### 2020-01-08 types20-abstract.tex
first commit

\documentclass{easychair}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{proof}
\usepackage{tikz}
\usetikzlibrary{cd}

\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Setoid}{\mathbf{Setoid}}
\newcommand{\Cat}{\mathbf{Cat}}
\newcommand{\Con}{\mathrm{Con}}
\newcommand{\Ty}{\mathrm{Ty}}
\newcommand{\Tm}{\mathrm{Tm}}
\newcommand{\Tms}{\mathrm{Tms}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\cat}[1]{\underline{\mathbf{#1}}}
\newcommand{\U}{\mathrm{U}}
\newcommand{\J}{\mathrm{J}}
\newcommand{\El}{\mathrm{El}}

% \newcommand{\Set}{\mathbf{Set}}
% \newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Type}{\mathbf{Type}}
\renewcommand{\iff}{\leftrightarrow}
\newcommand{\Nat}{\mathbb{N}}
\newcommand{\Int}{\mathbb{Z}}
\newcommand{\Rat}{\mathbb{Q}}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\Bool}{\mathrm{Bool}}

\newcommand{\List}{\mathrm{List}}
\newcommand{\rev}{\mathrm{rev}}

%\newtheorem{theorem}{Theorem}

\providecommand\mathbbm{\mathbb}
\def\lv{\mathopen{{[\kern-0.14em[}}}    % opening [[ value delimiter
\def\rv{\mathclose{{]\kern-0.14em]}}}   % closing ]] value delimiter
\newcommand{\eval}[1]{{\lv{#1}\rv}}

% \usepackage{amsmath}
% \usepackage{amssymb}
% %\usepackage{mathpartir}

% \newcommand{\Set}{\mathbf{Set}}
% \newcommand{\Prop}{\mathbf{Prop}}
% \newcommand{\Type}{\mathbf{Type}}
% \renewcommand{\iff}{\leftrightarrow}
% \newcommand{\Nat}{\mathbb{N}}
% \newtheorem{theorem}{Theorem}
% \newcommand{\cont}[2]{#1 \lhd #2}
% \def\lv{\mathopen{{[\kern-0.14em[}}}    % opening [[ value delimiter
% \def\rv{\mathclose{{]\kern-0.14em]}}}   % closing ]] value delimiter
% \newcommand{\eval}[1]{{\lv{#1}\rv}}

%\usepackage{natbib}
%\setcitestyle{round}
% Note that the \doi command from the doi package doesn't enable the
% same kinds of line breaks as the command below.
\newcommand{\doi}[1]{doi:\href{http://doi.org/#1}{%
    \urlstyle{same}\nolinkurl{#1}}}

% TODO notes.
\usepackage[textsize=small]{todonotes}
\setlength{\marginparwidth}{2cm}
\newcommand{\txatodo}[1]{\todo[fancyline,color=green!20]{#1}{}}

\title{Constructing a universe for the setoid model}

\author{Thorsten Altenkirch\inst{1}\thanks{Supported by USAF grant
    FA9550-16-1-0029.} \and AmbrusKaposi \inst{2} \and Filippo Sestini\inst{1}}

\institute{
  School of Computer Science, University of Nottingham, UK\\
  \email{\{psztxa,psxfs5\}@nottingham.ac.uk}
  \and
  E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary\\
  \email{akaposi@inf.elte.hu}
}

\begin{document}
\maketitle

The setoid model gives rise to a translation of a type theory with
functional and propositional extensionality to one without. It is thus
a way to explain extensionality in a type-theoretic and
computationally adequate way \cite{alti99,mpc19}. This translation
relies on the existence of a strict universe of propositions $\Prop$
as implemented in Coq and Agda \cite{Sprop}. This interpretation
in turn models a universe of propositions where equality of propositions
is logical equivalence, this is a very basic instance of
univalence. However, we also would like to be able to refer to a
universe of setoids, which cannot be univalent, because this would not
be a setoid.

To provide such a universe we need $\U : \Setoid$ and a family of
setoids $\El : \U \to \Setoid$ which provide codes for basic type
formers, like $\Pi$-types and booleans. This can be obtained as an
inductive recursive type.

\emph{Insert a summary of the construction}

However, we don't want to assume inductive-recursive types in the
basic type theory which is the target of the setoid model. We know
that we can translate basic instances of induction-recursion into
inductive families using the equivalence of families and discrete
fibrations. E.g. the inductive recursive definition of a universe $\U:\Set$
and $\El : \U \to \Setoid$ with $\Pi$-types and booleans is:

\emph{insert definition}

Assuming a universe of sets in the target theory we can model this as
an inductive type as follows:

\emph{insert definition}

Our result is that a modified form of this translation also works for
the more complex type we need to model the setoid universe. However,
we now need an inductive-inductive type in the target theory.

\emph{Insert a summary of the construction}

We know that finitary indutive-inductive definitions can be translated
into inductive families \cite{jakob,ambroise} but it is not clear wether this construction
extends to an infinitary type like the one above. This is subject of
further work. If successful we would be able to give a translation of
the setoid model with a universe into a very basic core type theory.

\bibliographystyle{alpha}
\bibliography{local}

\end{document}

### 2020-01-08 types20-abstract.tex
Write more

@@
 \usepackage{tikz}
+\usepackage{bbm}
+\usepackage[bw]{agda}
 \usetikzlibrary{cd}
 
-\newcommand{\Set}{\mathbf{Set}}
-\newcommand{\Prop}{\mathbf{Prop}}
-\newcommand{\Setoid}{\mathbf{Setoid}}
-\newcommand{\Cat}{\mathbf{Cat}}
+\newcommand{\setoidU}{\mathcal{U}}
+\newcommand{\ad}[1]{\AgdaFunction{#1}}
+\newcommand{\Set}{\textsf{Set}}
+\newcommand{\Prop}{\textsf{Prop}}
+\newcommand{\Setoid}{\textsf{Setoid}}
 \newcommand{\Con}{\mathrm{Con}}
@@
 \newcommand{\cat}[1]{\underline{\mathbf{#1}}}
-\newcommand{\U}{\mathrm{U}}
+\newcommand{\U}{\textsf{U}}
 \newcommand{\J}{\mathrm{J}}
-\newcommand{\El}{\mathrm{El}}
+\newcommand{\El}{\textsf{El}}
 
@@
 
-The setoid model gives rise to a translation of a type theory with
-functional and propositional extensionality to one without. It is thus
-a way to explain extensionality in a type-theoretic and
-computationally adequate way \cite{alti99,mpc19}. This translation
-relies on the existence of a strict universe of propositions $\Prop$
-as implemented in Coq and Agda \cite{Sprop}. This interpretation
-in turn models a universe of propositions where equality of propositions
-is logical equivalence, this is a very basic instance of
-univalence. However, we also would like to be able to refer to a
-universe of setoids, which cannot be univalent, because this would not
-be a setoid.
+The setoid interpretation gives rise to a model of a type theory with functional
+and propositional extensionality. It is thus a way to explain extensionality in
+a type-theoretic and computationally adequate way \cite{alti99}. The setoid
+model has been recently adapted into Setoid Type Theory (SeTT), which is
+justified by a syntactic translation into a very basic \emph{target theory}.
+\cite{mpc19}. This translation relies on the existence in the target theory of a
+universe of definitionally proof-irrelevant propositions $\Prop$, as recently
+implemented in Coq and Agda \cite{Sprop}. The setoid interpretation models a
+universe of propositions where equality of propositions is logical equivalence,
+thus providing a very basic instance of Univalence. However, we also would like
+to be able to equip SeTT with a universe of setoids. This universe must in turn
+be a setoid, hence in particular it cannot be univalent, but rather its
+equivalence relation must reflect simple equality of codes.
 
-To provide such a universe we need $\U : \Setoid$ and a family of
-setoids $\El : \U \to \Setoid$ which provide codes for basic type
-formers, like $\Pi$-types and booleans. This can be obtained as an
-inductive recursive type.
+To provide such a universe we need to define a setoid $\setoidU : \Setoid$ and a
+family of setoids $\El : \setoidU \to \Setoid$, with codes for basic type
+formers like $\Pi$-types and booleans. This can be obtained as an inductive
+recursive type.
+%
+\begin{align*}
+  & \textsf{data} \ \setoidU : \Set_1 \\
+  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \Prop_1 \\
+  & \El : \setoidU \to \Set \\
+  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
+\end{align*}
 
-\emph{Insert a summary of the construction}
+However, we don't want to assume inductive-recursive types in the basic type
+theory which is the target of the setoid model. We know that we can translate
+basic instances of induction-recursion into inductive families using the
+equivalence of families and discrete fibrations. For example, the
+inductive-recursive definition of a universe $\U:\Set_1$ and $\El : \U \to
+\Set$ with $\Pi$-types and booleans is
+\footnote{In Agda, this definition would also go through if $\U$ were in $\Set$, but
+  this seems to be a non-conservative extension.}:
+%
+\begin{align*}
+  & \textsf{data}\ \U\ : \Set_1 \\
+  & \quad \textsf{bool} : \U \\
+  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
+  & \El\ \textsf{bool} = \mathbbm{2} \\
+  & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
+\end{align*}
 
-However, we don't want to assume inductive-recursive types in the
-basic type theory which is the target of the setoid model. We know
-that we can translate basic instances of induction-recursion into
-inductive families using the equivalence of families and discrete
-fibrations. E.g. the inductive recursive definition of a universe $\U:\Set$
-and $\El : \U \to \Setoid$ with $\Pi$-types and booleans is:
+We can model this as an inductive type $\ad{in-U}$ that \emph{carves out} all
+types in $\Set$ that are in the image of $\El$:
+%
+\begin{align*}
+  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
+  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
+  & \quad \textsf{inPi}
+  :  \{A : \Set\} \{B : A \to \Set\}\\
+  & \ \qquad \to \textsf{in-U}\ A
+  \to ((a : A) \to \textsf{in-U}\ (B\ a))
+  \to \textsf{in-U}\ ((a : A) \to (B\ a))
+\end{align*}
+\begin{align*}
+  \U & = \Sigma (A : \Set) (\textsf{in-U}\ A) \\
+  \El & = \textsf{proj}_1
+\end{align*}
 
-\emph{insert definition}
+Note that this construction gives rise to an inductive-recursive universe in
+$\Set_1$, rather than $\Set$, thus the meta-theory must contain at least one
+universe. Our result is that a modified form of this translation also works for
+the more complex inductive-recursive type we need to model the universe of
+setoids. 
+%
+In particular, in addition to $\ad{in-U}$ for defining $\U$ as before, we also
+introduce a family $\ad{in-U∼}$ of binary relations between types in the
+universe, from which we then define $\_\ad{∼U}\_$.
+%
+However, we now need an inductive-inductive type in the target theory.
+%
+\begin{align*}
+  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
+  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1 \\
+  & \setoidU : \Set_1 \\
+  & \El : \setoidU \to \Set \\
+  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \Prop_1 \\
+  & \_\vdash\_\sim_{\El}\_ : \{a\ a' : \U\} \to a\ \sim_{\setoidU}\ a' \to \El\ a \to \El\ a' \to \Prop \\
+  & \\
+  & \U = \Sigma\ (X : \Set)\ (\AgdaDatatype{in-U}\ X) \\
+  & \El = \textsf{proj}_1 \\
+  & (X , p) \sim_{\setoidU} (X' , p') =
+   \ \parallel \Sigma\ (R : X \to X' \to \Prop) (\ad{in-U~} p\ p'\ R) \parallel
+\end{align*}
 
-Assuming a universe of sets in the target theory we can model this as
-an inductive type as follows:
+where $\parallel A \parallel\ : \Prop$ is propositional truncation for any $A :
+\Set$. Intuition would suggest to define $\_\vdash\_\sim_{\El}\_$ by projecting
+the relation out of the proof of $a \sim_{\setoidU} a'$, in much the same way as
+$\El$ is defined by projecting out of $\U$. That is, $p \vdash x \sim_{\El} y =
+\ad{proj}_1\ p\ x\ y$.
+%
+However, this doesn't work since the type of $p$ is propositionally truncated,
+hence it cannot be used to construct a proof-relevant object. We can work around
+this by defining $\_\vdash\_\sim_{\El}\_$ explicitly for all type formers, by
+induction on the codes $a\ a' : \U$, mutually with a proof that this definition
+is in some sense equivalent to what we would get if we could project out of the
+truncation.
 
-\emph{insert definition}
+We know that finitary inductive-inductive definitions can be translated into
+inductive families \cite{iit-erasure,iit-to-ix,induction-is-enough} but it is
+not clear whether this construction extends to an infinitary type like the one
+above. This is subject of further work. If successful we would be able to give a
+translation of the setoid model with a universe into a very basic core type
+theory.
 
-Our result is that a modified form of this translation also works for
-the more complex type we need to model the setoid universe. However,
-we now need an inductive-inductive type in the target theory.
-
-\emph{Insert a summary of the construction}
-
-We know that finitary indutive-inductive definitions can be translated
-into inductive families \cite{jakob,ambroise} but it is not clear wether this construction
-extends to an infinitary type like the one above. This is subject of
-further work. If successful we would be able to give a translation of
-the setoid model with a universe into a very basic core type theory.
-
-\bibliographystyle{alpha}
-\bibliography{local}
+\bibliography{bibliography}
 

### 2020-01-08 types20-abstract.tex
Add missing citations

@@
 and propositional extensionality. It is thus a way to explain extensionality in
-a type-theoretic and computationally adequate way \cite{alti99}. The setoid
-model has been recently adapted into Setoid Type Theory (SeTT), which is
-justified by a syntactic translation into a very basic \emph{target theory}.
+a type-theoretic and computationally adequate way \cite{setoid-model}. The
+setoid model has been recently adapted into Setoid Type Theory (SeTT), which is
+justified by a syntactic translation into a very basic \emph{target theory}
 \cite{mpc19}. This translation relies on the existence in the target theory of a
 universe of definitionally proof-irrelevant propositions $\Prop$, as recently
-implemented in Coq and Agda \cite{Sprop}. The setoid interpretation models a
+implemented in Coq and Agda \cite{gilbert}. The setoid interpretation models a
 universe of propositions where equality of propositions is logical equivalence,
@@
 introduce a family $\ad{in-U∼}$ of binary relations between types in the
-universe, from which we then define $\_\ad{∼U}\_$.
+universe, from which we then define $\_\sim_{\setoidU}\_$.
 %
@@
   & \\
-  & \U = \Sigma\ (X : \Set)\ (\AgdaDatatype{in-U}\ X) \\
+  & \setoidU = \Sigma\ (X : \Set)\ (\AgdaDatatype{in-U}\ X) \\
   & \El = \textsf{proj}_1 \\
   & (X , p) \sim_{\setoidU} (X' , p') =
-   \ \parallel \Sigma\ (R : X \to X' \to \Prop) (\ad{in-U~} p\ p'\ R) \parallel
+   \ \parallel \Sigma\ (R : X \to X' \to \Prop) (\ad{in-U∼}\ p\ p'\ R) \parallel
 \end{align*}
@@
 induction on the codes $a\ a' : \U$, mutually with a proof that this definition
-is in some sense equivalent to what we would get if we could project out of the
+is logically equivalent to what we would obtain if we could project out of the
 truncation.
@@
 
+\bibliographystyle{alpha}
 \bibliography{bibliography}

### 2020-01-08 types20-abstract.tex
Fix some macros

@@
   :  \{A : \Set\} \{B : A \to \Set\}\\
-  & \ \qquad \to \textsf{in-U}\ A
-  \to ((a : A) \to \textsf{in-U}\ (B\ a))
-  \to \textsf{in-U}\ ((a : A) \to (B\ a))
+  & \ \qquad \to \ad{in-U}\ A
+  \to ((a : A) \to \ad{in-U}\ (B\ a))
+  \to \ad{in-U}\ ((a : A) \to (B\ a))
 \end{align*}
@@
   & \\
-  & \setoidU = \Sigma\ (X : \Set)\ (\AgdaDatatype{in-U}\ X) \\
+  & \setoidU = \Sigma\ (X : \Set)\ (\ad{in-U}\ X) \\
   & \El = \textsf{proj}_1 \\
@@
 this by defining $\_\vdash\_\sim_{\El}\_$ explicitly for all type formers, by
-induction on the codes $a\ a' : \U$, mutually with a proof that this definition
-is logically equivalent to what we would obtain if we could project out of the
-truncation.
+induction on the codes $a\ a' : \setoidU$, mutually with a proof that this
+definition is logically equivalent to what we would obtain if we could project
+out of the truncation.
 

### 2020-01-08 types20-abstract.tex
small fixes

@@
 family of setoids $\El : \setoidU \to \Setoid$, with codes for basic type
-formers like $\Pi$-types and booleans. This can be obtained as an inductive
-recursive type.
+formers like $\Pi$-types and booleans. This can be obtained as an inductive-recursive type.
 %
@@
 However, we don't want to assume inductive-recursive types in the basic type
-theory which is the target of the setoid model. We know that we can translate
+theory which is the target of the setoid translation. We know that we can translate
 basic instances of induction-recursion into inductive families using the
@@
 inductive-recursive definition of a universe $\U:\Set_1$ and $\El : \U \to
-\Set$ with $\Pi$-types and booleans is
-\footnote{In Agda, this definition would also go through if $\U$ were in $\Set$, but
+\Set$ with $\Pi$-types and booleans is\footnote{In Agda, this definition would also go through if $\U$ were in $\Set$, but
   this seems to be a non-conservative extension.}:
@@
 \end{align*}
+Using this $\U$ and $\El$ can be given as follows:
 \begin{align*}
-  \U & = \Sigma (A : \Set) (\textsf{in-U}\ A) \\
+  \U & = \Sigma (A : \Set)\ (\textsf{in-U}\ A) \\
   \El & = \textsf{proj}_1
@@
 
-Note that this construction gives rise to an inductive-recursive universe in
+Note that this construction gives rise to a universe in
 $\Set_1$, rather than $\Set$, thus the meta-theory must contain at least one
@@
   & (X , p) \sim_{\setoidU} (X' , p') =
-   \ \parallel \Sigma\ (R : X \to X' \to \Prop) (\ad{in-U∼}\ p\ p'\ R) \parallel
+   \ \parallel \Sigma\ (R : X \to X' \to \Prop)\ (\ad{in-U∼}\ p\ p'\ R) \parallel
 \end{align*}

### 2020-01-09 types20-abstract.tex
added titlerunning and authorrunning
added blank before Kaposi

@@
 \title{Constructing a universe for the setoid model}
+\titlerunning{Constructing a universe for the setoid model}
+\authorrunning{Altenkirch, Kaposi, Sestini}
 
 \author{Thorsten Altenkirch\inst{1}\thanks{Supported by USAF grant
-    FA9550-16-1-0029.} \and AmbrusKaposi \inst{2} \and Filippo Sestini\inst{1}}
+    FA9550-16-1-0029.} \and Ambrus Kaposi \inst{2} \and Filippo Sestini\inst{1}}
 

### 2020-01-20 types20-abstract.tex
Updated author list.

@@
 \titlerunning{Constructing a universe for the setoid model}
-\authorrunning{Altenkirch, Kaposi, Sestini}
+\authorrunning{Altenkirch, Kaposi, Sattler, Sestini}
 
-\author{Thorsten Altenkirch\inst{1}\thanks{Supported by USAF grant
-    FA9550-16-1-0029.} \and Ambrus Kaposi \inst{2} \and Filippo Sestini\inst{1}}
+\author{
+  Thorsten Altenkirch \inst{1}\thanks{Supported by USAF grant FA9550-16-1-0029.}
+\and
+  Ambrus Kaposi \inst{2}
+\and
+  Christian Sattler\inst{1}\thanks{Supported by USAF grant FA9550-16-1-0029.}
+\and
+  Filippo Sestini\inst{1}
+}
 
@@
   School of Computer Science, University of Nottingham, UK\\
-  \email{\{psztxa,psxfs5\}@nottingham.ac.uk}
+  \email{\{psztxa,pszcs1,psxfs5\}@nottingham.ac.uk}
   \and

### 2020-01-20 types20-abstract.tex
Minor edits, more explicit equivalence [I, Set] \simeq Set/I.

@@
 universe of propositions where equality of propositions is logical equivalence,
-thus providing a very basic instance of Univalence. However, we also would like
+thus providing a very basic instance of univalence. However, we also would like
 to be able to equip SeTT with a universe of setoids. This universe must in turn
@@
 basic instances of induction-recursion into inductive families using the
-equivalence of families and discrete fibrations. For example, the
+equivalence of $I$-indexed families of sets and sets over $I$,
+i.e.\ $[I, \Set_i] \simeq \Set_i/I$ for $I : \Set_i$. For example, the
 inductive-recursive definition of a universe $\U:\Set_1$ and $\El : \U \to
-\Set$ with $\Pi$-types and booleans is\footnote{In Agda, this definition would also go through if $\U$ were in $\Set$, but
+\Set$ with $\Pi$-types and Booleans is\footnote{In Agda, this definition would also go through if $\U$ were in $\Set$, but
   this seems to be a non-conservative extension.}:
@@
 %
-However, we now need an inductive-inductive type in the target theory.
+However, we now need an inductive-inductive type in the target theory:
 %
@@
 
-where $\parallel A \parallel\ : \Prop$ is propositional truncation for any $A :
+where $\parallel A \parallel\ : \Prop$ denotes the propositional truncation of $A :
 \Set$. Intuition would suggest to define $\_\vdash\_\sim_{\El}\_$ by projecting
@@
 hence it cannot be used to construct a proof-relevant object. We can work around
-this by defining $\_\vdash\_\sim_{\El}\_$ explicitly for all type formers, by
-induction on the codes $a\ a' : \setoidU$, mutually with a proof that this
-definition is logically equivalent to what we would obtain if we could project
-out of the truncation.
+this by instead defining $\_\vdash\_\sim_{\El}\_$ by induction on the
+codes $a\ a' : \setoidU$, explicitly for each type former, mutually with a proof
+that this definition is logically equivalent to what we would obtain if we could
+project out of the truncation.
 

### 2020-05-05 paper.tex
Add paper.tex

--- 2020-01-20 types20-abstract.tex
@@
 
+\newcommand{\refl}{\textsf{refl}}
+\newcommand{\GG}{\Gamma}
+\newcommand{\DD}{\Delta}
+\newcommand{\GD}{\Delta}
+\newcommand{\R}{\textsf{R}}
+\newcommand{\Id}{\textsf{Id}}
+\newcommand{\coe}{\textsf{coe}}
+\newcommand{\coh}{\textsf{coh}}
+\newcommand{\J}{\textsf{J}}
+\newcommand{\tyj}[2]{#2 : \Ty\ #1}
+\newcommand{\tmj}[3]{#2 : \Tm\ #1\ #3}
+\newcommand{\subj}[3]{#1 : \mathrm{Sub}\ #2\ #3}
 \newcommand{\setoidU}{\mathcal{U}}
@@
 \newcommand{\U}{\textsf{U}}
-\newcommand{\J}{\mathrm{J}}
 \newcommand{\El}{\textsf{El}}
 
-% \newcommand{\Set}{\mathbf{Set}}
-% \newcommand{\Prop}{\mathbf{Prop}}
 \newcommand{\Type}{\mathbf{Type}}
@@
 % %\usepackage{mathpartir}
-
-% \newcommand{\Set}{\mathbf{Set}}
-% \newcommand{\Prop}{\mathbf{Prop}}
-% \newcommand{\Type}{\mathbf{Type}}
-% \renewcommand{\iff}{\leftrightarrow}
-% \newcommand{\Nat}{\mathbb{N}}
-% \newtheorem{theorem}{Theorem}
-% \newcommand{\cont}[2]{#1 \lhd #2}
-% \def\lv{\mathopen{{[\kern-0.14em[}}}    % opening [[ value delimiter
-% \def\rv{\mathclose{{]\kern-0.14em]}}}   % closing ]] value delimiter
-% \newcommand{\eval}[1]{{\lv{#1}\rv}}
 
@@
 
+\section{Introduction}
+
+...
+
+Intensional Type Theory has nice computational properties, like decidability of
+type-checking, that can make it more suitable for computer formalization, since
+proof-checking can be implemented as an entirely automatic procedure.
+%
+However, a serious drawback of ITT is given by the lack of extensionality.
+%
+Two extensionality principles are particularly missed from intensional type
+theory, in that they arguably constitute the bare minimum for type theory to be
+comparable to set theory as a foundational system for set-level mathematics.
+%
+The first of these is functional extensionality, i.e. the principle according to
+which two functions can be considered equal if they map equal inputs to equal
+outputs.
+%
+The second principle is propositional extensionality, according to which two
+propositions should be considered equal when logically equivalent.
+
+These extensionality principles are not theorems of intensional type theory as
+usually defined [cite], although they are consistent with it [cite].
+
+As a consequence, these principles are often added as axioms when formalizing
+mathematics using proof-assistants based on ITT. This solution, however, is far
+from optimal, as the introduction of axioms destroys the nice computational
+properties of the system, particularly canonicity. Unfortunately, one can show
+that whenever equality is defined according to the usual inductive definition by
+ML, there is no hope to have a theory where both canonicity and funext/propext
+coexist.
+
+The first step is therefore to give up the idea of defining equality as a single
+inductive definition, and adopt the approach according to which equality should
+be defined in a specific way for each type former in the type theory. That is,
+every type former should come equipped with its own notion of equality, from
+which a general notion of equality is derived.
+
+Two systems of ITT with extensional properties that follow this approach are OTT
+and cubical TT. In the first one ... TODO. In cubical TT ... TODO. Note that we
+do not consider HoTT since, although it is a version of ITT with extensional
+properties, it's not known to enjoy canonicity.
+
+The system that we will consider in this paper follows the same approach of OTT,
+namely the setoid model... TODO
+
+* Internal encoding of setoids
+
+* Setoid hell
+
+* Setoid TT: MLTT + setoid constructs
+
+* universe of setoids
+
+\section{Setoid Type Theory}
+
+\subsection{Setoid model}
+
 The setoid interpretation gives rise to a model of a type theory with functional
 and propositional extensionality. It is thus a way to explain extensionality in
-a type-theoretic and computationally adequate way \cite{setoid-model}. The
-setoid model has been recently adapted into Setoid Type Theory (SeTT), which is
-justified by a syntactic translation into a very basic \emph{target theory}
+a type-theoretic and computationally adequate way \cite{setoid-model}.
+
+\todo{say something more about this}
+
+\subsection{SeTT}
+
+The setoid model has been recently adapted into Setoid Type Theory (SeTT), which
+is justified by a syntactic translation into a very basic \emph{target theory}
 \cite{mpc19}. This translation relies on the existence in the target theory of a
@@
 universe of propositions where equality of propositions is logical equivalence,
-thus providing a very basic instance of univalence. However, we also would like
-to be able to equip SeTT with a universe of setoids. This universe must in turn
-be a setoid, hence in particular it cannot be univalent, but rather its
-equivalence relation must reflect simple equality of codes.
+thus providing a very basic instance of univalence.
 
-To provide such a universe we need to define a setoid $\setoidU : \Setoid$ and a
-family of setoids $\El : \setoidU \to \Setoid$, with codes for basic type
-formers like $\Pi$-types and booleans. This can be obtained as an inductive-recursive type.
+Technically speaking, SeTT is an extension of MLTT with extra component to
+support setoid reasoning. These extra components are nothing more than an
+internalization of the idea that every type is a setoid. More precisely, we have
+rules [explaining TODO] that contexts (aka closed types) are setoids, and that
+dependent types are setoids indexed over setoids.
+
+\[
+\infer{
+  \GG^\sim\ \rho_0\ \rho_1 : \Tm\ \DD\ \Prop
+}{
+  \GG : \Con \qquad \subj{\rho_0, \rho_1}{\DD}{\GG}
+}
+\qquad
+\infer{
+  A^{\sim}\ p\ a_0\ a_1 : \Tm\ \DD\ \Prop
+}{
+  \deduce{
+    a_0 : \Tm\ \DD\ A[\rho_0] \qquad a_1 : \Tm\ \DD\ A[\rho_1]
+  }{
+    A : \Ty\ \GG \qquad p : \Tm\ \DD\ (\El\ (\GG^{\sim}\ \rho_0\ \rho_1))
+  }
+}
+\]
+
+Note that the setoid equivalence relations are strict propositions. In addition,
+we also have rules representing the fact that every construction in SeTT
+respects setoid equality, so that we can transport along any such equality:
+
+[coe and coh rules here]
+
+    \[
+    \infer{
+      \deduce{
+        \tmj{\GD}{\coh_A\ p\ a}{\underline{A^\sim\ p\ a\ (\coe_A\ p\ a)}}
+      }{
+        \tmj{\GD}{\coe_A\ p\ a}{A[\rho_1]}
+      }
+    }{
+      \tyj{\GG}{A} \qquad \subj{\rho_0, \rho_1}{\GD}{\GG} \qquad
+      \tmj{\GD}{p}{\underline{\GG^\sim\ \rho_0\ \rho_1}}\qquad
+      \tmj{\GD}{a}{A[\rho_0]}
+    }
+    \]
+
+Equality in SeTT differs from the standard approach by Martin-Loef in that every
+type comes equipped with an equality relation. This is in contrast with ML's
+identity type which is defined uniformly for any type.
+
+Nevertheless, we can easily recover the usual ML identity type from setoid
+equality:
+
+\[
+\infer{
+  \textsf{Id}_A\ a_0\ a_1 :\equiv A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
+}{
+  \tyj{\GG}{A} \qquad \tmj{\GG}{a_0, a_1}{A}
+}
+\]
+
+Note that ... we have UIP... Moreover, contractibility of singletons holds by
+UIP and by definition of equality on $\Sigma$ types. Therefore we can define the
+$\J$ eliminator:
+
+\[
+\infer{
+  \tmj{\GG}{\J_M\ p\ d}{M[a_0, a_1, p]}
+}{
+  \deduce{
+    \tmj{\GG}{d}{M[a_0, a_0, \refl_A\ a_0]}
+    \qquad
+    \tmj{\GG}{p}{\underline{\Id_A\ a_0\ a_1}}
+  }{
+  \tyj{(\GG, x : A, y : A, p : \Id_A\ x\ y)}{M}
+  }
+}
+\]
+
+\section{Universe of setoids}
+
+As currently defined, SeTT is a system already powerful enough to support the
+formalization of part of set-level mathematics. However, we also would like to
+be able to equip SeTT with a universe of setoids. This universe must in turn be
+a setoid, hence in particular it cannot be univalent, but rather its equivalence
+relation must reflect simple equality of codes.
+
+As we use Type Theory as our metatheory, this universe of setoids is going to be
+an internal construction in Type Theory. The usual way to encode universes in
+Type Theory is by an inductive-recursive definition. As we will see in Section
+TODO, it is indeed possible to give a IR definition of a universe of setoids.
 %
-\begin{align*}
-  & \textsf{data} \ \setoidU : \Set_1 \\
-  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \Prop_1 \\
-  & \El : \setoidU \to \Set \\
-  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
-\end{align*}
+However, we would prefer not to be forced to assume inductive-recursive types,
+or IR in general, in the basic type theory which is the target of the setoid
+translation. This is because the point of the translation is to explain a
+complex system in terms of a very simple, well understood core, and we think
+introducing IR would go against this goal.
 
-However, we don't want to assume inductive-recursive types in the basic type
-theory which is the target of the setoid translation. We know that we can translate
-basic instances of induction-recursion into inductive families using the
-equivalence of $I$-indexed families of sets and sets over $I$,
-i.e.\ $[I, \Set_i] \simeq \Set_i/I$ for $I : \Set_i$. For example, the
-inductive-recursive definition of a universe $\U:\Set_1$ and $\El : \U \to
-\Set$ with $\Pi$-types and Booleans is\footnote{In Agda, this definition would also go through if $\U$ were in $\Set$, but
-  this seems to be a non-conservative extension.}:
-%
+Fortunately, it is possible to reduce some instances of induction-recursion to
+plain induction (more specifically, inductive families), and the well known IR
+universe construction is indeed one of them.
+
+In the next section, we recall the IR construction of universes in type theory,
+as well as lay out the process by which such construction can be turned to one
+that only relies on inductive families. In Section TODO, we give a first IR
+definition of the universe of setoids. After that we will see if and how such IR
+definition can be turned into one that does not rely on IR.
+
+\subsection{Encoding of universes via IR and II}
+
+A standard, well-known way of encoding universes in type theory is via induction-recursion.
+\todo{a couple of words about induction recursion}.
+
+To define a universe, we simultaneously define the universe of codes itself, as
+an inductive definition, and a family $\El : \U \to \Set$ that assigns, to each
+code corresponding to some type, the meta-theoretic type of its elements.
+
+For example, the definition of a small universe with booleans and $\Pi$ types
+would be the following:
+
 \begin{align*}
@@
 
-We can model this as an inductive type $\ad{in-U}$ that \emph{carves out} all
-types in $\Set$ that are in the image of $\El$:
+Note that $\U$ is defined to be in $\Set_1$ rather than $\Set$. This is
+because... \todo{TODO}
+
+As it turns out, induction-recursion is not really required to encode such
+a universe, although is certainly makes the definition particularly nice.
+%
+In fact, we know that we can translate basic instances of induction-recursion
+into inductive families using the equivalence of $I$-indexed families of sets
+and sets over $I$, i.e.\ $[I, \Set_i] \simeq \Set_i/I$ for $I : \Set_i$.
+
+Following this, we seen that we can actually encode the universe above as an
+inductive type $\ad{in-U}$ that \emph{carves out} all types in $\Set$ that are
+in the image of $\El$:
 %
@@
 
-Note that this construction gives rise to a universe in
-$\Set_1$, rather than $\Set$, thus the meta-theory must contain at least one
-universe. Our result is that a modified form of this translation also works for
+Note that this construction gives rise to a universe in $\Set_1$, rather than
+$\Set$, since the definition of $\U$ quantifies over all possible sets. Thus,
+the meta-theory must contain at least one universe for this construction to go
+through.
+
+%% ----------------------
+
+%% However, we don't want to assume inductive-recursive types in the basic type
+%% theory which is the target of the setoid translation. We know that we can translate
+%% basic instances of induction-recursion into inductive families using the
+%% equivalence of $I$-indexed families of sets and sets over $I$,
+%% i.e.\ $[I, \Set_i] \simeq \Set_i/I$ for $I : \Set_i$. For example, the
+%% inductive-recursive definition of a universe $\U:\Set_1$ and $\El : \U \to
+%% \Set$ with $\Pi$-types and Booleans is\footnote{In Agda, this definition would also go through if $\U$ were in $\Set$, but
+%%   this seems to be a non-conservative extension.}:
+%% %
+%% \begin{align*}
+%%   & \textsf{data}\ \U\ : \Set_1 \\
+%%   & \quad \textsf{bool} : \U \\
+%%   & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
+%%   & \El\ \textsf{bool} = \mathbbm{2} \\
+%%   & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
+%% \end{align*}
+
+%% We can model this as an inductive type $\ad{in-U}$ that \emph{carves out} all
+%% types in $\Set$ that are in the image of $\El$:
+%% %
+%% \begin{align*}
+%%   & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
+%%   & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
+%%   & \quad \textsf{inPi}
+%%   :  \{A : \Set\} \{B : A \to \Set\}\\
+%%   & \ \qquad \to \ad{in-U}\ A
+%%   \to ((a : A) \to \ad{in-U}\ (B\ a))
+%%   \to \ad{in-U}\ ((a : A) \to (B\ a))
+%% \end{align*}
+%% Using this $\U$ and $\El$ can be given as follows:
+%% \begin{align*}
+%%   \U & = \Sigma (A : \Set)\ (\textsf{in-U}\ A) \\
+%%   \El & = \textsf{proj}_1
+%% \end{align*}
+
+%% Note that this construction gives rise to a universe in
+%% $\Set_1$, rather than $\Set$, thus the meta-theory must contain at least one
+%% universe.
+
+\subsection{Inductive-inductive universe}
+
+To provide such a universe we need to define a setoid $\setoidU : \Setoid$ and a
+family of setoids $\El : \setoidU \to \Setoid$, with codes for basic type
+formers like $\Pi$-types and booleans. This can be obtained as an inductive-recursive type.
+\todo{expand definition to include more details}
+%
+\begin{align*}
+  & \textsf{data} \ \setoidU : \Set_1 \\
+  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \Prop_1 \\
+  & \El : \setoidU \to \Set \\
+  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
+\end{align*}
+
+As said before, however, we would like to avoid having to assume
+inductive-recursive types, of more generally induction-recursion, in the
+metatheory.
+
+Our result is that a modified form of this translation also works for
 the more complex inductive-recursive type we need to model the universe of

### 2020-05-11 paper.tex
Add content to paper.tex

@@
 inductive definition, and adopt the approach according to which equality should
-be defined in a specific way for each type former in the type theory. That is,
-every type former should come equipped with its own notion of equality, from
-which a general notion of equality is derived.
+be defined specifically for each type former in the type theory. That is, every
+type former $A$ should be introduced together with its own notion of equality
+$A^\sim$, and the internal notion of identity in type theory should refer to
+that relation.
+%
+If one does this, then funext can be obtained by simply defining equality on
+function types to be point-wise equality. That is, $(A \to B)^\sim\ f\ g :\equiv
+\forall (x : A) , B^\sim\ (f x)\ (g x)$.
 
-Two systems of ITT with extensional properties that follow this approach are OTT
-and cubical TT. In the first one ... TODO. In cubical TT ... TODO. Note that we
-do not consider HoTT since, although it is a version of ITT with extensional
-properties, it's not known to enjoy canonicity.
+* One possible answer: setoids
 
-The system that we will consider in this paper follows the same approach of OTT,
-namely the setoid model... TODO
-
-* Internal encoding of setoids
-
-* Setoid hell
+This form of extensional reasoning is already available in type theory via
+\emph{setoids}. A setoid is just a type $A$ (in this setting, we assume that all
+types are sets in the HoTT sense) together with a propositional equivalence
+relation $A^\sim : A \to A \to \Prop$.
+%
+One can thus reason extensionally in ITT by defining an internal notion of
+setoid, and working with those instead of normal sets/types. Note that this
+approach only works provided one ensures that every construction performed on
+setoids respects the setoid relations involved. Unfortunately doing so is not an
+easy task, and most constructions beyond the more trivial ones will generate a
+great amount of proof obligations, especially when dependent setoids come into
+play, leading to a situation commonly known as \emph{setoid hell}.
 
@@
 
+The good news is that all this work proving that our type theoretic
+constructions respect the setoid relations is not necessary, because these
+already respect setoid equality \emph{by construction}. More precisely, there
+exists a sound interpretation of MLTT (with pi, sigma, etc...) into setoids,
+that maps closed types/contexts to setoids and dependent types to indexed
+setoids.
+%
+Thus there is no need to explicitly work with setoids and setoid relations in
+TT, as one can simply carry out their constructions in plain MLTT and use the
+aforementioned interpretation to map that construction into the world of setoid.
+
+This is interesting in principle, but practically useless. Since our
+constructions still take place in MLTT, we can't take advantage of the features
+of the setoid model our syntactic constructions are mapped to, since they are
+not accessible internall, from within the system itself.
+%
+This is where a key property of models comes into play. Models inform consistent
+extensions of the type theory, based on what holds true in the model, by
+reflecting the semanics into the syntax via additional syntactic constructs that
+are then mapped back to the original notions via the semantics.
+
+In our case, the existence of a setoid model suggests us ways to extend the
+syntax of MLTT with syntactic constructs that correspond to the setoid
+relations, allowing us to access and use them when proving theorems within the
+theory.
+%
+This is the idea behind Setoid Type Theory, SeTT for short.
+
+* Two incarnations of ITT that follow this approach are OTT and cubical TT. In OTT
+... TODO. In cubical TT ... TODO. Note that we do not include HoTT in the
+discussion since, although it is a version of ITT with extensional properties,
+it's not yet known to enjoy canonicity.
+
 * universe of setoids
+
+%% The system that we will consider in this paper follows the same motivation of
+%% OTT, of giving a syntactic presentation of the setoid model in [cite].
 
@@
 \todo{say something more about this}
+
+\todo{Say a bit about how one can turn the setoid model to a setoid translation
+  (strict model and such...)}
+
+\todo{This gives a translation/explanation/implementation of extensionality in
+  terms of intensional TT}
 
@@
 
+\todo{Mention that the universe cannot be univalent}
+
+\todo{Maybe mention the fact that it should be possible to give an elimination
+  principle for the setoid universe. This also makes sense since it's not univalent.}
+
 As currently defined, SeTT is a system already powerful enough to support the
@@
 inductive type $\ad{in-U}$ that \emph{carves out} all types in $\Set$ that are
-in the image of $\El$:
+in the image of $\El$. In other words, $\ad{in-U}$ is a predicate that holds for
+any type that would have been obtained via $\El$ in the usual
+inductive-recursive definition. As $\El$ is indexed on the type of codes, the
+definition of $\ad{in-U}$ quite expectedly reflecs the inductive structure of
+codes.
 %
@@
 family of setoids $\El : \setoidU \to \Setoid$, with codes for basic type
-formers like $\Pi$-types and booleans. This can be obtained as an inductive-recursive type.
-\todo{expand definition to include more details}
+formers like $\Pi$-types and booleans. This can be defined as the following
+inductive-recursive type. \todo{expand definition to include more details}
 %
@@
 
-As said before, however, we would like to avoid having to assume
+The point of having $\setoidU$ and $\_\sim_\setoidU\_$ is quite clear: our
+setoid universe is a type, hence itself a setoid, hence described as any other
+setoid by a carrier set and an equivalence relation on it. These functions are
+achieved by $\setoidU$ and $\_\sim_{\setoidU}\_$, respectively.
+
+For any code $A$ in the universe, we want a type $\El\ A$; this, in turn, should
+be a setoid as well. In other words, we want a family of setoids indexed over
+setoids. We thus need to explain, for any $A$, what is the carrier and
+equivalence relation of $\El\ A$. This is achieved by $\El : \setoidU \to \Set$,
+and $\_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to
+\El\ a \to \El\ a' \to \Prop$. Note that we need this to be indexed over code
+equality, because El is a type family indexed over U, and hence it must respect
+the equality on U. \todo{rewrite in better form}
+
+As already pointed out, however, we would like to avoid having to assume
 inductive-recursive types, of more generally induction-recursion, in the
@@
 %
-However, we now need an inductive-inductive type in the target theory:
-%
 \begin{align*}
   & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
-  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1 \\
+  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1
+\end{align*}
+
+Note that $\ad{in-U}$ appears in the sort of $\ad{in-U∼}$, thus making this
+definition as instance of induction-induction [cite]. \todo{add brief
+  explanation of induction-induction}.
+
+As in the previous example, we can now define $\setoidU$ as a $\Sigma$ type, and
+$\El$ as the corresponding first projection.
+
+\begin{align*}
   & \setoidU : \Set_1 \\
+  & \setoidU = \Sigma\ (X : \Set)\ (\ad{in-U}\ X) \\
   & \El : \setoidU \to \Set \\
+  & \El = \textsf{proj}_1
+\end{align*}
+
+What is left now is to define the setoid equality relation on the universe, as
+well as the setoid equality relation on $\El\ A$ for any $A$ in $\setoidU$.
+%
+The definition of $\_\sim_{\setoidU}\_$ follows, in principle, that of
+$\setoidU$. A ...
+
+Similarly, two codes $A, B$ in the universe $\setoidU$ are equal when there
+exists a setoid equivalence relation on their respective sets $\El\ A$ and $\El
+B$. Intuitively, since elements of a setoid are only ever compared to elements
+of the same setoid, this is only possible if $A, B$ are codes for the same
+setoid, that is, if $A \sim_{\setoidU} B$. Existence and well-formedness of such
+relations is expressed via the type $\ad{in-U~}$ just defined, hence formally we
+should have something like:
+
+\[
+A \sim_{\setoidU} B :\equiv \Sigma (R: ...) ...
+\]
+
+or, using pattern matching:
+
+\[
+(A, p) \sim_{\setoidU} (A', p') :\equiv \Sigma (R: ...) ...
+\]
+
+Unfortunately this definition only captures the idea we are aiming for, but does
+not actually typecheck. In fact, $\_\sim_{\setoidU}\_$ is a $\Prop_1$-valued
+relation, so $A \sim_{\setoidU} B$ should be a proposition, but $\Sigma (R: ...)
+...$ clearly isn't.
+
+The solution is actually quite simple, and it just involves truncating the
+$\Sigma$ type described above to force it to be in $\Prop_1$. More precisely, we
+assume to have in our metatheory, in addition to a hierarchy of universes of
+strict propositions $\Prop_i$, a truncation operation $\parallel\_\parallel :
+\Set_i \to \Prop_i$ with constructor $|\_| : \{A : \Set_i\} \to A \to \parallel
+A \parallel$ and eliminator $\ad{withTrunc} : \{B : \Prop_i\} \to (A \to B) \to
+\parallel A \parallel \to B$. Note that truncation can be defined in Agda
+extended with the universe of strict propositions, with the eliminator replaced
+by pattern matching.
+
+\begin{align*}
   & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \Prop_1 \\
-  & \_\vdash\_\sim_{\El}\_ : \{a\ a' : \U\} \to a\ \sim_{\setoidU}\ a' \to \El\ a \to \El\ a' \to \Prop \\
-  & \\
-  & \setoidU = \Sigma\ (X : \Set)\ (\ad{in-U}\ X) \\
-  & \El = \textsf{proj}_1 \\
   & (X , p) \sim_{\setoidU} (X' , p') =
@@
 
-where $\parallel A \parallel\ : \Prop$ denotes the propositional truncation of $A :
-\Set$. Intuition would suggest to define $\_\vdash\_\sim_{\El}\_$ by projecting
-the relation out of the proof of $a \sim_{\setoidU} a'$, in much the same way as
+This trick might seem to simple to the point of being naive, but as we will see
+it is sufficient to make the entire definition go through.
+
+We are now left to define $\_\vdash\_\sim_{\El}\_$.
+
+\begin{align*}
+  & \_\vdash\_\sim_{\El}\_ : \{a\ a' : \U\} \to a\ \sim_{\setoidU}\ a' \to \El\ a \to \El\ a' \to \Prop
+\end{align*}
+
+Intuition would suggest to define $\_\vdash\_\sim_{\El}\_$ by projecting
+the relation out of the proof of $A \sim_{\setoidU} A'$, in much the same way as
 $\El$ is defined by projecting out of $\U$. That is, $p \vdash x \sim_{\El} y =
@@
 However, this doesn't work since the type of $p$ is propositionally truncated,
-hence it cannot be used to construct a proof-relevant object. We can work around
-this by instead defining $\_\vdash\_\sim_{\El}\_$ by induction on the
-codes $a\ a' : \setoidU$, explicitly for each type former, mutually with a proof
-that this definition is logically equivalent to what we would obtain if we could
-project out of the truncation.
+hence it cannot be used to construct a proof-relevant object. In other words,
+the proof $p$ cannot be eliminated into $El A \to El A' \to Prop$, because
+$El A \to El A' \to Prop$ is not a proposition, but a $\Set$.
+
+We can work around this limitation by defining $p \vdash x \sim_{\El} y$ by
+induction on the codes $a\ a' : \setoidU$, in a way that is logically equivalent
+to the proposition we would have obtained by $\ad{proj}_1\ p\ x\ y$ if that
+expression was well-typed.
+
+Note that for this definition to go through, we sometimes need the fact that the
+relation we are building is indeed equivalent to the one we would have obtained
+if we could project out of the truncation. That is, we need an additional proof
+that for any concrete $R$ and $\ad{inR}$, $| (R, \ad{inR}) | \vdash x \sim_{\El}
+y \iff R\ x\ y$. We thus need to define $\_\vdash\_\sim_{\El}\_$ mutually with a proof
+of the following
+
+\[
+...
+\]
 

### 2020-05-13 paper.tex
Write more

@@
 usually defined [cite], although they are consistent with it [cite].
-
+%
 As a consequence, these principles are often added as axioms when formalizing
@@
 that whenever equality is defined according to the usual inductive definition by
-ML, there is no hope to have a theory where both canonicity and funext/propext
-coexist.
+Martin-Loef, there is no hope to have a theory where both canonicity and
+funext/propext coexist.
 
-The first step is therefore to give up the idea of defining equality as a single
-inductive definition, and adopt the approach according to which equality should
-be defined specifically for each type former in the type theory. That is, every
-type former $A$ should be introduced together with its own notion of equality
-$A^\sim$, and the internal notion of identity in type theory should refer to
-that relation.
+The first step out of this is to give up the idea of equality as a single
+inductive definition. Instead, equality should be defined specifically for each
+type former in the type theory. That is, every type former $A$ should be
+introduced together with its own notion of equality $A^\sim$.
 %
@@
 
-* One possible answer: setoids
+%% * One possible answer: setoids
 
@@
 
-* Setoid TT: MLTT + setoid constructs
+%% * Setoid TT: MLTT + setoid constructs
 
@@
 
+... TODO ...
+
 * universe of setoids
 
-%% The system that we will consider in this paper follows the same motivation of
-%% OTT, of giving a syntactic presentation of the setoid model in [cite].
+The current formulation of SeTT is already a satisfactory (\todo{maybe not the
+  right word: satisfactory to who? according to what metric?}) presentation of
+the setoid model, in that it includes all the features of the semantics that we
+wished to reflect in the syntax. It is also arguably expressive enough to
+formalize a good chunk of set-level mathematics. However, it is also currently
+lacking in a serious way, as the lack of universes seriously limits its
+introspection capabilities, and consequently its strength as a foundational
+system. Note that SeTT already includes a (univalent) universe of strict
+propositions; hence what we are referring to is a universe of setoids.
 
-\section{Setoid Type Theory}
+The inclusion of a universe of setoids in SeTT poses several interesting
+questions. First of all, since a universe of setoids is itself a type in SeTT,
+is must also be a setoid. However, it certainly cannot be a univalent one. One
+question is therefore what kind of equality to equip such universe.
+%
+The second question is how one can justify such universe in the setoid model.
+In particular what principles, if any, should be added to the metatheory in
+order to be able to define the universe.
 
-\subsection{Setoid model}
+This paper is an attempt to give an answer to the questions above.
+...TODO...
 
-The setoid interpretation gives rise to a model of a type theory with functional
-and propositional extensionality. It is thus a way to explain extensionality in
-a type-theoretic and computationally adequate way \cite{setoid-model}.
+\section{Setoid model}
 
-\todo{say something more about this}
+%% The setoid interpretation gives rise to a model of a type theory with functional
+%% and propositional extensionality. It is thus a way to explain extensionality in
+%% a type-theoretic and computationally adequate way \cite{setoid-model}.
 
-\todo{Say a bit about how one can turn the setoid model to a setoid translation
-  (strict model and such...)}
+By \emph{setoid model} we intend a class of models of (intensional) type theory
+where contexts/closed types are interpreted as setoids, i.e. sets with an
+equivalence relation, and dependent types are interpreted as dependent/indexed
+setoids.
+%
+A setoid model was first given by M. Hofmann in \cite{hofmann-ext-int} in order
+to give a semantics to extensionality principles such as functional and
+propositional extensionality. Here we consider a similar model construction, due
+to T. Altenkirch \cite{setoid-model}.
 
-\todo{This gives a translation/explanation/implementation of extensionality in
-  terms of intensional TT}
+The peculiarity of this model is that it is presented in a type theoretic and
+intensional metatheory. The metatheory is also very simple, requiring only $\Pi$
+types, $\Sigma$ types, and a universe of definitionally proof-irrelevant
+propositions itself closed under such type formers.
+%
+Notably, the use of strict prop makes this model a strict one, namely one where
+every equation in the model is a definitional equality in the metatheory.  The
+model itself validates the extensionality principles mentioned above, like
+funext, proext, and quotient types. As a consequence, this model construction
+provides a way to explain extensionality, by giving a way to implement it from
+stratch in terms of a base theory that is both very simple and intensional.
 
-\subsection{SeTT}
+This idea of implementing extensionality via this model can be made both more
+precise and more practical. Indeed, the strictness of this model implies that we
+can turn this model into a syntactic translation, from which a syntactic
+presentation of the model can be derived. (See next section, and [cite mpc'19])
+\todo{rewrite this.}
+
+\subsection{Setoid Type Theory}
 

### 2020-05-19 paper.tex
WIP

@@
 
-\subsection{Inductive-inductive universe}
+\subsection{A first definition of a setoid universe}
 
@@
 family of setoids $\El : \setoidU \to \Setoid$, with codes for basic type
-formers like $\Pi$-types and booleans. This can be defined as the following
-inductive-recursive type. \todo{expand definition to include more details}
+formers like $\Pi$-types and booleans. That is, we need the following data
 %
@@
 
-As already pointed out, however, we would like to avoid having to assume
-inductive-recursive types, of more generally induction-recursion, in the
-metatheory.
+\todo{Acknowledge Simon Boulier for the IIR universe construction}
+\todo{QIIT II LICS paper. this part is already in the HIIT paper}
 
-Our result is that a modified form of this translation also works for
-the more complex inductive-recursive type we need to model the universe of
-setoids. 
+We given an inductive definition of $\setoidU, ...$, mutually with a recursive
+definition of $\El, \_\vdash\_\sim_\El\_, \sim_\setoidU, \textsf{refl}_\setoidU, ...$.
+
+\begin{align*}
+  & \textsf{data}\ \setoidU : \Set_1 \\
+  & \quad \textsf{bool} : \setoidU \\
+  %% Πs : (a : U i)(b : El a → U i)
+     %% (refb : {x x' : El a}(x~ : refU a ⊢ x ~El x') → b x ~U b x') → U i
+  & \quad \textsf{pi}
+  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
+  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x' \to B\ x \sim_\setoidU B\ x')
+  \to \setoidU
+\end{align*}
+
+\[
+\textsf{refl}_\setoidU : ?
+\]
+
+Note that in the definition of $\setoidU$, we crucially require that the family
+$B : \El\ A \to \setoidU$ be a setoid morphism, hence that it respects the
+corresponding setoid equalities.
+
+This definition of the setoid universe relies on both induction-recursion and
+induction-induction. Although appily accepted by Agda, the meaning of this
+combination is not really understood. The goal of the next sections is therefore
+to find an alternative definition of the setoid universe that relies on
+principles that are better understood. Not only that, we would like to reduce
+the number of extensions to the metatheory that are needed to be able to give
+such definition, so to stay as close as possible to our initial idea of reducing
+SeTT to a small core.  \todo{reword}
+
+The next section explains our main result, which is that a modified form of the
+translation of IR to IF also works for the setoid universe defined above. This
+type is more complex that the basic IR types we have seen before, in that it
+uses a combination of IR and II, and includes the universe of strict
+propositions.
+
+\subsection{Inductive-inductive universe}
+
+The reduction of the IIR universe to induction works, conceptually, as before.
 %
-In particular, in addition to $\ad{in-U}$ for defining $\U$ as before, we also
-introduce a family $\ad{in-U∼}$ of binary relations between types in the
-universe, from which we then define $\_\sim_{\setoidU}\_$.
+However, in addition to $\ad{in-U}$ for defining $\U$, we also introduce a
+family $\ad{in-U∼}$ of binary relations between types in the universe, from
+which we then define $\_\sim_{\setoidU}\_$.
 %

### 2020-05-28 paper.tex

@@
 
+\newcommand{\reflu}{\textsf{refl}_\setoidU}
+\newcommand{\symu}{\textsf{sym}_\setoidU}
+\newcommand{\transu}{\textsf{trans}_\setoidU}
+\newcommand{\reflel}{\textsf{refl}_\El}
+\newcommand{\symel}{\textsf{sym}_\El}
+\newcommand{\transel}{\textsf{trans}_\El}
+\newcommand{\coeel}{\textsf{coe}_\El}
+\newcommand{\cohel}{\textsf{coh}_\El}
+\newcommand{\equ}[2]{#1 \sim_\setoidU #2}
+\newcommand{\eqel}[3]{#1 \vdash #2 \sim_\El #3}
+
+\newcommand{\metaprop}{\mathbf{Prop}}
+
 \newcommand{\List}{\mathrm{List}}
@@
 \titlerunning{Constructing a universe for the setoid model}
-\authorrunning{Altenkirch, Kaposi, Sattler, Sestini}
+\authorrunning{Altenkirch, Boulier, Kaposi, Sattler, Sestini}
 
@@
 
+\todo{mention that everything takes place in a type theoretic metatheory, with
+  strict prop etc...}
+
 \subsection{Setoid Type Theory}
@@
 
-Technically speaking, SeTT is an extension of MLTT with extra component to
-support setoid reasoning. These extra components are nothing more than an
-internalization of the idea that every type is a setoid. More precisely, we have
-rules [explaining TODO] that contexts (aka closed types) are setoids, and that
-dependent types are setoids indexed over setoids.
+Technically, SeTT is an extension of MLTT with extra component to support setoid
+reasoning. These extra components are essentially an internalization of the idea
+that every type is a setoid. More precisely, we have rules [explaining TODO]
+that contexts (aka closed types) are setoids, and that dependent types are
+setoids indexed over setoids.
 
@@
 \infer{
-  \GG^\sim\ \rho_0\ \rho_1 : \Tm\ \DD\ \Prop
+  \GG^\sim\ \rho_0\ \rho_1 : \metaprop
 }{
@@
 \infer{
-  A^{\sim}\ p\ a_0\ a_1 : \Tm\ \DD\ \Prop
+  A^{\sim}\ \rho_{01}\ a_0\ a_1 : \Tm\ \DD\ \Prop
 }{
@@
   }{
-    A : \Ty\ \GG \qquad p : \Tm\ \DD\ (\El\ (\GG^{\sim}\ \rho_0\ \rho_1))
+    A : \Ty\ \GG \qquad \rho_{01} : \GG^{\sim}\ \rho_0\ \rho_1
   }
@@
 
-Note that the setoid equivalence relations are strict propositions. In addition,
-we also have rules representing the fact that every construction in SeTT
-respects setoid equality, so that we can transport along any such equality:
+%% Note that the setoid equivalence relations are strict propositions.
 
-[coe and coh rules here]
+In addition, we also have rules representing the fact that every construction in
+SeTT respects setoid equality, so that we can transport along any such equality:
 
@@
       \tyj{\GG}{A} \qquad \subj{\rho_0, \rho_1}{\GD}{\GG} \qquad
-      \tmj{\GD}{p}{\underline{\GG^\sim\ \rho_0\ \rho_1}}\qquad
+      p : \GG^\sim\ \rho_0\ \rho_1\qquad
       \tmj{\GD}{a}{A[\rho_0]}
@@
 identity type which is defined uniformly for any type.
-
+%
 Nevertheless, we can easily recover the usual ML identity type from setoid
@@
 As currently defined, SeTT is a system already powerful enough to support the
-formalization of part of set-level mathematics. However, we also would like to
-be able to equip SeTT with a universe of setoids. This universe must in turn be
-a setoid, hence in particular it cannot be univalent, but rather its equivalence
-relation must reflect simple equality of codes.
+formalization of part of set-level mathematics. Still, we would also like to
+equip SeTT with a universe of setoids. As any type in SeTT, this universe must
+in turn be a setoid. However, unlike the already existing universe of
+propositions, a universe of setoids cannot be univalent, as this would force it
+to be a groupoid. Instead, we must define its equivalence relation so that it
+only reflects a simple, syntactic equality of codes.
 
@@
 
-%% ----------------------
+\subsection{A first definition of the setoid universe}
 
-%% However, we don't want to assume inductive-recursive types in the basic type
-%% theory which is the target of the setoid translation. We know that we can translate
-%% basic instances of induction-recursion into inductive families using the
-%% equivalence of $I$-indexed families of sets and sets over $I$,
-%% i.e.\ $[I, \Set_i] \simeq \Set_i/I$ for $I : \Set_i$. For example, the
-%% inductive-recursive definition of a universe $\U:\Set_1$ and $\El : \U \to
-%% \Set$ with $\Pi$-types and Booleans is\footnote{In Agda, this definition would also go through if $\U$ were in $\Set$, but
-%%   this seems to be a non-conservative extension.}:
-%% %
-%% \begin{align*}
-%%   & \textsf{data}\ \U\ : \Set_1 \\
-%%   & \quad \textsf{bool} : \U \\
-%%   & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
-%%   & \El\ \textsf{bool} = \mathbbm{2} \\
-%%   & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
-%% \end{align*}
-
-%% We can model this as an inductive type $\ad{in-U}$ that \emph{carves out} all
-%% types in $\Set$ that are in the image of $\El$:
-%% %
-%% \begin{align*}
-%%   & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
-%%   & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
-%%   & \quad \textsf{inPi}
-%%   :  \{A : \Set\} \{B : A \to \Set\}\\
-%%   & \ \qquad \to \ad{in-U}\ A
-%%   \to ((a : A) \to \ad{in-U}\ (B\ a))
-%%   \to \ad{in-U}\ ((a : A) \to (B\ a))
-%% \end{align*}
-%% Using this $\U$ and $\El$ can be given as follows:
-%% \begin{align*}
-%%   \U & = \Sigma (A : \Set)\ (\textsf{in-U}\ A) \\
-%%   \El & = \textsf{proj}_1
-%% \end{align*}
-
-%% Note that this construction gives rise to a universe in
-%% $\Set_1$, rather than $\Set$, thus the meta-theory must contain at least one
-%% universe.
-
-\subsection{A first definition of a setoid universe}
-
-To provide such a universe we need to define a setoid $\setoidU : \Setoid$ and a
-family of setoids $\El : \setoidU \to \Setoid$, with codes for basic type
-formers like $\Pi$-types and booleans. That is, we need the following data
+To provide such a universe we need to first define a setoid for the universe,
+i.e. a set of codes together with an equivalence relation reflecting equality of
+codes. That is, we need the following data
 %
@@
   & \textsf{data} \ \setoidU : \Set_1 \\
-  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \Prop_1 \\
+  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \Prop_1
+  %% & \El : \setoidU \to \Set \\
+  %% & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
+\end{align*}
+%
+in addition to proofs that these are indeed equivalence relations
+%
+\begin{align*}
+  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
+  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
+  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
+\end{align*}
+
+Moreover, we need a way to reflect codes into types. Thus, we define a family of
+setoids indexed over the universe. This is given by the data below, that
+together assign to each code in the universe a carrier set and an equivalence
+relation.
+%
+\begin{align*}
   & \El : \setoidU \to \Set \\
@@
 
-The point of having $\setoidU$ and $\_\sim_\setoidU\_$ is quite clear: our
-setoid universe is a type, hence itself a setoid, hence described as any other
-setoid by a carrier set and an equivalence relation on it. These functions are
-achieved by $\setoidU$ and $\_\sim_{\setoidU}\_$, respectively.
+Note that $\eqel{}{}{}$ is indexed over equality on the universe, because $\El$
+is a setoid morphism, hence in particular it must respect the setoid equality of
+$\setoidU$.
+%
+We also require data and proofs that make sure we get setoids out of $\El$:
 
-For any code $A$ in the universe, we want a type $\El\ A$; this, in turn, should
-be a setoid as well. In other words, we want a family of setoids indexed over
-setoids. We thus need to explain, for any $A$, what is the carrier and
-equivalence relation of $\El\ A$. This is achieved by $\El : \setoidU \to \Set$,
-and $\_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to
-\El\ a \to \El\ a' \to \Prop$. Note that we need this to be indexed over code
-equality, because El is a type family indexed over U, and hence it must respect
-the equality on U. \todo{rewrite in better form}
+\begin{align*}
+  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
+  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
+  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
+  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
+  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
+\end{align*}
 
-\todo{Acknowledge Simon Boulier for the IIR universe construction}
 \todo{QIIT II LICS paper. this part is already in the HIIT paper}
 
-We given an inductive definition of $\setoidU, ...$, mutually with a recursive
-definition of $\El, \_\vdash\_\sim_\El\_, \sim_\setoidU, \textsf{refl}_\setoidU, ...$.
+We given an inductive definition of $\setoidU$, mutually with a recursive
+definition of everything else.
 
@@
 
-\[
-\textsf{refl}_\setoidU : ?
-\]
-
 Note that in the definition of $\setoidU$, we crucially require that the family
@@
 corresponding setoid equalities.
+
+An example of this is in the definition of $\El$:
+
+\begin{align*}
+  & \El\ \textsf{bool} :\equiv ? \\
+  & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
+  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
+  & \qquad \quad
+  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \to \eqel{h\ p}{f\ x}{f\ x'})
+\end{align*}
+
+We define a $\Pi$ type to be the set of functions that map equal elements to
+equal results. However, to state the latter property we need to compare elements
+in different types, coming from applying $f$ to different arguments $x$ and
+$x'$.  We know that $x$ and $x'$ are equal, but to conclude $\equ{B\ x}{B\ x'}$
+we need to know that $B$ respects setoid equality. This is exactly what we get
+from our definition of $\setoidU$.
 

### 2020-06-22 paper.tex
More paper writing

@@
 \usepackage{mathpartir}
+\usepackage{dsfont}
 \usepackage{proof}
@@
 \newcommand{\Set}{\textsf{Set}}
+\newcommand{\mType}{\mathbf{Type}}
 \newcommand{\Prop}{\textsf{Prop}}
+\newcommand{\mProp}{\mathbf{Prop}}
 \newcommand{\Setoid}{\textsf{Setoid}}
@@
 \newcommand{\Tm}{\mathrm{Tm}}
+\newcommand{\Sub}{\mathrm{Sub}}
 \newcommand{\Tms}{\mathrm{Tms}}
@@
 \newcommand{\El}{\textsf{El}}
+\newcommand{\Uty}{\mathds{U}}
+\newcommand{\EL}{\mathds{E}l}
 
@@
 \and
+  Simon Boulier
+\and
   Ambrus Kaposi \inst{2}
@@
 
-...
+A central aspect of Type Theory is the coexistence of two notions of
+equality. On the one hand definitional equality, the computational equality that
+is built into the formalism. On the other hand propositional equality, the
+internal notion of equality that is actually used to state and prove equational
+theorems within the system. The precise balance between these two notions is at
+the center of type theory research; however, it is generally understood that to
+properly support formalization of mathematics in type theory, one should aim for
+a definitional equality that is decidable and therefore machine-checkable, and a
+notion of propositional equality that is as extensional as possible.
 
-Intensional Type Theory has nice computational properties, like decidability of
-type-checking, that can make it more suitable for computer formalization, since
-proof-checking can be implemented as an entirely automatic procedure.
+Two extensionality principles seem particularly desirable, in that they arguably
+constitute the bare minimum for type theory to be comparable to set theory as a
+foundational system for set-level mathematics, in terms of power and ergonomics.
+One is function extensionality, according to which functions are equal if
+point-wise equal. Another is propositional extensionality, equating all
+propositions that are logically equivalent.
+
+Extensional Type Theory (ETT) does support extensional reasoning to some degree,
+thus seemingly being a convenient setting for formalization of
+mathematics. However, equality reflection makes type-checking ETT terms
+computationally unfeasible, if not impossible.
+
+On the other hand, Intensional Type Theory has nice computational properties
+that can make it more suitable for computer implementation, but as normally
+defined [cite ML paper?] it lacks the amount of extensionality necessary to
+effectively \emph{use} it.
 %
-However, a serious drawback of ITT is given by the lack of extensionality.
+It is known from model constructions that extensional principles like function
+extensionality (funext) are consistent with ITT.
 %
-Two extensionality principles are particularly missed from intensional type
-theory, in that they arguably constitute the bare minimum for type theory to be
-comparable to set theory as a foundational system for set-level mathematics.
+Moreover, ITT extended with uniqueness of identity proofs (UIP) and funext is
+known to be as powerful as ETT [cite hofmann]. These principles can be added as
+axioms, however destroying some computational properties of ITT such as
+canonicity.
+
+Ultimately, what is needed is a formulation of ITT that supports the
+aforementioned extensionality principles, while retaining convenient
+computational properties like canonicity.
 %
-The first of these is functional extensionality, i.e. the principle according to
-which two functions can be considered equal if they map equal inputs to equal
-outputs.
+Unfortunately, one can show that whenever identity types are defined according
+to Martin-Loef's inductive definition, there is no hope to have canonicity and
+funext coexist.
 %
-The second principle is propositional extensionality, according to which two
-propositions should be considered equal when logically equivalent.
-
-These extensionality principles are not theorems of intensional type theory as
-usually defined [cite], although they are consistent with it [cite].
-%
-As a consequence, these principles are often added as axioms when formalizing
-mathematics using proof-assistants based on ITT. This solution, however, is far
-from optimal, as the introduction of axioms destroys the nice computational
-properties of the system, particularly canonicity. Unfortunately, one can show
-that whenever equality is defined according to the usual inductive definition by
-Martin-Loef, there is no hope to have a theory where both canonicity and
-funext/propext coexist.
-
-The first step out of this is to give up the idea of equality as a single
+The first step towards a solution is to give up the idea of equality as a single
 inductive definition. Instead, equality should be defined specifically for each
-type former in the type theory. That is, every type former $A$ should be
-introduced together with its own notion of equality $A^\sim$.
-%
-If one does this, then funext can be obtained by simply defining equality on
-function types to be point-wise equality. That is, $(A \to B)^\sim\ f\ g :\equiv
-\forall (x : A) , B^\sim\ (f x)\ (g x)$.
+type former in the type theory. According to this principle, funext can be
+justified by simply defining equality on function types to be point-wise
+equality.
 
@@
 
-This form of extensional reasoning is already available in type theory via
-\emph{setoids}. A setoid is just a type $A$ (in this setting, we assume that all
-types are sets in the HoTT sense) together with a propositional equivalence
-relation $A^\sim : A \to A \to \Prop$.
+Sets, or types, together with a propositional equivalence relation as their
+notion of equality are known as \emph{setoids}. There are several definitions of
+the setoid model of Type Theory, the first of which was constructed to show
+consistency of funext for ITT [cite]. This consistency result is implied by the
+observation that UIP and funext hold in the setoid model. Moreover, it was shown
+by Altenkirch [cite] that is the model construction is carried out in a
+metatheory with a universe of strict (definitionally proof-irrelevant)
+propositions, it is possible to define a univalent universe of propositions,
+therefore satisfying propositional extensionality. The setoid model thus
+satisfies all the extensionality principles that we would like to extend ITT
+with. The question is whether there exists a version of ITT that supports setoid
+reasoning, and hence the forms of extensionality enables by it.
+
+This question was revisited and answered in Altenkirch et al. [cite]. In this
+paper, the authors define SeTT an extension of intensional MLTT + Prop with
+constructs for setoid reasoning, where funext and propext hold by
+definition. Moreover, they justify SeTT via a syntactic translation, after
+observing that this is made possible by the strictness of the model (every
+equation holds definitionally). This is in contrast with other type theories
+based on the setoid model, like OTT [cite] and XTT [cite], which instead rely on
+ETT as metatheory for their justification. A major contribution of SeTT is
+therefore to provide a way to bootstrap extensionality, by translating it into a
+small core intensional theory by means that do not require extensionality to
+begin with.
+
+The current formulation of SeTT is already a rich theory. However, its
+introspection capabilities are currently lacking, as they are limited to the
+univalent universe of propositions. We would like to internalise the notion of
+type in SeTT, thus extending the theory with a universe of setoids.
 %
-One can thus reason extensionally in ITT by defining an internal notion of
-setoid, and working with those instead of normal sets/types. Note that this
-approach only works provided one ensures that every construction performed on
-setoids respects the setoid relations involved. Unfortunately doing so is not an
-easy task, and most constructions beyond the more trivial ones will generate a
-great amount of proof obligations, especially when dependent setoids come into
-play, leading to a situation commonly known as \emph{setoid hell}.
+This brings up several questions, one of which has to do with the notion of
+equality with which the universe should come equipped. The universe of setoids
+is itself a setoid (as any type is), so it certainly cannot be a univalent
+universe, since a univalent setoid would be a groupoid rather than a setoid.
+%
+Another question asks how one can justify such universe in the setoid model. In
+particular what principles, are needed from the metatheory in order to define
+the universe.
 
-%% * Setoid TT: MLTT + setoid constructs
+\paragraph{Contributions}
 
-The good news is that all this work proving that our type theoretic
-constructions respect the setoid relations is not necessary, because these
-already respect setoid equality \emph{by construction}. More precisely, there
-exists a sound interpretation of MLTT (with pi, sigma, etc...) into setoids,
-that maps closed types/contexts to setoids and dependent types to indexed
-setoids.
-%
-Thus there is no need to explicitly work with setoids and setoid relations in
-TT, as one can simply carry out their constructions in plain MLTT and use the
-aforementioned interpretation to map that construction into the world of setoid.
+This paper documents our work towards the construction of a universe of setoids
+inside the setoid model, and tries to answer these and other questions related
+to the design and implementation of this construction. Our main contribution is
+the definition of the universe in the setoid model; this is initially given as
+an inductive-inductive-recursive definition, which is then translated firstly to
+an inductive-inductive definition, and secondly to a purely inductive type. As a
+consequence, we show that we only need to assume indexed W-types in the
+metatheory (along with some obligatory basic tools like $\Sigma$ and $\Pi$
+types) to justify the universe.
 
-This is interesting in principle, but practically useless. Since our
-constructions still take place in MLTT, we can't take advantage of the features
-of the setoid model our syntactic constructions are mapped to, since they are
-not accessible internall, from within the system itself.
-%
-This is where a key property of models comes into play. Models inform consistent
-extensions of the type theory, based on what holds true in the model, by
-reflecting the semanics into the syntax via additional syntactic constructs that
-are then mapped back to the original notions via the semantics.
+\paragraph{Structure of the paper}
 
-In our case, the existence of a setoid model suggests us ways to extend the
-syntax of MLTT with syntactic constructs that correspond to the setoid
-relations, allowing us to access and use them when proving theorems within the
-theory.
-%
-This is the idea behind Setoid Type Theory, SeTT for short.
+In Section~\label{sec:metatheory}...
 
-* Two incarnations of ITT that follow this approach are OTT and cubical TT. In OTT
-... TODO. In cubical TT ... TODO. Note that we do not include HoTT in the
-discussion since, although it is a version of ITT with extensional properties,
-it's not yet known to enjoy canonicity.
+\section{$MLTT^{Prop}$}\label{metatheory}
 
-... TODO ...
+This section describes $MLTT^{Prop}$, the metatheory in which we model
+construction takes place. It is crucial to keep the metatheory as minimal as
+possible.
 
-* universe of setoids
+$MLTT^{Prop}$ is an intensional MLTT with $\Pi, \Sigma, \textsf{Bool},
+\mathbf{0}, \mathbf{1}$, and a universe of strict propositions $\mProp$. We also
+have definitional $\eta$ rules for $\Pi, \Sigma$, and $\mathbf{1}$
+types. Propositions can be lifted to types via $\textsf{Lift} : \mProp \to
+\mType$. The universe of propositions is inhabited by false $\bot$ and
+propositional truncation $\|\_\| : \mType \to \mProp$. We also require indexed
+W-types, both in $\mType$ and $\mProp$:
 
-The current formulation of SeTT is already a satisfactory (\todo{maybe not the
-  right word: satisfactory to who? according to what metric?}) presentation of
-the setoid model, in that it includes all the features of the semantics that we
-wished to reflect in the syntax. It is also arguably expressive enough to
-formalize a good chunk of set-level mathematics. However, it is also currently
-lacking in a serious way, as the lack of universes seriously limits its
-introspection capabilities, and consequently its strength as a foundational
-system. Note that SeTT already includes a (univalent) universe of strict
-propositions; hence what we are referring to is a universe of setoids.
+\begin{align*}
+  W & : (S : \mType) \to (S \to \mType) \to \mType \\
+  W_p & : (S : \mType) \to (S \to \mType) \to \mProp
+\end{align*}
 
-The inclusion of a universe of setoids in SeTT poses several interesting
-questions. First of all, since a universe of setoids is itself a type in SeTT,
-is must also be a setoid. However, it certainly cannot be a univalent one. One
-question is therefore what kind of equality to equip such universe.
-%
-The second question is how one can justify such universe in the setoid model.
-In particular what principles, if any, should be added to the metatheory in
-order to be able to define the universe.
+We also require the propositional versions of $\mathbf{1}$, $\Pi$, and $\Sigma$,
+however these can be defined from their $\mType$ counterparts via truncation:
 
-This paper is an attempt to give an answer to the questions above.
-...TODO...
+\begin{align*}
+  \Pi_p\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
+  \Sigma_P\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
+  \mathbf{1}_p & :\equiv \| \mathbf{1} \|
+\end{align*}
+
+Note that a universe of strict propositions has been recently added to the Agda
+proof assistant [cite], thus making $MLTT^{Prop}$ a subset of Agda. In fact,
+most of the developments presented here have been formalized and proof-checked
+using Agda extended with strict propositions. For convenience we will employ
+Agda notation for metatheoretic constructions throughout the paper, with the
+implication that these can be understood as, or easily translated to, equivalent
+$MLTT^{Prop}$ definitions.
 
@@
 to T. Altenkirch \cite{setoid-model}.
+%
+The peculiarity of this model is that it is presented in a type theoretic and
+intensional metatheory, in our case $MLTT^{Prop}$.
+%
+The use of strict propositions to define the setoid equivalence relations makes
+this model \emph{strict}. A strict model is one where every equation holds
+definitionally. The setoid model thus defined validates extensionality
+principles like funext, proext, and quotient types. As a consequence, it
+provides a way to bootstrap and explain extensionality by implementing it in a
+core intensional type theory.
 
-The peculiarity of this model is that it is presented in a type theoretic and
-intensional metatheory. The metatheory is also very simple, requiring only $\Pi$
-types, $\Sigma$ types, and a universe of definitionally proof-irrelevant
-propositions itself closed under such type formers.
+As any model of Type Theory, the setoid model can be framed categorically as a
+category with families (CwF) with extra structure for the various type and term
+formers. The core structure can be defined as the following signature:
+
+\begin{align*}
+  \Con &: \mType \\
+  \Ty &: (\GG : \Con) \to \mType \\
+  \Sub &: (\GG \GD : \Con) \to \mType \\
+  \Tm &: (\GG : \Con) \to \Ty\ \GG \to \mType
+\end{align*}
+
+Contexts are given by setoids, that is types together with an equivalence relation.
+
+\[
+\infer{
+  \begin{aligned}
+    & | \GG | : \mType \\
+    & \GG^\sim : | \GG | \to | \GG | \to \mProp \\
+    & ...
+  \end{aligned}
+}{
+  \GG : \Con
+}
+\]
+
+Types in a context $\GG$ are given by setoid morphisms from $\GG$ to the
+(strict) category of setoids. The definition can be unfolded as follows:
+
+\[
+\infer{
+  \begin{aligned}
+    & | A | : | \GG | \to \mType \\
+    & A^\sim : \{\gamma_0\ \gamma_1 : | \GG |\}
+    \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
+    & \textsf{coe} : \{\gamma_0\ \gamma_1 : | \GG |\} \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
+    & \dots
+  \end{aligned}
+}{
+  A : \Ty\ \GG
+}
+\]
+
+This definition of types in the setoid model is different from the original one
+in [cite LICS 99 paper], but it is equivalent to it [cite].
 %
-Notably, the use of strict prop makes this model a strict one, namely one where
-every equation in the model is a definitional equality in the metatheory.  The
-model itself validates the extensionality principles mentioned above, like
-funext, proext, and quotient types. As a consequence, this model construction
-provides a way to explain extensionality, by giving a way to implement it from
-stratch in terms of a base theory that is both very simple and intensional.
+Substitutions are interpreted as functors between the corresponding setoids,
+whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
+fibration $\GG . A \to \GG$. Note that we only need to include components for
+the functorial action on objects and morphisms, since the functor laws follow
+from proof-irrelevance.
 
-This idea of implementing extensionality via this model can be made both more
-precise and more practical. Indeed, the strictness of this model implies that we
-can turn this model into a syntactic translation, from which a syntactic
-presentation of the model can be derived. (See next section, and [cite mpc'19])
-\todo{rewrite this.}
+\[
+\infer{
+  \begin{aligned}
+    & | \sigma | : | \GG | \to | \GD | \\
+    & \sigma^\sim : \forall\{\rho_0\, \rho_1\} \to \GG^\sim\, \rho_0\, \rho_1 \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
+  \end{aligned}
+}{
+  \sigma : \Sub\ \GG\ \GD
+}
+\]
 
-\todo{mention that everything takes place in a type theoretic metatheory, with
-  strict prop etc...}
+\[
+\infer{
+  \begin{aligned}
+    & | t | : (\gamma : | \GG |) \to | A | \, \gamma \\
+    & t^\sim : ...
+  \end{aligned}
+}{
+  t : \Tm\ \GG\ A
+}
+\]
+
+As defined, we can show that the setoid model validates the usual basic type
+formers ($\Pi, \Sigma$, etc.), including function extensionality and a universe
+of strict propositions satisfying propositional extensionality [cite LICS
+  99]. Note that we don't need neither identity types nor inductive types
+(W-types) in the metatheory to achieve this. However, W-types are needed when
+constructing the setoid universe in the next sections.
 
@@
 
-The setoid model has been recently adapted into Setoid Type Theory (SeTT), which
-is justified by a syntactic translation into a very basic \emph{target theory}
-\cite{mpc19}. This translation relies on the existence in the target theory of a
-universe of definitionally proof-irrelevant propositions $\Prop$, as recently
-implemented in Coq and Agda \cite{gilbert}. The setoid interpretation models a
-universe of propositions where equality of propositions is logical equivalence,
-thus providing a very basic instance of univalence.
+The setoid model presented in the previous section is peculiar in several
+ways. One is that it can be presented in a very small type theoretic
+metatheory. The other important aspect is that it is \emph{strict}, namely every
+equation of the original syntax holds by definition in the semantics. One
+advantage of strict models is that they can be turned into \emph{syntactic
+  translations}, in which syntactic objects of the source theory are interpreted
+as their counterparts in another \emph{target} theory. In the case of the setoid
+model, this gives rise to a \emph{setoid translation}, where source contexts are
+interpreted as target contexts together with a target type representing the
+equivalence relation, and so on.
 
-Technically, SeTT is an extension of MLTT with extra component to support setoid
-reasoning. These extra components are essentially an internalization of the idea
-that every type is a setoid. More precisely, we have rules [explaining TODO]
-that contexts (aka closed types) are setoids, and that dependent types are
-setoids indexed over setoids.
+A setoid translation is used in \cite{mpc19} to justify Setoid Type Theory
+(SeTT), an extension of $MLTT^{Prop}$ (without W-types) with equality types for
+contexts and dependent types that reflect the setoid equality of the model in
+the syntax.
+
+We recall the rules of SeTT that extend $MLTT^{Prop}$ below, but with a
+variation. Whereas the equality types in \cite{mpc19} are stated as elements of
+the universe of propositions internal to SeTT itself, we here state the context
+equalities as elements of the external, metatheoretic universe
+$\mProp$. Equality on dependent types is defined as before.
+
+We have a universe of propositions $\Prop$ defined as follows:
+
+\[
+\infer{
+  \Prop : \Ty\ \GG
+}{
+  \GG : \Con
+}
+\qquad
+\infer{
+  \underline{P} : \Ty\ \GG
+}{
+  P : \Tm\ \GG\ \Prop
+}
+\]
+
+Equality type constructors for contexts and dependent types internalize the idea
+that every context and type comes equipped with a setoid equivalence
+relation. As in the model, equality for dependent types is indexed over context
+equality.
 
@@
 
-%% Note that the setoid equivalence relations are strict propositions.
+We have rules witnessing that these are indeed equivalence relations. We only
+recall reflexivity:
+
+\[
+\infer{
+  \textsf{R}\ \rho : \GG^\sim\ \rho\ \rho
+}{
+  \rho : \Sub\ \GD\ \GG
+}
+\qquad
+\infer{
+  \textsf{R}\ a : \Tm\ \GG\ \underline{A^\sim\ (\textsf{R}\ \rho)\ a\ a}
+}{
+  A : \Ty\ \GG
+  \qquad
+  \rho : \Sub\ \GD\ \GG
+  \qquad
+  a : \Tm\ \GD\ A[\rho]
+}
+\]
 
@@
 
-Equality in SeTT differs from the standard approach by Martin-Loef in that every
-type comes equipped with an equality relation. This is in contrast with ML's
-identity type which is defined uniformly for any type.
+Notably, equality constructors in SeTT compute definitionally on concrete type
+formers. In particular, they compute to their obvious intended meaning, so that
+an equality of pairs is a pair of equalities, an equality of functions is a map
+of equalities, and so on. From this, we get definitional versions of function
+and propositional extensionality.
 %
-Nevertheless, we can easily recover the usual ML identity type from setoid
-equality:
+\begin{align*}
+  & (\Pi\ A\ B)^\sim\ p\ f_0\ f_1 = ... \\
+  & (\Sigma\ A\ B)^\sim\ p\ (a_0, b_0)\ (a_1, b_1) = ... \\
+  & \Prop^\sim\ p\ P\ Q = ...
+\end{align*}
+
+We can easily recover the usual Martin-L\"of identity type from setoid equality,
+with transport implemented via coercion.
 
@@
 \]
-
-Note that ... we have UIP... Moreover, contractibility of singletons holds by
-UIP and by definition of equality on $\Sigma$ types. Therefore we can define the
-$\J$ eliminator:
-
 \[
 \infer{
-  \tmj{\GG}{\J_M\ p\ d}{M[a_0, a_1, p]}
+  \textsf{transp}\ P\ p\ t :\equiv
+    \coe\ P\ (\textsf{R}\ \textsf{id} , p)\ t : \Tm\ \GG\ P[a_1]
 }{
-  \deduce{
-    \tmj{\GG}{d}{M[a_0, a_0, \refl_A\ a_0]}
+  P : \Ty\ (\GG . A)
   \qquad
-    \tmj{\GG}{p}{\underline{\Id_A\ a_0\ a_1}}
-  }{
-  \tyj{(\GG, x : A, y : A, p : \Id_A\ x\ y)}{M}
-  }
+  p : \Tm\ \GG\ (\Id\ A\ a_0\ a_1)
+  \qquad
+  t : \Tm\ \GG\ P[a_0]
 }
@@
 
+Note that contractibility of singletons holds in SeTT by UIP and the computation
+rules of equality on $\Sigma$ types. Hence, the J eliminator can also be
+derived.
+
 \section{Universe of setoids}
 
-\todo{Mention that the universe cannot be univalent}
+In its current form, SeTT is already rich and interesting in its own right.
+However, as pointed out in the introduction, it is also seriously limited by the
+lack of a universe internalizing the notion of types beyond propositions, that
+is, a universe of setoids. Our goal is to extend SeTT with such universe; since
+SeTT is a direct syntactic reflection of the setoid model, this essentially
+amounts to showing that a universe of setoids with the necessary structure and
+equations can be constructed within the setoid model. This opens several
+questions and possible design choices.
 
-\todo{Maybe mention the fact that it should be possible to give an elimination
-  principle for the setoid universe. This also makes sense since it's not univalent.}
+A first fundamental consideration has to with the very definition of the setoid
+universe: as any type in the setoid model, this universe must be a setoid and
+thus come equipped with an equivalence relation. However, unlike the already
+existing universe of propositions, a universe of setoids cannot be univalent,
+since this would force it to be a groupoid. Instead, we can define the relation
+so that it reflects a simpler syntactic equality of codes.
 
-As currently defined, SeTT is a system already powerful enough to support the
-formalization of part of set-level mathematics. Still, we would also like to
-equip SeTT with a universe of setoids. As any type in SeTT, this universe must
-in turn be a setoid. However, unlike the already existing universe of
-propositions, a universe of setoids cannot be univalent, as this would force it
-to be a groupoid. Instead, we must define its equivalence relation so that it
-only reflects a simple, syntactic equality of codes.
+Another question has to do with the metatheoretic tools required to carry out
+the construction of the universe. In fact, one of the main aspects of the setoid
+model construction recalled in Section TODO and shown originally in [cite LICS
+  99] is that it can be carried out in a very small type theoretic metatheory,
+thus providing a way to reduce extensionality to a small intensional core.
+%
+In extending the setoid model with a universe, we would like this property to
+still hold, hence it is of primary concern to reduce to a minimum the extensions
+required in the metatheory, so to keep it as small as possible.
 
-As we use Type Theory as our metatheory, this universe of setoids is going to be
-an internal construction in Type Theory. The usual way to encode universes in
-Type Theory is by an inductive-recursive definition. As we will see in Section
-TODO, it is indeed possible to give a IR definition of a universe of setoids.
-%
-However, we would prefer not to be forced to assume inductive-recursive types,
-or IR in general, in the basic type theory which is the target of the setoid
-translation. This is because the point of the translation is to explain a
-complex system in terms of a very simple, well understood core, and we think
-introducing IR would go against this goal.
+A known and established method for constructing universes in Type Theory uses
+induction-recursion (IR), a definition schema developed by Dybjer
+[cite]. Inductive-recursive definitions can be found throughout the literature
+on type theory, from the already mentioned type theoretic universes, including
+the original formulation a la Tarski by Martin-L\"of [cite], to metamathematical
+tools like computability predicates [cite].
 
-Fortunately, it is possible to reduce some instances of induction-recursion to
-plain induction (more specifically, inductive families), and the well known IR
-universe construction is indeed one of them.
+Although universe constructions in Type Theory---including our own setoid
+universe---are naturally presented as inductive-recursive definitions, they
+may not necessarily require a metatheory that supports induction-recursion. In
+fact, it is possible to reduce some instances of induction-recursion to plain
+induction (more specifically, inductive families), including some universe
+definitions. We recall this reduction in Section TODO.
 
-In the next section, we recall the IR construction of universes in type theory,
-as well as lay out the process by which such construction can be turned to one
-that only relies on inductive families. In Section TODO, we give a first IR
-definition of the universe of setoids. After that we will see if and how such IR
-definition can be turned into one that does not rely on IR.
+%% However, we would prefer not to be forced to assume inductive-recursive types,
+%% or IR in general, in the basic type theory which is the target of the setoid
+%% translation. This is because the point of the translation is to explain a
+%% complex system in terms of a very simple, well understood core, and we think
+%% introducing IR would go against this goal.
+
+Other design choices on the setoid universe are less essential, but still
+require careful consideration. For instance, one question is whether the setoid
+universe should support elimination, thus exposing the inductive structure of
+the codes. Universe elimination is known to be inconsistent with univalence. As
+it turns our, univalence is not an issue in our case; nevertheless it is not
+immediately clear if the elimination principle can be justified by the
+semantics.
+
+Another choice is whether to include codes for other universes, besides the
+usual type formers. One could have $\Prop$ be a code in the setoid universe,
+rather than a type. Moreover, the universes could form a hierarchy, where the
+universe at level $i$ is an element of the universe of level $i+1$. Another
+option is whether to make the hierarchy cumulative.
+
+%% In the next section, we recall the inductive-recursive construction of universes
+%% in type theory, as well as lay out the process by which such construction can be
+%% turned to one that only relies on inductive families. In Section TODO, we give a
+%% first inductive-recursive definition of the universe of setoids. After that we
+%% will see if and how such IR definition can be turned into one that does not rely
+%% on IR.
 
@@
 
-A standard, well-known way of encoding universes in type theory is via induction-recursion.
-\todo{a couple of words about induction recursion}.
+An inductive-recursive universe is given by a type of codes $\U$, and a family
+$\El : \U \to \mType$ that assigns, to each code corresponding to some type, the
+meta-theoretic type of its elements. The resulting definition is
+inductive-recursive because the inductive type of codes is defined
+simultaneously with the recursive function $\El$.
 
-To define a universe, we simultaneously define the universe of codes itself, as
-an inductive definition, and a family $\El : \U \to \Set$ that assigns, to each
-code corresponding to some type, the meta-theoretic type of its elements.
-
-For example, the definition of a small universe with booleans and $\Pi$ types
-would be the following:
+As an example, consider the following definition of a small universe with
+booleans and $\Pi$ types.
 
 \begin{align*}
-  & \textsf{data}\ \U\ : \Set_1 \\
+  & \textsf{data}\ \U\ : \mType_1 \\
   & \quad \textsf{bool} : \U \\
@@
 
-Note that $\U$ is defined to be in $\Set_1$ rather than $\Set$. This is
-because... \todo{TODO}
+Note that $\U$ is defined to be in $\mType_1$ rather than $\mType$. This is
+because... \todo{TODO add explanation}. This fact also implies that the
+meta-theory must contain at least one universe.
 
-As it turns out, induction-recursion is not really required to encode such
-a universe, although is certainly makes the definition particularly nice.
+Induction-recursion is arguably the nicest and more natural way to define
+internal universes in Type Theory, however it is not always required.
 %
-In fact, we know that we can translate basic instances of induction-recursion
-into inductive families using the equivalence of $I$-indexed families of sets
-and sets over $I$, i.e.\ $[I, \Set_i] \simeq \Set_i/I$ for $I : \Set_i$.
+We can translate basic instances of induction-recursion into inductive families
+using the equivalence of $I$-indexed families of sets and sets over $I$,
+i.e.\ $[I, \mType_i] \simeq \mType_i/I$ for $I : \mType_i$.
 
@@
 \begin{align*}
-  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
+  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
   & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
   & \quad \textsf{inPi}
-  :  \{A : \Set\} \{B : A \to \Set\}\\
+  :  \{A : \mType\} \{B : A \to \mType\}\\
   & \ \qquad \to \ad{in-U}\ A
@@
 \begin{align*}
-  \U & = \Sigma (A : \Set)\ (\textsf{in-U}\ A) \\
+  \U & = \Sigma (A : \mType)\ (\textsf{in-U}\ A) \\
   \El & = \textsf{proj}_1
@@
 
-Note that this construction gives rise to a universe in $\Set_1$, rather than
-$\Set$, since the definition of $\U$ quantifies over all possible sets. Thus,
-the meta-theory must contain at least one universe for this construction to go
-through.
+Note that this construction gives rise to a universe in $\mType_1$, rather than
+$\Set$, since the definition of $\U$ quantifies over all possible sets.
 
@@
 
-To provide such a universe we need to first define a setoid for the universe,
-i.e. a set of codes together with an equivalence relation reflecting equality of
-codes. That is, we need the following data
+In this section we give a first definition of the setoid universe, as a direct
+generalization of the simple inductive-recursive definition just shown. We only
+consider a very small universe with unit type $\mathds{1}$ and $\Pi$ for
+simplicity; a more realistic universe can be found in the Agda formalization,
+and includes more type formers.
+
+To construct the universe of setoids in the setoid model, we first of all need
+to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
+\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling Section TODO,
+these are essentially record types made of several Since. components $\Uty$ is a
+closed type, it requires the same data of a setoid; in particular, we need a set
+of codes together with an equivalence relation reflecting equality of codes
 %
 \begin{align*}
-  & \textsf{data} \ \setoidU : \Set_1 \\
-  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \Prop_1
+  & \textsf{data} \ \setoidU : \mType_1 \\
+  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
   %% & \El : \setoidU \to \Set \\
@@
 
-Moreover, we need a way to reflect codes into types. Thus, we define a family of
-setoids indexed over the universe. This is given by the data below, that
-together assign to each code in the universe a carrier set and an equivalence
-relation.
+To define $\EL$, we need a family of setoids indexed over the universe. In other
+words, a way to assign to each code in the universe a carrier set and an
+equivalence relation.
 %
@@
 
-\todo{QIIT II LICS paper. this part is already in the HIIT paper}
+%% \todo{QIIT II LICS paper. this part is already in the HIIT paper}
 
-We given an inductive definition of $\setoidU$, mutually with a recursive
-definition of everything else.
-
+We give an inductive definition of $\setoidU$, mutually with a recursive
+definition of everything else. Below we only show the more interesting definitions of $\setoidU$ and $\El$:
+%
 \begin{align*}
   & \textsf{data}\ \setoidU : \Set_1 \\
-  & \quad \textsf{bool} : \setoidU \\
+  & \quad \textsf{one} : \setoidU \\
   %% Πs : (a : U i)(b : El a → U i)
@@
 \end{align*}
-
-Note that in the definition of $\setoidU$, we crucially require that the family
-$B : \El\ A \to \setoidU$ be a setoid morphism, hence that it respects the
-corresponding setoid equalities.
-
-An example of this is in the definition of $\El$:
-
+%
 \begin{align*}
-  & \El\ \textsf{bool} :\equiv ? \\
+  & \El\ \textsf{one} :\equiv \textbf{1} \\
   & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
@@
 
-We define a $\Pi$ type to be the set of functions that map equal elements to
-equal results. However, to state the latter property we need to compare elements
-in different types, coming from applying $f$ to different arguments $x$ and
-$x'$.  We know that $x$ and $x'$ are equal, but to conclude $\equ{B\ x}{B\ x'}$
-we need to know that $B$ respects setoid equality. This is exactly what we get
-from our definition of $\setoidU$.
+Note that in the definition of $\setoidU$ we require that the family $B : \El\ A
+\to \setoidU$ be a setoid morphism, respecting the setoid equalities involved.
+%
+This choice is crucial for the definition of $\El$ to go through, in particular
+since we eliminate the code for $\Pi$ types into the setoid of functions that
+map equal elements to equal results. To state this mapping property we need to
+compare elements in different types, coming from applying $f$ to different
+arguments $x$ and $x'$. We know that $x$ and $x'$ are equal, but to conclude
+$\equ{B\ x}{B\ x'}$ we need to know that $B$ respects setoid equality. This is
+exactly what we get from our definition of $\setoidU$.
 
-This definition of the setoid universe relies on both induction-recursion and
-induction-induction. Although appily accepted by Agda, the meaning of this
-combination is not really understood. The goal of the next sections is therefore
-to find an alternative definition of the setoid universe that relies on
-principles that are better understood. Not only that, we would like to reduce
-the number of extensions to the metatheory that are needed to be able to give
-such definition, so to stay as close as possible to our initial idea of reducing
-SeTT to a small core.  \todo{reword}
+\todo{why is this inductive-inductive-recursive?}
 
-The next section explains our main result, which is that a modified form of the
-translation of IR to IF also works for the setoid universe defined above. This
-type is more complex that the basic IR types we have seen before, in that it
-uses a combination of IR and II, and includes the universe of strict
-propositions.
+%% The above definition of the setoid universe relies on both induction-recursion
+%% and induction-induction. Although happily accepted by Agda, we don't fully
+%% understand the meaning of this combination, so we would like to find an
+%% equivalent definition that does not rely on it. The goal of the next sections is
+%% therefore to get to such an alternative definition. This is done by first
+%% transforming the inductive-inductive-recursive definition to one that doesn't
+%% use induction-recursion, and secondly by reducing the induction-induction to
+%% inductive families.
+
+%% Our transformation of the inductive-inductive-recursive definition is inspired
+%% by the well-known trick described in Section X, but modified in a novel way to
+%% account for the presence of induction-induction and $\mProp$-valued types. To
+%% our knowledge, this is the first time this reduction trick is applied to an
+%% inductive-recursive type of this complexity.
 
@@
 
-Note that $\ad{in-U}$ appears in the sort of $\ad{in-U∼}$, thus making this
-definition as instance of induction-induction [cite]. \todo{add brief
-  explanation of induction-induction}.
+Note that $\ad{in-U}$ appears in the sort of $\ad{in-U∼}$. This is an instance
+of \emph{induction-induction}, a schema that allows the definition of a type
+mutually with other types that contain the first one in their signature [cit].
 
-As in the previous example, we can now define $\setoidU$ as a $\Sigma$ type, and
-$\El$ as the corresponding first projection.
+As in the universe example in Section TODO, we now define $\setoidU$ as a
+$\Sigma$ type, and $\El$ as the corresponding first projection.
 
 \begin{align*}
-  & \setoidU : \Set_1 \\
-  & \setoidU = \Sigma\ (X : \Set)\ (\ad{in-U}\ X) \\
-  & \El : \setoidU \to \Set \\
+  & \setoidU : \mType_1 \\
+  & \setoidU = \Sigma\ (X : \mType)\ (\ad{in-U}\ X) \\
+  & \El : \setoidU \to \mType \\
   & \El = \textsf{proj}_1
@@
 %
-The definition of $\_\sim_{\setoidU}\_$ follows, in principle, that of
-$\setoidU$. A ...
-
-Similarly, two codes $A, B$ in the universe $\setoidU$ are equal when there
-exists a setoid equivalence relation on their respective sets $\El\ A$ and $\El
-B$. Intuitively, since elements of a setoid are only ever compared to elements
-of the same setoid, this is only possible if $A, B$ are codes for the same
-setoid, that is, if $A \sim_{\setoidU} B$. Existence and well-formedness of such
-relations is expressed via the type $\ad{in-U~}$ just defined, hence formally we
-should have something like:
+Two codes $A, B$ in the universe $\setoidU$ are equal when there exists a setoid
+equivalence relation on their respective sets $\El\ A$ and $\El B$. Intuitively,
+since elements of a setoid are only ever compared to elements of the same
+setoid, this should only be possible if $A, B$ are codes for the same setoid,
+that is, if $A \sim_{\setoidU} B$. Existence and well-formedness of such
+relations is expressed via the type $\ad{in-U~}$ just defined, hence we would
+expect $A \sim_{\setoidU} B$ to be defined as follows:
 
 \[
-A \sim_{\setoidU} B :\equiv \Sigma (R: ...) ...
+A \sim_{\setoidU} B :\equiv \Sigma (R: \El\,A \to \El\,B \to \mProp)(\ad{in-U∼}\ (\pi_2\,A)\ (\pi_2\,B)\ R)
 \]
@@
 \[
-(A, p) \sim_{\setoidU} (A', p') :\equiv \Sigma (R: ...) ...
+(A, p_A) \sim_{\setoidU} (B, p_B) :\equiv \Sigma (R: A \to B \to \mProp)(\ad{in-U∼}\ p_A\ p_B\ R)
 \]
 
-Unfortunately this definition only captures the idea we are aiming for, but does
-not actually typecheck. In fact, $\_\sim_{\setoidU}\_$ is a $\Prop_1$-valued
-relation, so $A \sim_{\setoidU} B$ should be a proposition, but $\Sigma (R: ...)
-...$ clearly isn't.
-
-The solution is actually quite simple, and it just involves truncating the
-$\Sigma$ type described above to force it to be in $\Prop_1$. More precisely, we
-assume to have in our metatheory, in addition to a hierarchy of universes of
-strict propositions $\Prop_i$, a truncation operation $\parallel\_\parallel :
-\Set_i \to \Prop_i$ with constructor $|\_| : \{A : \Set_i\} \to A \to \parallel
-A \parallel$ and eliminator $\ad{withTrunc} : \{B : \Prop_i\} \to (A \to B) \to
-\parallel A \parallel \to B$. Note that truncation can be defined in Agda
-extended with the universe of strict propositions, with the eliminator replaced
-by pattern matching.
-
+Unfortunately this definition only manages to capture the idea, but does not
+actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be $\Prop_1$-valued
+relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
+$\Sigma$ type shown above clearly isn't, since it quantifies over a type of
+relations, which is not a proposition.
+%
+One possible solution is actually quite simple, and it just involves truncating
+the $\Sigma$ type above to force it to be in $\Prop_1$.
+%
 \begin{align*}
@@
 
-This trick might seem to simple to the point of being naive, but as we will see
-it is sufficient to make the entire definition go through.
-
-We are now left to define $\_\vdash\_\sim_{\El}\_$.
-
+We are now left to define the indexed equivalence relation on $\El$:
+%
 \begin{align*}
-  & \_\vdash\_\sim_{\El}\_ : \{a\ a' : \U\} \to a\ \sim_{\setoidU}\ a' \to \El\ a \to \El\ a' \to \Prop
+  & \_\vdash\_\sim_{\El}\_ : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
+  & p \vdash a \sim_{\El} b :\equiv \ ?
 \end{align*}
 
-Intuition would suggest to define $\_\vdash\_\sim_{\El}\_$ by projecting
-the relation out of the proof of $A \sim_{\setoidU} A'$, in much the same way as
-$\El$ is defined by projecting out of $\U$. That is, $p \vdash x \sim_{\El} y =
-\ad{proj}_1\ p\ x\ y$.
+In the definition above, $p$ has type $\|| \Sigma (R: \El\,A \to \El\,B \to
+\mProp)(...) \||$. If the type wasn't propositionally truncated, we could define
+$p \vdash a \sim_{\El} b$ by extracting the relation out of the first component
+of $p$, and apply it to $a, b$. That is, $p \vdash a \sim_{\El} b :\equiv
+\pi_1\,p\,a\,b$. This would make the definition of $\_\sim_\setoidU\_$ and
+$\_\vdash\_\sim_{\El}\_$ in line with how we defined $\setoidU$ and $\El$.
+
+However, this doesn't work in our case, since the type of $p$ \emph{is}
+propositionally truncated, hence it cannot be eliminated to construct a
+proof-relevant object.
 %
-However, this doesn't work since the type of $p$ is propositionally truncated,
-hence it cannot be used to construct a proof-relevant object. In other words,
-the proof $p$ cannot be eliminated into $El A \to El A' \to Prop$, because
-$El A \to El A' \to Prop$ is not a proposition, but a $\Set$.
+Fortunately, we can work around this limitation by defining $p \vdash x
+\sim_{\El} y$ by induction on the codes $A\ B : \setoidU$, in a way that ends
+up being logically equivalent to the proposition we would have obtained by
+$\ad{proj}_1\ p\ x\ y$ if there were no truncation.
+%
+More precisely, we need to construct proofs that for any concrete $R$ and
+$\ad{inR}$, the types $| (R, \ad{inR}) | \vdash x \sim_{\El} y$ and $R\ x\ y$
+are logically equivalent. These in turn need to be defined mutually with
+$\_\vdash\_\sim_{\El}\_$. We direct the interested reader to the Agda
+formalization for the full details of these definitions, as they are quite
+involved.
 
-We can work around this limitation by defining $p \vdash x \sim_{\El} y$ by
-induction on the codes $a\ a' : \setoidU$, in a way that is logically equivalent
-to the proposition we would have obtained by $\ad{proj}_1\ p\ x\ y$ if that
-expression was well-typed.
+The full definition of the universe is concluded with the remaining definitions,
+like $\reflu, \reflel$, etc., which can be adapted from their IR counterparts
+more or less straighforwardly. The final result does not use
+induction-recursion, but it's nevertheless an instance of infinitary
+induction-induction. The ability to define arbitrary, infinitary
+inductive-inductive types clashes with our objective of keeping the metatheory
+as small and simple as possible. The next step is therefore to reduce this
+inductive-inductive universe to one that does not require (infinitary)
+induction-induction.
 
-Note that for this definition to go through, we sometimes need the fact that the
-relation we are building is indeed equivalent to the one we would have obtained
-if we could project out of the truncation. That is, we need an additional proof
-that for any concrete $R$ and $\ad{inR}$, $| (R, \ad{inR}) | \vdash x \sim_{\El}
-y \iff R\ x\ y$. We thus need to define $\_\vdash\_\sim_{\El}\_$ mutually with a proof
-of the following
+It is known that finitary inductive-inductive definitions can be translated into
+inductive families \cite{iit-erasure,iit-to-ix,induction-is-enough} but it is
+not clear whether this extends to an infinitary type like our universe.
+%
+Of course, the absence of a general reduction method does not mean that
+particular concrete instances of induction-induction cannot be reduced to
+simpler principles, which is exactly what we hope for our universe construction.
 
-\[
-...
-\]
+\section{Indexed inductive universe (maybe?)}
 
-We know that finitary inductive-inductive definitions can be translated into
-inductive families \cite{iit-erasure,iit-to-ix,induction-is-enough} but it is
-not clear whether this construction extends to an infinitary type like the one
-above. This is subject of further work. If successful we would be able to give a
-translation of the setoid model with a universe into a very basic core type
-theory.
+TODO: depends on whether we manage to finish the formalization.
+
+\section{Related work}
+
+TODO
+
+\section{Further work}
+
+TODO
 

### 2020-06-24 paper.tex
More paper writing

@@
 equality. On the one hand definitional equality, the computational equality that
-is built into the formalism. On the other hand propositional equality, the
+is built into the formalism. On the other hand ``propositional'' equality, the
 internal notion of equality that is actually used to state and prove equational
@@
 properly support formalization of mathematics in type theory, one should aim for
-a definitional equality that is decidable and therefore machine-checkable, and a
-notion of propositional equality that is as extensional as possible.
+a notion of definitional equality that is decidable and therefore
+machine-checkable, and a notion of propositional equality that is as as
+extensional possible.
 
@@
 foundational system for set-level mathematics, in terms of power and ergonomics.
-One is function extensionality, according to which functions are equal if
-point-wise equal. Another is propositional extensionality, equating all
-propositions that are logically equivalent.
+One is function extensionality (funext), according to which functions are equal
+if point-wise equal. Another is propositional extensionality (propext), equating
+all propositions that are logically equivalent.
 
 Extensional Type Theory (ETT) does support extensional reasoning to some degree,
-thus seemingly being a convenient setting for formalization of
-mathematics. However, equality reflection makes type-checking ETT terms
+and thus would seem a convenient setting for formalization of
+mathematics. However, ETT's equality reflection makes type-checking ETT terms
 computationally unfeasible, if not impossible.
 
-On the other hand, Intensional Type Theory has nice computational properties
-that can make it more suitable for computer implementation, but as normally
-defined [cite ML paper?] it lacks the amount of extensionality necessary to
-effectively \emph{use} it.
+On the other hand, Intensional Type Theory (ITT) has nice computational
+properties that can make it more suitable for computer implementation, but as
+normally defined [cite Martin-L\"of paper?] it lacks extensionality.
 %
-It is known from model constructions that extensional principles like function
-extensionality (funext) are consistent with ITT.
+It is known from model constructions that extensional principles like funext are
+consistent with ITT.
 %
@@
 
-Ultimately, what is needed is a formulation of ITT that supports the
+What we would like instead is a formulation of ITT that supports the
 aforementioned extensionality principles, while retaining convenient
-computational properties like canonicity.
+computational properties like canonicity and decidability of type checking.
 %
@@
 %
-The first step towards a solution is to give up the idea of equality as a single
-inductive definition. Instead, equality should be defined specifically for each
-type former in the type theory. According to this principle, funext can be
-justified by simply defining equality on function types to be point-wise
-equality.
+The first step towards a solution is to give up the idea of propositional
+equality as single type family inductively generated by reflexivity. Instead,
+equality should be \emph{specific} to each type former in the type theory, or in
+other words, every type former should be introduced alongside an explanation of
+what counts as equality for its elements. In this setting, function
+extensionality can simply be introduced as the chosen notion of equality on
+function types.
 
-%% * One possible answer: setoids
-
-Sets, or types, together with a propositional equivalence relation as their
-notion of equality are known as \emph{setoids}. There are several definitions of
-the setoid model of Type Theory, the first of which was constructed to show
-consistency of funext for ITT [cite]. This consistency result is implied by the
-observation that UIP and funext hold in the setoid model. Moreover, it was shown
-by Altenkirch [cite] that is the model construction is carried out in a
-metatheory with a universe of strict (definitionally proof-irrelevant)
-propositions, it is possible to define a univalent universe of propositions,
-therefore satisfying propositional extensionality. The setoid model thus
-satisfies all the extensionality principles that we would like to extend ITT
-with. The question is whether there exists a version of ITT that supports setoid
-reasoning, and hence the forms of extensionality enables by it.
+This conception of types together with their notion of equality goes back to the
+notion of \emph{setoids} or \emph{Bishop sets}. A setoid is simply a set
+equipped with a propositional equivalence relation. Setoids provide a quite
+natural and useful semantic domain in which to interpret type theory. The first
+setoid model was constructed to show consistency of funext for ITT [cite]. This
+consistency result is implied by the observation that UIP and funext hold in the
+setoid model. Moreover, it was shown by Altenkirch [cite] that is the model
+construction is carried out in a metatheory with a universe of strict
+(definitionally proof-irrelevant) propositions, it is possible to define a
+univalent universe of propositions, therefore satisfying propositional
+extensionality. The setoid model thus satisfies all the extensionality
+principles that we would like to extend ITT with. The question is whether there
+exists a version of ITT that supports setoid reasoning, and hence the forms of
+extensionality enables by it.
 
@@
 
-This section describes $MLTT^{Prop}$, the metatheory in which we model
-construction takes place. It is crucial to keep the metatheory as minimal as
-possible.
+This section describes $MLTT^{Prop}$, the metatheory in which the model
+construction described in the following sections takes place. One aspect of the
+metatheory that we pay attention to is it's minimality, as we want to keep it as
+simple as possible. We want to particularly avoid having to assume strong
+definition schemas that go beyond inductive families (or, semantically, what can
+be modeled with W types.)
 
-$MLTT^{Prop}$ is an intensional MLTT with $\Pi, \Sigma, \textsf{Bool},
-\mathbf{0}, \mathbf{1}$, and a universe of strict propositions $\mProp$. We also
-have definitional $\eta$ rules for $\Pi, \Sigma$, and $\mathbf{1}$
-types. Propositions can be lifted to types via $\textsf{Lift} : \mProp \to
-\mType$. The universe of propositions is inhabited by false $\bot$ and
-propositional truncation $\|\_\| : \mType \to \mProp$. We also require indexed
-W-types, both in $\mType$ and $\mProp$:
+$MLTT^{Prop}$ is an intensional type theory in the style of Martin-L\"of Type
+Theory with $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$, and a universe
+of strict propositions $\mProp$. $\Sigma$ types are defined negatively by
+pairing $\_,\_$ and projections $\pi_1, \pi_2$. We also have definitional $\eta$
+rules for $\Pi, \Sigma$, and $\mathbf{1}$ types. Propositions can be lifted to
+types via $\textsf{Lift} : \mProp \to \mType$. The universe of propositions is
+inhabited by false $\bot$ and propositional truncation $\|\_\| : \mType \to
+\mProp$. We also require indexed W-types, both in $\mType$ and $\mProp$:
 
@@
 
-We also require the propositional versions of $\mathbf{1}$, $\Pi$, and $\Sigma$,
-however these can be defined from their $\mType$ counterparts via truncation:
+In addition to the proof-relevant types just listed, we will need the
+propositional versions of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can
+be defined from their $\mType$ counterparts via truncation. That is, given $P :
+\mProp$ and $Q : P \to \mProp$:
 
@@
     & A^\sim : \{\gamma_0\ \gamma_1 : | \GG |\}
-    \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
+    \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
     & \textsf{coe} : \{\gamma_0\ \gamma_1 : | \GG |\} \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
@@
 
-An inductive-recursive universe is given by a type of codes $\U$, and a family
-$\El : \U \to \mType$ that assigns, to each code corresponding to some type, the
-meta-theoretic type of its elements. The resulting definition is
+An inductive-recursive universe is given by a type of codes $\U : \mType$, and a
+family $\El : \U \to \mType$ that assigns, to each code corresponding to some
+type, the meta-theoretic type of its elements. The resulting definition is
 inductive-recursive because the inductive type of codes is defined
@@
 \begin{align*}
-  & \textsf{data}\ \U\ : \mType_1 \\
+  & \textsf{data}\ \U\ : \mType \\
   & \quad \textsf{bool} : \U \\
@@
 
-Note that $\U$ is defined to be in $\mType_1$ rather than $\mType$. This is
-because... \todo{TODO add explanation}. This fact also implies that the
-meta-theory must contain at least one universe.
+%% Note that $\U$ is defined to be in $\mType_1$ rather than $\mType$. This is
+%% because... \todo{TODO add explanation}.
 
@@
 We can translate basic instances of induction-recursion into inductive families
-using the equivalence of $I$-indexed families of sets and sets over $I$,
-i.e.\ $[I, \mType_i] \simeq \mType_i/I$ for $I : \mType_i$.
+using the equivalence of $I$-indexed families of types and types over $I$
+(that is, $\mType_i \to I$).
 
@@
 Note that this construction gives rise to a universe in $\mType_1$, rather than
-$\Set$, since the definition of $\U$ quantifies over all possible sets.
+$\mType$, since the definition of $\U$ quantifies over all possible sets. This
+also means that a metatheory with at least one universe is required for this
+construction.
 
@@
 \Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling Section TODO,
-these are essentially record types made of several Since. components $\Uty$ is a
-closed type, it requires the same data of a setoid; in particular, we need a set
-of codes together with an equivalence relation reflecting equality of codes
+these are essentially record types made of several components. Since $\Uty$ is a
+closed type, it requires the same data of a setoid; in particular, we need a type
+of codes together with an equivalence relation reflecting equality of codes:
 %
@@
 
-Note that $\eqel{}{}{}$ is indexed over equality on the universe, because $\El$
-is a setoid morphism, hence in particular it must respect the setoid equality of
-$\setoidU$.
+Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
+$\El$ is a setoid morphism, hence in particular it must respect the setoid
+equality of $\setoidU$.
 %
@@
 
-\todo{why is this inductive-inductive-recursive?}
+[todo... explanation]
+%
+\begin{align*}
+  & | \Uty | :\equiv \lambda\, \gamma . \, \setoidU \\
+  & \Uty^\sim :\equiv \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
+  & \textsf{refl}\ \Uty :\equiv \reflu \\
+  & ... \\
+  & \textsf{coe}\ \Uty :\equiv \lambda\,p\,a . \, a \\
+  & \textsf{coh}\ \Uty :\equiv \lambda\,p . \, \reflu
+\end{align*}
+
+Given $A : \Tm\ \GG\ \Uty$, we define $\EL\ A$ as follows:
+%
+\begin{align*}
+  & | \EL\ A | :\equiv \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
+  & (\EL\ A)^\sim :\equiv \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
+  & \textsf{refl}\ (\EL\ A) :\equiv \reflel \\
+  & ... \\
+  & \textsf{coe}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
+  & \textsf{coh}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
+\end{align*}
+
+We can show that $\Uty$ is closed under $\Pi$ types and \textsf{Bool}, including
+$\beta$ and $\eta$ laws. Moreover, all the necessary equations hold
+definitionally in the metatheory.
+
+This gives a complete definition of a universe of setoids. However, as already
+pointed out, this definition uses induction-recursion. Moreover, the kind of
+recursion involved in this definition is particularly complex, since it involves
+several recursive definitions whose types contain mutually defined functions. It
+is not clear whether this definition schema, which we could call
+induction-recursion-recursion, can be reduced to simpler, better understood
+notions of induction-recursion like in [dybjer-setzer]. Nevertheless, even the
+need to extend the metatheory with simple forms of induction-recursion goes
+against our objective to keep the metatheory as small and essential as possible.
+
+The goal of the next section is therefore to transform the
+inductive-recursvie-recursive definition ourlined above to one that doesn't use
+induction-recursion.
+%
+This transformation is inspired by the well-known trick to eliminate
+induction-recursion described in Section X, but modified in a novel way to
+account for the presence of $\mProp$-valued types. To our knowledge, this is the
+first time this reduction trick is applied to an inductive-recursive type of
+this kind.
 
@@
 
-The reduction of the IIR universe to induction works, conceptually, as before.
+The reduction of the IR universe to induction works, conceptually, as in the
+small example of Section X.
 %
@@
 
-Note that $\ad{in-U}$ appears in the sort of $\ad{in-U∼}$. This is an instance
-of \emph{induction-induction}, a schema that allows the definition of a type
-mutually with other types that contain the first one in their signature [cit].
+Note that $\ad{in-U}$ appears in the sort of $\ad{in-U∼}$. Since these types are
+mutually defined, this is an instance of \emph{induction-induction}, a schema
+that allows the definition of a type mutually with other types that contain the
+first one in their signature [cit]. The main example of induction-induction can
+be found in intrinsic definitions of type theory in type theory [cite tt in tt].
 
@@
 induction-induction. The ability to define arbitrary, infinitary
-inductive-inductive types clashes with our objective of keeping the metatheory
-as small and simple as possible. The next step is therefore to reduce this
-inductive-inductive universe to one that does not require (infinitary)
+inductive-inductive types clashes, again, with our objective of keeping the
+metatheory as small and simple as possible. The next step is therefore to reduce
+this inductive-inductive universe to one that does not require (infinitary)
 induction-induction.

### 2020-06-25 paper.tex
More paper writing

@@
 
-In Section~\label{sec:metatheory}...
+We begin by describing the metatheory that we will use throughout the paper, in
+Section~\ref{sec:metatheory}. In Section~\ref{sec:setoid-model}, after briefly
+recalling the abstract notion of models of Type Theory known as categories with
+familes (CwF), we outline Altenkirch's setoid model of Type Theory in the
+variant used in \cite{mpc19}. We then show how the model has been turned into
+Setoid Type Theory in Section~\ref{sett}.
+
+In Section~\ref{setoid-universe} we provide the actual construction of the
+setoid universe in the setoid model. In Section~\ref{ir-universes} we first
+recall the well known inductive-recursive definition of universes in Type
+Theory, and the way these can be turned into a construction that does not
+require induction-recursion.
+%
+We then provide, in Section~\ref{ir-setoid-universe}, a first complete
+definition of the setoid universe using a quite complex form of
+induction-recursion. This is followed by an alternative definition of the
+universe in Section~\ref{ii-universe}, that does not rely on induction-recursion
+but instead on infinitary induction-induction.
+%
+Further reductions of this inductive-inductive universe to even simpler forms,
+in addition to other long term goals, are discussed in
+Section~\ref{further-work}.
 
@@
 metatheory that we pay attention to is it's minimality, as we want to keep it as
-simple as possible. We want to particularly avoid having to assume strong
-definition schemas that go beyond inductive families (or, semantically, what can
-be modeled with W types.)
+simple as possible. In particular, we want to avoid having to assume strong
+definition schemas that go beyond inductive families.
 
@@
 
-\section{Setoid model}
+\section{Setoid model}\label{setoid-model}
 
@@
 
-By \emph{setoid model} we intend a class of models of (intensional) type theory
-where contexts/closed types are interpreted as setoids, i.e. sets with an
-equivalence relation, and dependent types are interpreted as dependent/indexed
-setoids.
+By \emph{setoid model} we intend a class of models of Type Theory where
+contexts/closed types are interpreted as setoids, i.e. sets with an equivalence
+relation, and dependent types are interpreted as dependent/indexed setoids.
 %
-A setoid model was first given by M. Hofmann in \cite{hofmann-ext-int} in order
-to give a semantics to extensionality principles such as functional and
-propositional extensionality. Here we consider a similar model construction, due
-to T. Altenkirch \cite{setoid-model}.
+A setoid model was first given for intensional Type Theory by M. Hofmann in
+\cite{hofmann-ext-int}, in order to give a semantics to extensionality
+principles such as function and propositional extensionality. Here we consider
+a similar model construction due to Altenkirch \cite{setoid-model}.
 %
 The peculiarity of this model is that it is presented in a type theoretic and
-intensional metatheory, in our case $MLTT^{Prop}$.
+intensional metatheory which includes a strict universe of propositions.
 %
@@
 this model \emph{strict}. A strict model is one where every equation holds
-definitionally. The setoid model thus defined validates extensionality
-principles like funext, proext, and quotient types. As a consequence, it
-provides a way to bootstrap and explain extensionality by implementing it in a
-core intensional type theory.
+definitionally. The setoid model thus defined validates function extensionality,
+a univalent universe of propositions, and quotient types. Therefore, it provides
+a way to bootstrap and ``explain'' extensionality, since the model construction
+effectively gives an implementation of various extensionality principles it in
+terms of a small, entirely intensional type theory.
+
+\subsection{Setoid model as a CwF}
 
@@
 category with families (CwF) with extra structure for the various type and term
-formers. The core structure can be defined as the following signature:
+formers. The core structure of a CwF can be given as the following signature:
 
@@
 
-Contexts are given by setoids, that is types together with an equivalence relation.
+In our presentation of the setoid model, contexts are given by setoids, that is
+types together with an equivalence relation.
 
@@
     & \GG^\sim : | \GG | \to | \GG | \to \mProp \\
-    & ...
+    & \textsf{refl}\ \GG : (\gamma : | \GG |) \to \GG^\sim\,\gamma\,\gamma \\
+    & \textsf{sym} \ \GG : \forall\{\gamma_0\ \gamma_1\} \to \GG^\sim\,\gamma_0\,\gamma_1
+    \to \GG^\sim\,\gamma_1\,\gamma_0 \\
+    & \textsf{trans} \ \GG : \forall\{\gamma_0\ \gamma_1\ \gamma_2\}
+    \to \GG^\sim\,\gamma_0\,\gamma_1
+    \to \GG^\sim\,\gamma_1\,\gamma_2
+    \to \GG^\sim\,\gamma_0\,\gamma_2
   \end{aligned}
@@
 Types in a context $\GG$ are given by setoid morphisms from $\GG$ to the
-(strict) category of setoids. The definition can be unfolded as follows:
+(strict) category of setoids. The definition can be unfolded as follows.
 
@@
     \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
-    & \textsf{coe} : \{\gamma_0\ \gamma_1 : | \GG |\} \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
-    & \dots
+    & ... \\
+    & \textsf{coe} : \forall\{\gamma_0\ \gamma_1 : | \GG |\} \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
+    & \textsf{coh} : \forall\{\gamma_0\ \gamma_1 : | \GG |\} \ (p : \GG^\sim\ \gamma_0\ \gamma_1) (a : | A | \gamma_0)
+    \to A^\sim\,p\,a\,(\textsf{coe}\,A\,p\,a) \\
   \end{aligned}
@@
 
-This definition of types in the setoid model is different from the original one
-in [cite LICS 99 paper], but it is equivalent to it [cite].
-%
+This definition of types in the setoid model is different from the one in [cite
+  LICS 99 paper], but it is equivalent to it [cite]. The main difference here is
+in the use of heterogeneous equality in the definition of types.
+
 Substitutions are interpreted as functors between the corresponding setoids,
 whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
-fibration $\GG . A \to \GG$. Note that we only need to include components for
-the functorial action on objects and morphisms, since the functor laws follow
-from proof-irrelevance.
+setoid fibration $\GG . A \to \GG$. Note that we only need to include components
+for the functorial action on objects and morphisms, since the functor laws
+follow from proof-irrelevance in the metatheory, and thus hold definitionally.
 
@@
 
-\subsection{Setoid Type Theory}
+\subsection{Setoid Type Theory}\label{sett}
 
@@
 interpreted as target contexts together with a target type representing the
-equivalence relation, and so on.
+equivalence relation, and so on. \todo{maybe expand this?}
 
@@
 
-\section{Universe of setoids}
+\section{Universe of setoids}\label{setoid-universe}
 
@@
 
-\subsection{Encoding of universes via IR and II}
+\subsection{Inductive-recursive universes in Type Theory}\label{ir-universes}
 
@@
 
-\subsection{A first definition of the setoid universe}
+\subsection{A first definition of the setoid universe}\label{ir-setoid-universe}
 
@@
 
-\subsection{Inductive-inductive universe}
+\subsection{Inductive-inductive universe}\label{ii-universe}
 
@@
 
-\section{Indexed inductive universe (maybe?)}
+\section{Inductive universe (maybe?)}
 
@@
 
-\section{Related work}
+\section{Further work}\label{further-work}
 
@@
 
-\section{Further work}
+\section{Related work}
 

### 2020-06-29 paper.tex
More paper writing

@@
 
+\subsection{Related work}
+
+TODO
+
 \section{$MLTT^{Prop}$}\label{metatheory}
@@
 This section describes $MLTT^{Prop}$, the metatheory in which the model
-construction described in the following sections takes place. One aspect of the
-metatheory that we pay attention to is it's minimality, as we want to keep it as
-simple as possible. In particular, we want to avoid having to assume strong
-definition schemas that go beyond inductive families.
+construction described in the following sections takes place. We employ Agda
+notation for metatheoretic terms throughout the paper.
+
+One aspect of the metatheory that we pay attention to is it's minimality, as we
+want to keep it as simple as possible. In particular, we want to avoid having to
+assume strong definition schemas that go beyond inductive families.
 
@@
 rules for $\Pi, \Sigma$, and $\mathbf{1}$ types. Propositions can be lifted to
-types via $\textsf{Lift} : \mProp \to \mType$. The universe of propositions is
-inhabited by false $\bot$ and propositional truncation $\|\_\| : \mType \to
-\mProp$. We also require indexed W-types, both in $\mType$ and $\mProp$:
+types via $\textsf{Lift} : \mProp \to \mType$.
+
+%% The universe of propositions is
+%% inhabited by false $\bot$ and propositional truncation $\|\_\| : \mType \to
+%% \mProp$.
+
+We also require indexed W-types, both in $\mType$ and $\mProp$:
 
@@
 
-In addition to the proof-relevant types just listed, we will need the
+From $W_p$ we can define propositional truncation:
+
+\begin{align*}
+  & \|\_\| : \mType \to \mProp \\
+  & |\_| : \{A : \mType \} \to A \to \| A \| \\
+  & \textsf{elim}_{\|\_\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P
+\end{align*}
+
+In addition to the above type formers in $\mType$, we will need the
 propositional versions of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can
@@
 
-Note that a universe of strict propositions has been recently added to the Agda
-proof assistant [cite], thus making $MLTT^{Prop}$ a subset of Agda. In fact,
-most of the developments presented here have been formalized and proof-checked
-using Agda extended with strict propositions. For convenience we will employ
-Agda notation for metatheoretic constructions throughout the paper, with the
-implication that these can be understood as, or easily translated to, equivalent
-$MLTT^{Prop}$ definitions.
+Finally, we will assume the presence of an identity type in the style of
+Martin-L\"of's inductive identity type. The main difference is that our identity
+type is a $\mProp$-valued relation. Naturally, we also have a \textsf{transp}
+combinator representing the corresponding elimination principle.
+
+\begin{align*}
+  & \textsf{Id} : \{A : \mType \} \to A \to A \to \mProp \\
+  & \textsf{refl} : \{A : \mType \} (a : A) \to \textsf{Id}\ a\ a \\
+  & \textsf{transp} : \{A : \mType \} (C : A \to \mType) \{a_0\ a_1 : A\} \to \textsf{Id}\ a_0\ a_1 \to C\ a_0 \to C\ a_1
+\end{align*}
+
+Note that this
+represents a strong elimination principle, as it allows to eliminate a strict
+proposition into arbitrary types. Indeed, the reason why we must assume this
+identity type as primitive is because, even though we could have defined
+\textsf{Id} from $W_p$, \textsf{transp} would not be definable from $W_p$
+elimination. \todo{what is the justification for this?}
+
+\subsection{Formalization}
+
+A universe of strict propositions has been recently added to the Agda proof
+assistant [cite], thus making $MLTT^{Prop}$ a subset of Agda. In fact, most of
+the developments presented here have been formalized and proof-checked using
+Agda extended with strict propositions.
+%
+For convenience, we also slightly deviate from $MLTT^{Prop}$ both in the paper
+and in the formalization, for instance by relying on pattern matching instead of
+eliminators, and using primitive versions of $\mProp$-valued $\Pi$ and $\Sigma$
+instead of their derived forms. We operate under the assumption that everything
+can be equivalently carried out in $MLTT^{Prop}$, although we haven't fully
+checked all the necessary details.
 
@@
 pointed out, this definition uses induction-recursion. Moreover, the kind of
-recursion involved in this definition is particularly complex, since it involves
-several recursive definitions whose types contain mutually defined functions. It
-is not clear whether this definition schema, which we could call
-induction-recursion-recursion, can be reduced to simpler, better understood
+recursion involved in this definition is particularly complex, since TODO. It is
+not clear whether this definition can be reduced to simpler, better understood
 notions of induction-recursion like in [dybjer-setzer]. Nevertheless, even the
@@
 
-It is known that finitary inductive-inductive definitions can be translated into
-inductive families \cite{iit-erasure,iit-to-ix,induction-is-enough} but it is
-not clear whether this extends to an infinitary type like our universe.
+\section{Inductive universe (maybe?)}
+
+\subsection{Reducing finitary induction-induction}
+
+It is known that finitary inductive-inductive definitions can be reduced to
+inductive families \cite{iit-erasure,iit-to-ix,induction-is-enough}.
+%
+To illustrate the idea, let us consider a well-known example of finitary
+inductive-inductive type, the intrinsic encoding of type theory in type theory
+itself. Actually, we only need to consider the type of contexts $\textsf{Con} :
+\mType$ and the type of types $\textsf{Ty} : \textsf{Con} \to \mType$; since the
+latter is indexed over the former, this is already an example of
+induction-induction.
+
+Contexts in \textsf{Con} are formed out of empty contexts $\bullet$ and context
+extension $\_,\_$. Types in \textsf{Ty} are either the base type $\iota$ or
+$\Pi$ types.
+
+\begin{align*}
+  & \bullet : \textsf{Con} \\
+  & \_,\_ : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \\
+  & \iota : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
+  & \Pi : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
+\end{align*}
+
+The general method to eliminate induction-induction is to split the original
+inductive-inductive types into a type of codes and an associated well-formedness
+predicate. In our \textsf{Con/Ty} example, these would be respectively given by
+codes $\textsf{Con}_0, \textsf{Ty}_0 : \mType$ and predicates $\textsf{Con}_1 :
+\textsf{Con}_0 \to \mType, \textsf{Ty}_1 : \textsf{Con}_0 \to \textsf{Ty}_0 \to
+\mType$.
+
+The definition of the codes and predicate types follows that of the original
+inductive-inductive type, and can be derived systematically from it. More
+importantly, they can be defined without induction-induction, since although
+$\textsf{Con}_0$ and $\textsf{Ty}_0$ are defined mutually, their sorts are not
+indexed.
+
+\begin{align*}
+  & \bullet_0 : \textsf{Con}_0 \\
+  & \_,_0\_ : (\GG : \textsf{Con}_0) \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
+  & \iota_0 : (\GG : \textsf{Con}_0) \to \textsf{Ty}_0 \\
+  & \Pi_0 : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \\
+  & \bullet_1 : \textsf{Con}_1\ \bullet_0 \\
+  & \_,_1\_ : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
+  & \iota_1 : \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
+  & \Pi_1 : \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
+\end{align*}
+
+We can recover the original inductive-inductive type as $\textsf{Con} :\equiv
+\Sigma (\GG_0 : \textsf{Con}_0)(\textsf{Con}_1\ \GG_0)$ and $\textsf{Ty}\ \GG
+:\equiv \Sigma(A_0 :
+\textsf{Ty}_0)(\textsf{Ty}_1\ (\pi_1\ \GG)\ A_0)$. Recovering the constructors
+is straighforward:
+
+\begin{align*}
+  & \bullet :\equiv (\bullet_0 , \bullet_1) \\
+  & (\GG_0 , \GG_1) , (A_0 , A_1) :\equiv ((\GG_0 ,_0 A_0) , (\GG_1 ,_1 A_1)) \\
+  & \iota\ (\GG_0 , \GG_1) :\equiv (\iota_0 \GG_0 , \iota_1\ \GG_1) \\
+  & \Pi\ \{\GG_0 , \GG_1\}(A_0 , A_1) (B_0 , B_1) :\equiv
+    ((\Pi_0\ \GG_0\ A_0\ B_0) , (\Pi_1\ \GG_1\ A_1\ B_1))
+\end{align*}
+
+Finally, we can define eliminators/induction principles for $\textsf{Con}$ and
+$\textsf{Ty}$ as just defined, by induction on the well-typing predicates.
+
+For any inductive-inductive type $A : \mType, B: A \to \mType$, the most general
+form of (dependent) eliminator is given by motives $F : A \to \mType$ and $G :
+(a : A) \to (b : B a) \to F\ a \to \mType$, and the functions $\textsf{elim}_A :
+(a : A) \to F\ a, \textsf{elim}_B : (a : A) \to (b : B\ a) \to
+G\ a\ b\ (\textsf{elim}_A\ a)$. A simpler notion of eliminator does not require
+the second component of the motive to mention the first. In particular, we have
+$F : A \to \mType$, $G : (a : A) \to B\ a \to \mType$, and eliminators
+$\textsf{elim}_A : (a : A) \to F\ a, \textsf{elim}_B : (a : A) (b : B\ a) \to
+G\ a\ b$. Note that this elimination principle---that we call \emph{simple
+  elimination} after [cite]---is no longer recursive-recursive, since
+$\textsf{elim}_A$ is not mentioned in the signature of $\textsf{elim}_B$.
+
+\subsection{Reducing the setoid universe}
+
+The reduction described in the previous section works generically for an
+arbitrary finitary inductive-inductive type, thus giving a systematic way to
+reduce finitary inductive-inductive definitions to inductive families. However,
+it is not clear whether this method extends to infinitary induction-induction,
+of which the setoid universe defined in Section~\ref{?} is an instance.
 %
 Of course, the absence of a general reduction method does not mean that
-particular concrete instances of induction-induction cannot be reduced to
-simpler principles, which is exactly what we hope for our universe construction.
-
-\section{Inductive universe (maybe?)}
-
-TODO: depends on whether we manage to finish the formalization.
+particular concrete instances of infinitary induction-induction cannot be
+reduced to simpler principles, which is exactly what we hope for our universe
+construction.
+
+The obvious challenge in successfully completing this reduction is to avoid the
+need for extensionality in the metatheory. In fact, consider the simple
+infinitary inductive-inductive type obtained from the previous $\textsf{Con/Ty}$
+example by replacing the finitary constructor $\Pi$ with an infinitary one: $\Pi
+: \{\GG : \textsf{Con}\} \to (\mathds{N} \to \textsf{Ty}\ \GG) \to
+\textsf{Ty}\ \GG$. Already with this simple example, we run into problems as
+soon as we try to define the general eliminator, since it requires function
+extensionality. \todo{Is this true?}
+
+Luckily, in our case we can take advantage of our specific circumstances. First
+of all, we can define the well-typing predicates as $\mProp$-valued families,
+rather than in $\mType$; without this, we would need function extensionality to
+prove that the predicates are propositions (with respect to propositional
+equality).
+
+\begin{align*}
+  & \textsf{data}\ \ad{in-U}_0 : \mType \to \mType_1 \\
+  & \textsf{data}\ \ad{in-U∼}_0 : \{A\ A' : \mType\} \to (A \to A' \to \mProp) \to \mType_1 \\
+  & \textsf{data}\ \ad{in-U}_1 : (A : \mType) \to \ad{in-U}_0\ A \to \mProp_1 \\
+  & \textsf{data}\ \ad{in-U∼}_1 : \{A\ A' : \mType\} \to (R : A \to A' \to \mProp) \to \ad{in-U∼}_0\ R \to \mType_1
+\end{align*}
+\todo{check that these signatures are correct}
+
+Using $\mProp$ would help only to some extent if we had to define general
+recursive-recursive eliminators, since that would also require
+extensionality. However, it turns out that for our limited use of this
+inductive-inductive universe the \emph{simple} elimination principle is
+sufficient. This eliminator can be defined by simple pattern matching, and does
+not require extensionality.
 
 \section{Further work}\label{further-work}
-
-TODO
-
-\section{Related work}
 

### 2020-06-29 paper.tex
Added footnote on functorial model conostruction PoV.

@@
 interpreted as target contexts together with a target type representing the
-equivalence relation, and so on. \todo{maybe expand this?}
+equivalence relation, and so on.\todo{maybe expand this?}%
+\footnote{%
+Semantically, this translation corresponds to a model construction, in particular a functor from the category of models of the target theory to the category of models of what will be Setoid Type Theory.
+Since the setoid translation is structural in the context component, we can work with models in the style of categories with families rather than contextual categories.
+}
 

### 2020-06-29 paper.tex
Remark on separate sorts for propositions and sets.

@@
 
+Yet another choice is whether to have two separate sorts, one for propositions
+and one for sets (with propositions convertible to sets) or a single sort of
+types (sets), with propositions given by elements of a universe of proposition,
+which is a (large) type. We have chosen to present the second option to fit with
+the standard notion of (unisorted) CwF. However, this has downsides: to even
+talk about propositions, we need to have a notion of large types. The first
+option is more symmetric: we can have parallel hierarchies for propositions and
+sets.
+
 \subsection{Inductive-recursive universes in Type Theory}\label{ir-universes}

### 2020-06-29 paper.tex
Further work: comparison with XTT.

@@
 
+One question is how models of SeTT and XTT \todo{reference} are related (assuming both sides are restricted to comparable collections of type formers).
+From a model of XTT, do we functorially obtain a model of SeTT?
+What about the reverse direction, assuming we have type case in SeTT? \todo{insert reference to where we define/discuss this in this paper}
+
 TODO

### 2020-06-30 paper.tex
Changed institution.
More paper writing

--- 2020-06-29 paper.tex
@@
 \and
-  Christian Sattler\inst{1}\thanks{Supported by USAF grant FA9550-16-1-0029.}
+  Christian Sattler\inst{3}\thanks{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.}
 \and
@@
   \email{akaposi@inf.elte.hu}
+  \and
+  Chalmers University of Technology, Gothenburg, Sweden
+  \email{sattler@chalmers.se}
 }
@@
 
-A central aspect of Type Theory is the coexistence of two notions of
+Intuitionistic type theory is a formal system designed by Per Martin-L\"of to be
+a full-fledged foundation in which to develop constructive mathematics
+\cite{mltt75, libretto}.
+%
+A central aspect of type theory is the coexistence of two notions of
 equality. On the one hand definitional equality, the computational equality that
@@
 the center of type theory research; however, it is generally understood that to
-properly support formalization of mathematics in type theory, one should aim for
-a notion of definitional equality that is decidable and therefore
-machine-checkable, and a notion of propositional equality that is as as
-extensional possible.
-
-Two extensionality principles seem particularly desirable, in that they arguably
+properly support formalization of mathematics, one should aim for a notion of
+propositional equality that is as \emph{extensional} as possible.
+
+Two extensionality principles seem particularly desirable, since they arguably
 constitute the bare minimum for type theory to be comparable to set theory as a
 foundational system for set-level mathematics, in terms of power and ergonomics.
-One is function extensionality (funext), according to which functions are equal
-if point-wise equal. Another is propositional extensionality (propext), equating
-all propositions that are logically equivalent.
-
-Extensional Type Theory (ETT) does support extensional reasoning to some degree,
-and thus would seem a convenient setting for formalization of
-mathematics. However, ETT's equality reflection makes type-checking ETT terms
+One is function extensionality (or \emph{funext}), according to which functions
+are equal if point-wise equal. Another is propositional extensionality
+(or \emph{propext}), that equates all propositions that are logically equivalent.
+
+Type theory with equality reflection, also known as \emph{extensional type
+  theory} (ETT) does support extensional reasoning to some degree, but
+unfortunately equality reflection makes the problem of type-checking ETT terms
 computationally unfeasible, if not impossible.
 
-On the other hand, Intensional Type Theory (ITT) has nice computational
-properties that can make it more suitable for computer implementation, but as
-normally defined [cite Martin-L\"of paper?] it lacks extensionality.
+On the other hand, \emph{intensional type theory} (ITT) has nice computational
+properties like decidable type checking that can make it more suitable for
+computer implementation, but as normally defined (for example, in \cite{mltt75})
+it severely lacks extensionality.
 %
@@
 %
-Moreover, ITT extended with uniqueness of identity proofs (UIP) and funext is
-known to be as powerful as ETT [cite hofmann]. These principles can be added as
-axioms, however destroying some computational properties of ITT such as
-canonicity.
-
-What we would like instead is a formulation of ITT that supports the
-aforementioned extensionality principles, while retaining convenient
-computational properties like canonicity and decidability of type checking.
-%
-Unfortunately, one can show that whenever identity types are defined according
-to Martin-Loef's inductive definition, there is no hope to have canonicity and
-funext coexist.
+Moreover, ITT extended with the principle of \emph{uniqueness of identity
+  proofs} (UIP) and funext is known to be as powerful as ETT
+\cite{conservativity}. We could recover the expressive power of ETT by adding
+these principles to ITT as axioms, however destroying some computational
+properties like canonicity.
+
+What we would like instead is a formulation of ITT that supports extensionality,
+while retaining its convenient computational behaviour.
+%
+Unfortunately, one can show that canonicity and function extensionality cannot
+coexist if propositional equality is given by Martin-L\"of's inductively-defined
+identity types.
 %
 The first step towards a solution is to give up the idea of propositional
-equality as single type family inductively generated by reflexivity. Instead,
-equality should be \emph{specific} to each type former in the type theory, or in
-other words, every type former should be introduced alongside an explanation of
-what counts as equality for its elements. In this setting, function
-extensionality can simply be introduced as the chosen notion of equality on
-function types.
-
-This conception of types together with their notion of equality goes back to the
-notion of \emph{setoids} or \emph{Bishop sets}. A setoid is simply a set
-equipped with a propositional equivalence relation. Setoids provide a quite
+equality as a single inductive definition given homogeneously for an arbitrary
+type. Instead, equality should be \emph{specific} to each type former in the
+type theory, or in other words, every type former should be introduced alongside
+an explanation of what counts as equality for its elements.
+%% In this setting, function extensionality can simply be introduced as the
+%% chosen notion of equality on function types.
+
+This idea of pairing types together with their own equality relation goes back
+to the notion of \emph{setoid} or \emph{Bishop set}. Setoids provide a quite
 natural and useful semantic domain in which to interpret type theory. The first
-setoid model was constructed to show consistency of funext for ITT [cite]. This
-consistency result is implied by the observation that UIP and funext hold in the
-setoid model. Moreover, it was shown by Altenkirch [cite] that is the model
-construction is carried out in a metatheory with a universe of strict
-(definitionally proof-irrelevant) propositions, it is possible to define a
-univalent universe of propositions, therefore satisfying propositional
-extensionality. The setoid model thus satisfies all the extensionality
-principles that we would like to extend ITT with. The question is whether there
-exists a version of ITT that supports setoid reasoning, and hence the forms of
-extensionality enables by it.
-
-This question was revisited and answered in Altenkirch et al. [cite]. In this
-paper, the authors define SeTT an extension of intensional MLTT + Prop with
-constructs for setoid reasoning, where funext and propext hold by
-definition. Moreover, they justify SeTT via a syntactic translation, after
-observing that this is made possible by the strictness of the model (every
-equation holds definitionally). This is in contrast with other type theories
-based on the setoid model, like OTT [cite] and XTT [cite], which instead rely on
-ETT as metatheory for their justification. A major contribution of SeTT is
-therefore to provide a way to bootstrap extensionality, by translating it into a
-small core intensional theory by means that do not require extensionality to
-begin with.
-
-The current formulation of SeTT is already a rich theory. However, its
-introspection capabilities are currently lacking, as they are limited to the
-univalent universe of propositions. We would like to internalise the notion of
-type in SeTT, thus extending the theory with a universe of setoids.
-%
-This brings up several questions, one of which has to do with the notion of
-equality with which the universe should come equipped. The universe of setoids
-is itself a setoid (as any type is), so it certainly cannot be a univalent
-universe, since a univalent setoid would be a groupoid rather than a setoid.
-%
-Another question asks how one can justify such universe in the setoid model. In
-particular what principles, are needed from the metatheory in order to define
-the universe.
+setoid model was constructed to show consistency of funext for ITT
+\cite{hofmann}. This consistency result is implied by the observation that UIP
+and funext hold in the setoid model. Moreover, it was shown by Altenkirch
+\cite{setoid99} that is the model construction is carried out in a type
+theoretic metatheory with a universe of strict (definitionally proof-irrelevant)
+propositions, it is possible to define a univalent universe of propositions
+satisfying propositional extensionality. The setoid model thus satisfies all the
+extensionality principles that we would like to have in type theory. The
+question is whether there exists a version of intensional type theory that
+supports setoid reasoning, and hence the forms of extensionality enabled by it.
+
+This question was revisited and answered in Altenkirch et al. \cite{mpc19}. In
+this paper, the authors define Setoid Type Theory (SeTT), an extension of
+intensional Martin-L\"of type theory with constructs for setoid reasoning, where
+funext and propext hold by definition. SeTT is based on the \emph{strict} setoid
+model of Altenkirch\footnote{a \emph{strict} model is one where every equation
+  holds definitionally}, which makes it possible to show consistency via a
+syntactic translation. This is in contrast with other type theories based on the
+setoid model, like Observational Type Theory \cite{alti:ott-conf, alti:ott-conf}
+and XTT \cite{xtt}, which instead rely on ETT for their justification. A major
+property of SeTT is thus to illustrate how to bootstrap extensionality, by
+translation into a small intensional core.
+
+SeTT as defined in \cite{mpc19} is already a rich theory, but its introspection
+capabilities are currently lacking, as its universes are limited to
+propositions. We would like to internalise the notion of type in SeTT, thus
+extending the theory with a universe of setoids.
+%
+This goal brings up several questions, one of which has to do with the notion of
+equality with which the universe should come equipped: the universe of setoids
+is itself a setoid (as any type is) so it certainly cannot be univalent, since
+setoids lack the necessary structure.
+%
+Another issue is the way such universe can be justified by the setoid model, and
+in particular what principles are needed in the metatheory to do so.
 
@@
 to the design and implementation of this construction. Our main contribution is
-the definition of the universe in the setoid model; this is initially given as
-an inductive-inductive-recursive definition, which is then translated firstly to
-an inductive-inductive definition, and secondly to a purely inductive type. As a
-consequence, we show that we only need to assume indexed W-types in the
-metatheory (along with some obligatory basic tools like $\Sigma$ and $\Pi$
-types) to justify the universe.
+the construction of the universe in the model; this is given in steps, first as
+an inductive-recursive definition, which is then translated to an
+inductive-inductive definition, and subsequently to an inductive type. As a
+consequence, we show that we only need to assume indexed W-types and
+proof-irrelevant identity types in the metatheory (along with some obligatory
+basic tools like $\Sigma$ and $\Pi$ types) to construct the universe.
+
+The universe constructions presented in this paper are, to our knowledge, the
+first examples of two kinds of data type reductions in an intensional
+metatheory: the first involving an inductive-recursive type which including
+strict propositions, and the second involving an infinitary inductive-inductive
+type.
 
@@
 Section~\ref{sec:metatheory}. In Section~\ref{sec:setoid-model}, after briefly
-recalling the abstract notion of models of Type Theory known as categories with
-familes (CwF), we outline Altenkirch's setoid model of Type Theory in the
-variant used in \cite{mpc19}. We then show how the model has been turned into
-Setoid Type Theory in Section~\ref{sett}.
-
-In Section~\ref{setoid-universe} we provide the actual construction of the
-setoid universe in the setoid model. In Section~\ref{ir-universes} we first
-recall the well known inductive-recursive definition of universes in Type
-Theory, and the way these can be turned into a construction that does not
-require induction-recursion.
+recalling \emph{categories with families} as an abstract notion of models of
+type theory, we outline Altenkirch's setoid model as given in \cite{mpc19}. We
+then briefly discuss the rules of Setoid Type Theory in Section~\ref{sett}.
+
+In Section~\ref{setoid-universe} we discuss the setoid model and various design
+choices related to it. We then recall inductive-recursive universes, and the way
+they can be equivalently defined as a plain inductive definition, in
+Section~\ref{ir-universes}.
 %
 We then provide, in Section~\ref{ir-setoid-universe}, a first complete
-definition of the setoid universe using a quite complex form of
-induction-recursion. This is followed by an alternative definition of the
-universe in Section~\ref{ii-universe}, that does not rely on induction-recursion
-but instead on infinitary induction-induction.
-%
-Further reductions of this inductive-inductive universe to even simpler forms,
-in addition to other long term goals, are discussed in
+definition of the setoid universe using a form of induction-recursion. This is
+followed by an alternative definition of the universe in
+Section~\ref{ii-universe}, that does not rely on induction-recursion but instead
+on infinitary induction-induction. This inductive-inductive encoding of the
+universe is obtained from the inductive-recursive one, following the method of
+Section~\ref{ir-universes}.
+%
+We end the series of universe constructions with
+Section~\ref{inductive-universe}, where we outline a purely inductive definition
+of the setoid universe, obtained from the inductive-inductive one.
+
+We conclude the paper with a discussion of further work in
 Section~\ref{further-work}.
@@
 
-TODO
+The setoid model was first described in \cite{hofmann} in order to add
+extensionality principles to Type Theory such as function extensionality and
+propositional extensionality. A strict variant of the setoid model was given in
+\cite{setoid99} using a definitionally proof-irrelevant universe of
+propositions. Recently, support for such a universe was added to the
+proof-assistants Agda and Coq \cite{agda-prop}, allowing a full formalization of
+Altenkirch's setoid model. Setoid Type Theory (SeTT) is a recently developed
+formal system derived from this model construction \cite{mpc19}. Observational
+Type Theory (OTT) \cite{alti:ott-conf, alti:ott-conf} is a syntax for the setoid
+model differing from SeTT for the use of a different notion of heterogeneous
+equality. Moreover, the consistency proof for OTT relies on Extensional Type
+Theory, whereas for SeTT it is obtained via a syntactic translation.
+%
+XTT \cite{xtt} is a cubical variant of OTT where the equality type is defined
+using an interval pretype. XTT's universes support universe induction, whereas
+it is left open whether the construction presented here supports this principle.
+
+The principle of propositional extensionality in the setoid model is an instance
+of Voevodsky's univalence axiom \cite{hottbook}. The cubical set model is a
+constructive model justifying this axiom \cite{cubical-sets}. A type theory
+extracted from this model is Cubical Type Theory \cite{cubical}. The
+relationship between the cubical set model and cubical type theory is similar to
+that between the setoid model and SeTT.
+%
+Compared to cubical type theories, SeTT has the advantage that the equality type
+satisfies more definitional equalities. For instance, whereas in cubical type
+theory equality of functions is isomorphic to pair-wise equality, in SeTT the
+isomorphism is replaced by a definitional equality. SeTT is also a syntactically
+straightforward extension of Martin-L\"of Type Theory, that does not require
+exotic objects like the internal pretype. In turn, the obvious advantage of
+cubical type theory is that it is not limited to setoids.
+
+An exceptional aspect of the metatheory used in this paper is the presence of a
+proof-irrelevant identity type with a strong transport rule allowing to
+eliminate into arbitrary types.
+%
+In \cite{abel:nbe09}, Abel given a proof of normalization by evaluation for the
+Logical Framework extended with a similar proof-irrelevant equality type.
+%
+Abel and Coquand show in \cite{abel2019failure} that the combination of
+impredicativity with a strong transport rule results in terms that fail to
+normalize.
 
@@
 
-Note that this
-represents a strong elimination principle, as it allows to eliminate a strict
-proposition into arbitrary types. Indeed, the reason why we must assume this
-identity type as primitive is because, even though we could have defined
-\textsf{Id} from $W_p$, \textsf{transp} would not be definable from $W_p$
-elimination. \todo{what is the justification for this?}
+Note that \textsf{transp} provides a strong elimination principle allowing to
+eliminate a strict proposition (the identity type) into arbitrary types. Indeed,
+the reason why we must assume this identity type as primitive is because, even
+though we could have defined \textsf{Id} from $W_p$, \textsf{transp} would not
+be definable from $W_p$ elimination. \todo{what is the justification for this?}
 
@@
 A universe of strict propositions has been recently added to the Agda proof
-assistant [cite], thus making $MLTT^{Prop}$ a subset of Agda. In fact, most of
-the developments presented here have been formalized and proof-checked using
-Agda extended with strict propositions.
-%
-For convenience, we also slightly deviate from $MLTT^{Prop}$ both in the paper
-and in the formalization, for instance by relying on pattern matching instead of
+assistant \cite{agda-prop}, thus making $MLTT^{Prop}$ a subset of Agda. In fact,
+most of the developments presented here have been formalized and proof-checked
+using Agda extended with strict propositions.
+%
+For convenience, we slightly deviate from $MLTT^{Prop}$ both in the paper and in
+the formalization, for instance by relying on pattern matching instead of
 eliminators, and using primitive versions of $\mProp$-valued $\Pi$ and $\Sigma$
-instead of their derived forms. We operate under the assumption that everything
-can be equivalently carried out in $MLTT^{Prop}$, although we haven't fully
-checked all the necessary details.
+types instead of deriving them from truncation. We operate under the assumption
+that everything can be equivalently carried out in $MLTT^{Prop}$, although we
+haven't fully checked all the necessary details.
 
 \section{Setoid model}\label{setoid-model}
-
-%% The setoid interpretation gives rise to a model of a type theory with functional
-%% and propositional extensionality. It is thus a way to explain extensionality in
-%% a type-theoretic and computationally adequate way \cite{setoid-model}.
 
@@
 A setoid model was first given for intensional Type Theory by M. Hofmann in
-\cite{hofmann-ext-int}, in order to give a semantics to extensionality
+\cite{hofmann}, in order to give a semantics to extensionality
 principles such as function and propositional extensionality. Here we consider
-a similar model construction due to Altenkirch \cite{setoid-model}.
+a similar model construction due to Altenkirch \cite{setoid99}.
 %
@@
 effectively gives an implementation of various extensionality principles it in
-terms of a small, entirely intensional type theory.
+terms of a small, entirely intensional theory.
 
@@
 category with families (CwF) with extra structure for the various type and term
-formers. The core structure of a CwF can be given as the following signature:
+formers \cite{internal-tt}. The core structure of a CwF can be given as the
+following signature:
 
@@
 
-This definition of types in the setoid model is different from the one in [cite
-  LICS 99 paper], but it is equivalent to it [cite]. The main difference here is
-in the use of heterogeneous equality in the definition of types.
+This definition of types in the setoid model is different from the one in
+\cite{setoid99}, but it is equivalent to it \todo{add citation}. The main
+difference here is in the use of heterogeneous equality in the definition of
+types.
 
@@
 formers ($\Pi, \Sigma$, etc.), including function extensionality and a universe
-of strict propositions satisfying propositional extensionality [cite LICS
-  99]. Note that we don't need neither identity types nor inductive types
-(W-types) in the metatheory to achieve this. However, W-types are needed when
-constructing the setoid universe in the next sections.
+of strict propositions satisfying propositional extensionality
+\cite{setoid99}. Note that we don't need neither identity types nor inductive
+types (W-types) in the metatheory to achieve this. However, W-types are needed
+when constructing the setoid universe in the next sections.
 
@@
 interpreted as target contexts together with a target type representing the
-equivalence relation, and so on.\todo{maybe expand this?}%
+equivalence relation, and so on.
 \footnote{%
-Semantically, this translation corresponds to a model construction, in particular a functor from the category of models of the target theory to the category of models of what will be Setoid Type Theory.
-Since the setoid translation is structural in the context component, we can work with models in the style of categories with families rather than contextual categories.
+Semantically, this translation corresponds to a model construction, in
+particular a functor from the category of models of the target theory to the
+category of models of what will be Setoid Type Theory.  Since the setoid
+translation is structural in the context component, we can work with models in
+the style of categories with families rather than contextual categories.
 }
@@
 Note that contractibility of singletons holds in SeTT by UIP and the computation
-rules of equality on $\Sigma$ types. Hence, the J eliminator can also be
-derived.
+rules of equality on $\Sigma$ types. Hence, Martin-L\"of's J eliminator can also
+be derived \todo{add citation}.
 
@@
 
-In its current form, SeTT is already rich and interesting in its own right.
-However, as pointed out in the introduction, it is also seriously limited by the
-lack of a universe internalizing the notion of types beyond propositions, that
-is, a universe of setoids. Our goal is to extend SeTT with such universe; since
-SeTT is a direct syntactic reflection of the setoid model, this essentially
-amounts to showing that a universe of setoids with the necessary structure and
-equations can be constructed within the setoid model. This opens several
-questions and possible design choices.
+As pointed out in the introduction, SeTT is seriously limited by the lack of a
+universes internalizing the notion of setoid. Our goal is to extend SeTT with a
+universe of setoids; since SeTT is a direct syntactic reflection of the setoid
+model, this essentially amounts to showing that a universe of setoids with the
+necessary structure and equations can be constructed within the setoid
+model. This opens several questions and possible design choices.
 
@@
 universe: as any type in the setoid model, this universe must be a setoid and
-thus come equipped with an equivalence relation. However, unlike the already
-existing universe of propositions, a universe of setoids cannot be univalent,
-since this would force it to be a groupoid. Instead, we can define the relation
-so that it reflects a simpler syntactic equality of codes.
+thus come equipped with an equivalence relation. However, unlike the universe of
+propositions, a universe of setoids cannot be univalent, since this would force
+it to be a groupoid. The obvious choice is therefore to have a non-univalent
+universe, and instead define the universe's relation so that it reflects a
+simple syntactic equality of codes rather than setoid equivalence.
 
@@
 the construction of the universe. In fact, one of the main aspects of the setoid
-model construction recalled in Section TODO and shown originally in [cite LICS
-  99] is that it can be carried out in a very small type theoretic metatheory,
-thus providing a way to reduce extensionality to a small intensional core.
-%
-In extending the setoid model with a universe, we would like this property to
-still hold, hence it is of primary concern to reduce to a minimum the extensions
-required in the metatheory, so to keep it as small as possible.
-
-A known and established method for constructing universes in Type Theory uses
+model construction recalled in Section TODO and shown originally in
+\cite{setoid99} is that it can be carried out in a very small type theoretic
+metatheory, thus providing a way to reduce extensionality to a small intensional
+core.
+%
+We would like to stay faithful to this ideal when constructing this setoid
+universe.
+
+A known and established method for defining universes in Type Theory relies on
 induction-recursion (IR), a definition schema developed by Dybjer
-[cite]. Inductive-recursive definitions can be found throughout the literature
-on type theory, from the already mentioned type theoretic universes, including
-the original formulation a la Tarski by Martin-L\"of [cite], to metamathematical
-tools like computability predicates [cite].
+\cite{ir-dybjer, dybjer-setzer}. Inductive-recursive definitions can be found
+throughout the literature, from the already mentioned type theoretic universes,
+including the original formulation \`{a} la Tarski by Martin-L\"of
+\cite{libretto}, to metamathematical tools like computability predicates.
 
@@
 induction (more specifically, inductive families), including some universe
-definitions. We recall this reduction in Section TODO.
-
-%% However, we would prefer not to be forced to assume inductive-recursive types,
-%% or IR in general, in the basic type theory which is the target of the setoid
-%% translation. This is because the point of the translation is to explain a
-%% complex system in terms of a very simple, well understood core, and we think
-%% introducing IR would go against this goal.
+definitions. We recall this reduction in Section~\ref{ir-universes}.
 
@@
 require careful consideration. For instance, one question is whether the setoid
-universe should support elimination, thus exposing the inductive structure of
-the codes. Universe elimination is known to be inconsistent with univalence. As
-it turns our, univalence is not an issue in our case; nevertheless it is not
-immediately clear if the elimination principle can be justified by the
-semantics.
-
-Another choice is whether to include codes for other universes, besides the
-usual type formers. One could have $\Prop$ be a code in the setoid universe,
-rather than a type. Moreover, the universes could form a hierarchy, where the
-universe at level $i$ is an element of the universe of level $i+1$. Another
-option is whether to make the hierarchy cumulative.
-
-%% In the next section, we recall the inductive-recursive construction of universes
-%% in type theory, as well as lay out the process by which such construction can be
-%% turned to one that only relies on inductive families. In Section TODO, we give a
-%% first inductive-recursive definition of the universe of setoids. After that we
-%% will see if and how such IR definition can be turned into one that does not rely
-%% on IR.
+universe should support universe induction, thus exposing the inductive
+structure of the codes. Such elimination principle is known to be inconsistent
+with univalence, although this is not an issue in our case; nevertheless it's
+not immediately clear if the elimination principle can be justified by the
+semantics, that is, if our encoding of the setoid universe in the model allows
+to define such universe eliminator.
+%
+The question arises because we only define a weak eliminator for our final
+inductive-inductive encoding of the setoid universe, for reasons that are
+explained in Section~\ref{inductive-universe}. Although not currently needed, a
+stronger eliminator might be necessary to justify universe induction. This
+problem should not arise in the other encodings of the setoid universe (as given
+in Section~\ref{ir-setoid-universe} and Section~\ref{ii-universe}).
+
+Another design choice has to do with how the setoid universe relates to the
+other universes. One could have $\Prop$ be a code of the setoid universe, rather
+than a type. Moreover, the setoid universes could form a hierarchy, possibly
+cumulative.
 
@@
 
-%% Note that $\U$ is defined to be in $\mType_1$ rather than $\mType$. This is
-%% because... \todo{TODO add explanation}.
-
-Induction-recursion is arguably the nicest and more natural way to define
-internal universes in Type Theory, however it is not always required.
+Induction-recursion is arguably a nicer and more natural way to define
+internal universes in Type Theory, however it is not always strictly required.
 %
@@
 
-Following this, we seen that we can actually encode the universe above as an
-inductive type $\ad{in-U}$ that \emph{carves out} all types in $\Set$ that are
-in the image of $\El$. In other words, $\ad{in-U}$ is a predicate that holds for
-any type that would have been obtained via $\El$ in the usual
-inductive-recursive definition. As $\El$ is indexed on the type of codes, the
-definition of $\ad{in-U}$ quite expectedly reflecs the inductive structure of
-codes.
+In our case, we can encode $\U$ as an inductive type $\ad{in-U}$ that
+\emph{carves out} all types in $\Set$ that are in the image of $\El$. In other
+words, $\ad{in-U}$ is a predicate that holds for any type that would have been
+obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed on
+the type of codes, the definition of $\ad{in-U}$ quite expectedly reflecs the
+inductive structure of codes.
 %
@@
 Note that this construction gives rise to a universe in $\mType_1$, rather than
-$\mType$, since the definition of $\U$ quantifies over all possible sets. This
-also means that a metatheory with at least one universe is required for this
-construction.
+$\mType$, since the definition of $\U$ quantifies over all possible sets. Hence
+this kind of construction requires a metatheory with at least one universe.
 
@@
 to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
-\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling Section TODO,
-these are essentially record types made of several components. Since $\Uty$ is a
-closed type, it requires the same data of a setoid; in particular, we need a type
-of codes together with an equivalence relation reflecting equality of codes:
+\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
+Section~\ref{setoid-model}, these are essentially record types made of several
+components. Since $\Uty$ is a closed type, it requires the same data of a
+setoid; in particular, we need a type of codes together with an equivalence
+relation reflecting equality of codes:
 %
@@
   & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
-  %% & \El : \setoidU \to \Set \\
-  %% & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
 \end{align*}
@@
 
-To define $\EL$, we need a family of setoids indexed over the universe. In other
-words, a way to assign to each code in the universe a carrier set and an
-equivalence relation.
+$\EL$ is given by a family of setoids indexed over the universe, that is, a way
+to assign to each code in the universe a carrier set and an equivalence
+relation.
 %
@@
 
-%% \todo{QIIT II LICS paper. this part is already in the HIIT paper}
-
 We give an inductive definition of $\setoidU$, mutually with a recursive
-definition of everything else. Below we only show the more interesting definitions of $\setoidU$ and $\El$:
+definition of everything else. Below we only show the more interesting
+definitions of $\setoidU$ and $\El$:
 %
@@
   & \quad \textsf{one} : \setoidU \\
-  %% Πs : (a : U i)(b : El a → U i)
-     %% (refb : {x x' : El a}(x~ : refU a ⊢ x ~El x') → b x ~U b x') → U i
   & \quad \textsf{pi}
@@
 
-This gives a complete definition of a universe of setoids. However, as already
-pointed out, this definition uses induction-recursion. Moreover, the kind of
-recursion involved in this definition is particularly complex, since TODO. It is
-not clear whether this definition can be reduced to simpler, better understood
-notions of induction-recursion like in [dybjer-setzer]. Nevertheless, even the
-need to extend the metatheory with simple forms of induction-recursion goes
-against our objective to keep the metatheory as small and essential as possible.
-
-The goal of the next section is therefore to transform the
-inductive-recursvie-recursive definition ourlined above to one that doesn't use
-induction-recursion.
-%
-This transformation is inspired by the well-known trick to eliminate
-induction-recursion described in Section X, but modified in a novel way to
-account for the presence of $\mProp$-valued types. To our knowledge, this is the
-first time this reduction trick is applied to an inductive-recursive type of
-this kind.
-
-%% The above definition of the setoid universe relies on both induction-recursion
-%% and induction-induction. Although happily accepted by Agda, we don't fully
-%% understand the meaning of this combination, so we would like to find an
-%% equivalent definition that does not rely on it. The goal of the next sections is
-%% therefore to get to such an alternative definition. This is done by first
-%% transforming the inductive-inductive-recursive definition to one that doesn't
-%% use induction-recursion, and secondly by reducing the induction-induction to
-%% inductive families.
-
-%% Our transformation of the inductive-inductive-recursive definition is inspired
-%% by the well-known trick described in Section X, but modified in a novel way to
-%% account for the presence of induction-induction and $\mProp$-valued types. To
-%% our knowledge, this is the first time this reduction trick is applied to an
-%% inductive-recursive type of this complexity.
+This gives a complete definition of a universe of setoids, which is, however,
+inductive-recursive. Moreover, the kind of recursion involved in this definition
+is particularly complex, and not obviously reducible to understood notions of
+induction-recursion like in \cite{dybjer-setzer}. Nevertheless, we would like to
+avoid extending the metatheory with any forms of induction-recursion, as per our
+goal to keep the metatheory as small and essential as possible.
+
+As a consequence, in the next section we transform our current
+inductive-recursive definition to one that doesn't use induction-recursion.
+%
+The way this is done is inspired by the well-known trick to eliminate
+induction-recursion described in Section~\ref{ir-universes}, but modified in a
+novel way to account for the presence of $\mProp$-valued types. To our
+knowledge, this is the first time this reduction trick is applied to an
+inductive-recursive type of this kind.
 
@@
 
-The reduction of the IR universe to induction works, conceptually, as in the
-small example of Section X.
-%
-However, in addition to $\ad{in-U}$ for defining $\U$, we also introduce a
-family $\ad{in-U∼}$ of binary relations between types in the universe, from
-which we then define $\_\sim_{\setoidU}\_$.
+We will follows the same method outlined in Section~\ref{ir-universes}. In
+addition to $\ad{in-U}$ for defining $\U$, we also introduce a family
+$\ad{in-U∼}$ of binary relations between types in the universe, from which we
+then define $\_\sim_{\setoidU}\_$.
 %
@@
 Note that $\ad{in-U}$ appears in the sort of $\ad{in-U∼}$. Since these types are
-mutually defined, this is an instance of \emph{induction-induction}, a schema
+mutually defined, they form an instance of \emph{induction-induction}, a schema
 that allows the definition of a type mutually with other types that contain the
-first one in their signature [cit]. The main example of induction-induction can
-be found in intrinsic definitions of type theory in type theory [cite tt in tt].
-
-As in the universe example in Section TODO, we now define $\setoidU$ as a
-$\Sigma$ type, and $\El$ as the corresponding first projection.
+first one in their signature \cite{ii}. The main example of induction-induction
+can be found in intrinsic definitions of Type Theory in Type Theory
+\cite{tt-in-tt}.
+
+As in the universe example in Section~\ref{ir-universes}, we now define
+$\setoidU$ as a $\Sigma$ type, and $\El$ as the corresponding first projection.
 
@@
 
-\section{Inductive universe (maybe?)}
-
-\subsection{Reducing finitary induction-induction}
+\subsection{Inductive universe}
+
+\subsubsection{Reducing finitary induction-induction}
 
@@
 G\ a\ b$. Note that this elimination principle---that we call \emph{simple
-  elimination} after [cite]---is no longer recursive-recursive, since
+  elimination} after \cite{ii}---is no longer recursive-recursive, since
 $\textsf{elim}_A$ is not mentioned in the signature of $\textsf{elim}_B$.
 
-\subsection{Reducing the setoid universe}
+Let us go back to our \textsf{Con}/\textsf{Ty} example. Given any dependent
+algebra of $\textsf{Con}$ and $\textsf{Ty}$ with motives $C : \textsf{Con} \to
+\mType$ and $T : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to C\ \GG \to
+\mType$, the general eliminators have the following signatures:
+
+\begin{align*}
+  & \textsf{elim}_{\textsf{Con}} : (\GG : \textsf{Con}) \to C\ \GG \\
+  & \textsf{elim}_{\textsf{Ty}} : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG)
+  \to T\ \GG\ A\ (\textsf{elim}_{\textsf{Con}}\ \GG)
+\end{align*}
+
+These can be derived from our encoding of $\textsf{Con}$ and $\textsf{Ty}$ via
+untyped codes and well-typing predicates. The way to do it is to first define
+the graph of the eliminators in the form of inductively-generated relations:
+
+\begin{align*}
+  & \textsf{data}\ \textsf{R-Con} :
+    (\GG : \textsf{Con}) \to C\ \GG \to \mType \\
+  & \textsf{data}\ \textsf{R-Ty} : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG)
+  (\gamma : C\ \GG)
+  \to T\ \GG\ A\ \gamma
+  \to \mType
+\end{align*}
+
+The next step is to prove that these relations are functional, by induction on
+the untyped codes $\textsf{Con}_0$ and $\textsf{Ty}_0$. From this result,
+defining the eliminators is immediate.
+
+\subsubsection{Reducing the setoid universe}\label{inductive-universe}
 
@@
 it is not clear whether this method extends to infinitary induction-induction,
-of which the setoid universe defined in Section~\ref{?} is an instance.
-%
-Of course, the absence of a general reduction method does not mean that
-particular concrete instances of infinitary induction-induction cannot be
-reduced to simpler principles, which is exactly what we hope for our universe
-construction.
+of which the setoid universe defined in Section~\ref{ii-universe} is an
+instance.
+%
+Of course, the absence of a general reduction method does not mean that we can't
+reduce particular concrete instances of infinitary induction-induction, which is
+exactly what we hope for our universe construction.
 
@@
 of all, we can define the well-typing predicates as $\mProp$-valued families,
-rather than in $\mType$; without this, we would need function extensionality to
-prove that the predicates are propositions (with respect to propositional
-equality).
+rather than in $\mType$:
 
@@
 
-Using $\mProp$ would help only to some extent if we had to define general
-recursive-recursive eliminators, since that would also require
-extensionality. However, it turns out that for our limited use of this
-inductive-inductive universe the \emph{simple} elimination principle is
-sufficient. This eliminator can be defined by simple pattern matching, and does
-not require extensionality.
+Defining the well-typing predicates in $\mProp$ does help, since otherwise we
+would need function extensionality to prove that these types are propositions
+(with respect to metatheoretic propositional equality).
+%
+However, it introduces a different issue: $\ad{in-U}_1$ and $\ad{in-U∼}_1$ give
+rise to equational constraints on their indices, in the form of proofs of the
+$\mProp$-valued identity type. The definition of the eliminators for $\ad{in-U}$
+and $\ad{in-U∼}$ relies on the ability to transport along these proofs, hence
+the need to extend our metatheory with a primitive, strong form of transport for
+$\textsf{Id}$.\footnote{Note that this issue cannot be solved by expressing the
+  equational constraints with an identity type in $\mType$, since the
+  well-typing predicates force it to necessarily be in $\mProp$.}
+
+Having $\mProp$ and a strong transport principle does help to some extent.
+%
+However, we would still need extensionality to derive the general eliminators
+for $\ad{in-U}$ and $\ad{in-U∼}$. In fact, as explained in the previous section,
+to derive the general recursive-recursive eliminators we need to prove that the
+correponding graph relations are functional, which can't be done without funext.
+
+Luckily, it turns out that the \emph{simple} elimination principle is sufficient
+for our purposes. This eliminator can be defined by simple pattern matching on
+the untyped codes, and does not require extensionality or any extra principles
+beyond strong transport.
 
@@
 
-TODO
+Previous work \cite{abel2019failure} seem to suggest that the strong transport
+rule for proof-irrelevant identity types is a non-trivial addition to the
+metatheory.
+%
+We would like to investigate a construction of the setoid universe that does not
+require such rule, not only to avoid unnecessary extensions to the metatheory,
+but also because we don't yet know whether $MLTT^{Prop}$ extended with this rule
+is normalizing.
+
+We want to figure out whether our inductive encoding of the setoid universe can
+support universe induction. A related question is whether the alternative
+definitions using induction-recursion and induction-induction can support it.
+
+SeTT \cite{mpc19} and XTT \cite{xtt} are both syntactic representations of the
+setoid model with similar design choices, like definitional
+proof-irrelevance. We would like to know whether their respective notions of
+models are equivalent, that is, whether we can obtain an XTT model from a SeTT
+model, and vice versa. Since XTT universes support universe induction, it is
+possible that we would need to extend our own universe with the same principle
+(see discussion in Section~\ref{setoid-model} and the previous paragraph). A
+further question is whether this mapping of models is functorial.
 

### 2020-06-30 paper.tex
More paper writing

@@
 
+\newcommand{\mlttp}{\textsf{MLTT}^{\mProp}}
 \newcommand{\refl}{\textsf{refl}}
@@
 
-\section{$MLTT^{Prop}$}\label{metatheory}
-
-This section describes $MLTT^{Prop}$, the metatheory in which the model
-construction described in the following sections takes place. We employ Agda
-notation for metatheoretic terms throughout the paper.
-
-One aspect of the metatheory that we pay attention to is it's minimality, as we
-want to keep it as simple as possible. In particular, we want to avoid having to
-assume strong definition schemas that go beyond inductive families.
-
-$MLTT^{Prop}$ is an intensional type theory in the style of Martin-L\"of Type
-Theory with $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$, and a universe
+\section{$\mlttp$}\label{metatheory}
+
+This section describes $\mlttp$, our ambient metatheory. We employ Agda notation
+to write down $\mlttp$ terms throughout the paper.
+
+One of the main appeals of Altenkirch's setoid model is that it can justify
+several useful extensionality principles while being defined in a small
+intensional metatheory.
+%
+We tried to stay true to this idea when figuring out the necessary
+metatheoretical tools for the universe construction in this paper. In
+particular, we wanted to avoid having to assume strong definition schemas that
+go beyond inductive families.
+
+$\mlttp$ is thus an intensional type theory in the style of Martin-L\"of type
+theory with $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$, and a universe
 of strict propositions $\mProp$. $\Sigma$ types are defined negatively by
@@
 rules for $\Pi, \Sigma$, and $\mathbf{1}$ types. Propositions can be lifted to
-types via $\textsf{Lift} : \mProp \to \mType$.
-
-%% The universe of propositions is
-%% inhabited by false $\bot$ and propositional truncation $\|\_\| : \mType \to
-%% \mProp$.
+types via a lifting $\textsf{Lift} : \mProp \to \mType$, with constructor
+$\textsf{lift} : \{P : \mProp \} \to P \to \textsf{Lift}\ P$ and destructor
+$\textsf{unlift} : \{P : \mProp \} \to \textsf{Lift}\ P \to P$.
 
@@
 
-In addition to the above type formers in $\mType$, we will need the
-propositional versions of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can
-be defined from their $\mType$ counterparts via truncation. That is, given $P :
-\mProp$ and $Q : P \to \mProp$:
+In addition to type formers in $\mType$, we will need the propositional versions
+of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can be defined from their
+$\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
+\mProp$:
 
@@
 
-Finally, we will assume the presence of an identity type in the style of
-Martin-L\"of's inductive identity type. The main difference is that our identity
-type is a $\mProp$-valued relation. Naturally, we also have a \textsf{transp}
-combinator representing the corresponding elimination principle.
+Finally, we will assume an identity type in the style of Martin-L\"of's
+inductive identity type. The main difference is that our identity type is a
+$\mProp$-valued relation. We also include a transport combinator
+\textsf{transp}.
 
@@
 
-Note that \textsf{transp} provides a strong elimination principle allowing to
-eliminate a strict proposition (the identity type) into arbitrary types. Indeed,
-the reason why we must assume this identity type as primitive is because, even
-though we could have defined \textsf{Id} from $W_p$, \textsf{transp} would not
-be definable from $W_p$ elimination. \todo{what is the justification for this?}
+The \textsf{transp} combinator provides a strong elimination principle allowing
+to eliminate a strict proposition (the identity type) into arbitrary types.
+%
+The reason to have an identity type in $\mProp$ with such a strong transport
+rule are made clear in Section~\ref{inductive-universe}.
+%
+Note that this identity type must necessarily be assumed as primitive; in fact,
+even though we could have defined \textsf{Id} from $W_p$, \textsf{transp} is not
+definable from $W_p$ elimination.
 
@@
 A universe of strict propositions has been recently added to the Agda proof
-assistant \cite{agda-prop}, thus making $MLTT^{Prop}$ a subset of Agda. In fact,
-most of the developments presented here have been formalized and proof-checked
-using Agda extended with strict propositions.
-%
-For convenience, we slightly deviate from $MLTT^{Prop}$ both in the paper and in
-the formalization, for instance by relying on pattern matching instead of
+assistant \cite{agda-prop}, making \emph{most} of $\mlttp$ a subset of Agda,
+with the exception of the proof-irrelevant identity type. Most of the universe
+constructions presented here have been formalized and proof-checked using Agda,
+with the proof-irrelevant identity type and the strong transport rule added via
+postulates and rewriting.
+
+For convenience, we slightly deviate from $\mlttp$ both in the paper and in the
+formalization, for instance by relying on pattern matching instead of
 eliminators, and using primitive versions of $\mProp$-valued $\Pi$ and $\Sigma$
 types instead of deriving them from truncation. We operate under the assumption
-that everything can be equivalently carried out in $MLTT^{Prop}$, although we
-haven't fully checked all the necessary details.
+that everything can be equivalently carried out in $\mlttp$, although we haven't
+fully checked all the necessary details.
 
@@
 
-By \emph{setoid model} we intend a class of models of Type Theory where
+By \emph{setoid model} we intend a class of models of type theory where
 contexts/closed types are interpreted as setoids, i.e. sets with an equivalence
@@
 %
-A setoid model was first given for intensional Type Theory by M. Hofmann in
-\cite{hofmann}, in order to give a semantics to extensionality
-principles such as function and propositional extensionality. Here we consider
-a similar model construction due to Altenkirch \cite{setoid99}.
-%
-The peculiarity of this model is that it is presented in a type theoretic and
-intensional metatheory which includes a strict universe of propositions.
-%
-The use of strict propositions to define the setoid equivalence relations makes
-this model \emph{strict}. A strict model is one where every equation holds
-definitionally. The setoid model thus defined validates function extensionality,
-a univalent universe of propositions, and quotient types. Therefore, it provides
-a way to bootstrap and ``explain'' extensionality, since the model construction
-effectively gives an implementation of various extensionality principles it in
-terms of a small, entirely intensional theory.
+A setoid model was first given for intensional type theory by M. Hofmann in
+\cite{hofmann}, in order to give a semantics to extensionality principles such
+as function and propositional extensionality.
+
+Here we consider a similar model construction due to Altenkirch
+\cite{setoid99}. The peculiarity of this model is that it is presented in a type
+theoretic and intensional metatheory which includes a strict universe of
+propositions.
+
+The setoid model thus defined validates function extensionality, a universe of
+propositions with propositional extensionality, and quotient types. Therefore,
+it provides a way to bootstrap and ``explain'' extensionality, since the model
+construction effectively gives an implementation of various extensionality
+principles it in terms of a small, completely intensional theory.
 
@@
 
-As any model of Type Theory, the setoid model can be framed categorically as a
-category with families (CwF) with extra structure for the various type and term
-formers \cite{internal-tt}. The core structure of a CwF can be given as the
-following signature:
+The setoid model can be framed categorically as a category with families (CwF,
+\cite{internal-tt}) with extra structure for the various type and term
+formers. The core structure of a CwF can be given as the following signature:
 
@@
     \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
-    & ... \\
+    & \textsf{refl*} : \{\gamma : | \GG | \} (a : | A | \gamma) \to A^\sim\ (\textsf{refl}\ \GG\ \gamma)\ a\ a \\
+    & \textsf{sym*} : \forall\{\gamma_0\,\gamma_1\,a_0\,a_1\}\{p : \GG^\sim\ \gamma_0\ \gamma_1\}
+    \to A^\sim\ p\ a_0\ a_1 \to A^\sim\ (\textsf{sym}\ \GG\ p)\ a_1\ a_0 \\
+    & \textsf{trans*} : \forall\{\gamma_0\,\gamma_1\,\gamma_2\,a_0\,a_1\,a_2\}\{p_0 : \GG^\sim\ \gamma_0\ \gamma_1\}\{p_1 : \GG^\sim\ \gamma_1\ \gamma_2\}\\
+    & \ \qquad \to A^\sim\ p_0\ a_0\ a_1\to A^\sim\ p_1\ a_1\ a_2\to A^\sim\ (\textsf{trans}\,\GG\,p_0\,p_1)\ a_0\ a_2\\
     & \textsf{coe} : \forall\{\gamma_0\ \gamma_1 : | \GG |\} \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
@@
 \cite{setoid99}, but it is equivalent to it \todo{add citation}. The main
-difference here is in the use of heterogeneous equality in the definition of
-types.
+difference here is in the use of a heterogeneous equivalence relation $A^\sim$
+in the definition of types.
 
@@
     & | t | : (\gamma : | \GG |) \to | A | \, \gamma \\
-    & t^\sim : ...
+    & t^\sim : \forall\{\gamma_0\, \gamma_1\}
+    (p : \GG^\sim\ \gamma_0\ \gamma_1) \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
   \end{aligned}
@@
 
-As defined, we can show that the setoid model validates the usual basic type
-formers ($\Pi, \Sigma$, etc.), including function extensionality and a universe
-of strict propositions satisfying propositional extensionality
-\cite{setoid99}. Note that we don't need neither identity types nor inductive
-types (W-types) in the metatheory to achieve this. However, W-types are needed
-when constructing the setoid universe in the next sections.
+We can show that the setoid model validates the usual basic type formers ($\Pi,
+\Sigma$, etc.), including function extensionality and a universe of strict
+propositions satisfying propositional extensionality \cite{setoid99}. Note that
+we don't need neither identity types nor inductive types (W-types) to do this
+much.
 
@@
 
-The setoid model presented in the previous section is peculiar in several
-ways. One is that it can be presented in a very small type theoretic
-metatheory. The other important aspect is that it is \emph{strict}, namely every
-equation of the original syntax holds by definition in the semantics. One
-advantage of strict models is that they can be turned into \emph{syntactic
-  translations}, in which syntactic objects of the source theory are interpreted
-as their counterparts in another \emph{target} theory. In the case of the setoid
-model, this gives rise to a \emph{setoid translation}, where source contexts are
+The setoid model presented in the previous section is \emph{strict}, that is,
+every equation of a CwF holds by definition in the semantics. One advantage of
+strict models is that they can be turned into \emph{syntactic translations}, in
+which syntactic objects of the source theory are interpreted as their
+counterparts in another \emph{target} theory. In the case of the setoid model,
+this gives rise to a \emph{setoid translation}, where source contexts are
 interpreted as target contexts together with a target type representing the
@@
 A setoid translation is used in \cite{mpc19} to justify Setoid Type Theory
-(SeTT), an extension of $MLTT^{Prop}$ (without W-types) with equality types for
-contexts and dependent types that reflect the setoid equality of the model in
-the syntax.
-
-We recall the rules of SeTT that extend $MLTT^{Prop}$ below, but with a
-variation. Whereas the equality types in \cite{mpc19} are stated as elements of
-the universe of propositions internal to SeTT itself, we here state the context
-equalities as elements of the external, metatheoretic universe
-$\mProp$. Equality on dependent types is defined as before.
+(SeTT), an extension of Martin-L\"of type theory (+ $\mProp$) with equality
+types for contexts and dependent types that reflect the setoid equality of the
+model.
+
+We recall the rules of SeTT that extend regular MLTT below, but with a
+variation: whereas the equality types in \cite{mpc19} are stated as elements of
+SeTT's internal universe of propositions, here we state the context equalities
+as elements of the external, metatheoretic universe $\mProp$. Equality on
+dependent types is defined as before in \cite{mpc19}.
 
@@
 
-Notably, equality constructors in SeTT compute definitionally on concrete type
+Notably, equality types in SeTT compute definitionally on concrete type
 formers. In particular, they compute to their obvious intended meaning, so that
@@
 and propositional extensionality.
-%
-\begin{align*}
-  & (\Pi\ A\ B)^\sim\ p\ f_0\ f_1 = ... \\
-  & (\Sigma\ A\ B)^\sim\ p\ (a_0, b_0)\ (a_1, b_1) = ... \\
-  & \Prop^\sim\ p\ P\ Q = ...
-\end{align*}
+%% %
+%% \begin{align*}
+%%   & (\Pi(a : A)\ B)^\sim\ p\ f_0\ f_1 = \Pi(a_0\,a_1 : A)... \\
+%%   & (\Sigma\ A\ B)^\sim\ p\ (a_0, b_0)\ (a_1, b_1) = ... \\
+%%   & \Prop^\sim\ p\ P\ Q = ...
+%% \end{align*}
 
@@
 
-Note that contractibility of singletons holds in SeTT by UIP and the computation
-rules of equality on $\Sigma$ types. Hence, Martin-L\"of's J eliminator can also
-be derived \todo{add citation}.
+We can also derive Martin-L\"of's J eliminator for this homogeneous identity
+type.
 
@@
 the construction of the universe. In fact, one of the main aspects of the setoid
-model construction recalled in Section TODO and shown originally in
-\cite{setoid99} is that it can be carried out in a very small type theoretic
+model construction recalled in Section~\ref{setoid-model} and shown originally
+in \cite{setoid99} is that it can be carried out in a very small type theoretic
 metatheory, thus providing a way to reduce extensionality to a small intensional
@@
 
-A known and established method for defining universes in Type Theory relies on
+A known and established method for defining universes in type theory relies on
 induction-recursion (IR), a definition schema developed by Dybjer
@@
 
-Although universe constructions in Type Theory---including our own setoid
-universe---are naturally presented as inductive-recursive definitions, they
-may not necessarily require a metatheory that supports induction-recursion. In
-fact, it is possible to reduce some instances of induction-recursion to plain
-induction (more specifically, inductive families), including some universe
-definitions. We recall this reduction in Section~\ref{ir-universes}.
+Although universe constructions in type theory---including our own setoid
+universe---are naturally presented as inductive-recursive definitions, they may
+not necessarily require a metatheory with induction-recursion. In fact, it is
+possible to reduce some instances of induction-recursion to plain induction
+(more specifically, inductive families), including some universe definitions. We
+recall this reduction in Section~\ref{ir-universes}.
 
@@
 %
-The question arises because we only define a weak eliminator for our final
-inductive-inductive encoding of the setoid universe, for reasons that are
-explained in Section~\ref{inductive-universe}. Although not currently needed, a
-stronger eliminator might be necessary to justify universe induction. This
-problem should not arise in the other encodings of the setoid universe (as given
-in Section~\ref{ir-setoid-universe} and Section~\ref{ii-universe}).
+The question arises because our final encoding of the setoid universe only
+supports a weak form of elimination, for reasons that are explained in
+Section~\ref{inductive-universe}. Although not currently needed, a stronger
+eliminator might be necessary to justify universe induction. This problem should
+not arise in the other encodings of the setoid universe (as given in
+Section~\ref{ir-setoid-universe} and Section~\ref{ii-universe}).
 
@@
 
-\subsection{Inductive-recursive universes in Type Theory}\label{ir-universes}
+\subsection{Inductive-recursive universes}\label{ir-universes}
 
@@
 
-Induction-recursion is arguably a nicer and more natural way to define
-internal universes in Type Theory, however it is not always strictly required.
+Induction-recursion is arguably a nicer and more natural way to define internal
+universes in type theory, however it is not always strictly required.
 %
@@
 In our case, we can encode $\U$ as an inductive type $\ad{in-U}$ that
-\emph{carves out} all types in $\Set$ that are in the image of $\El$. In other
+\emph{carves out} all types in $\mType$ that are in the image of $\El$. In other
 words, $\ad{in-U}$ is a predicate that holds for any type that would have been
 obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed on
-the type of codes, the definition of $\ad{in-U}$ quite expectedly reflecs the
+the type of codes, the definition of $\ad{in-U}$ quite expectedly reflects the
 inductive structure of codes.
@@
 \end{align*}
-Using this $\U$ and $\El$ can be given as follows:
+
+$\U$ and $\El$ can be given as follows:
 \begin{align*}
   \U & = \Sigma (A : \mType)\ (\textsf{in-U}\ A) \\
-  \El & = \textsf{proj}_1
+  \El & = \pi_1
 \end{align*}
@@
 Note that this construction gives rise to a universe in $\mType_1$, rather than
-$\mType$, since the definition of $\U$ quantifies over all possible sets. Hence
-this kind of construction requires a metatheory with at least one universe.
-
-\subsection{A first definition of the setoid universe}\label{ir-setoid-universe}
+$\mType$, since the definition of $\U$ quantifies over all possible types in
+$\mType$. Hence this kind of construction requires a metatheory with at least
+one universe.
+
+\subsection{Inductive-recursive setoid universe}\label{ir-setoid-universe}
 
@@
 
-[todo... explanation]
+We are now ready to give a full definition of the setoid universe:
 %
@@
 
-We can show that $\Uty$ is closed under $\Pi$ types and \textsf{Bool}, including
-$\beta$ and $\eta$ laws. Moreover, all the necessary equations hold
-definitionally in the metatheory.
+We can show that $\Uty$ is closed under $\Pi$ types and unit types, including
+$\beta$ and $\eta$ laws. Moreover, all the necessary equations (including
+substitution laws) hold definitionally in the metatheory.
 
@@
 inductive-recursive. Moreover, the kind of recursion involved in this definition
-is particularly complex, and not obviously reducible to understood notions of
-induction-recursion like in \cite{dybjer-setzer}. Nevertheless, we would like to
-avoid extending the metatheory with any forms of induction-recursion, as per our
-goal to keep the metatheory as small and essential as possible.
-
-As a consequence, in the next section we transform our current
-inductive-recursive definition to one that doesn't use induction-recursion.
+is particularly complex, and not obviously reducible to well-understood notions
+of induction-recursion like the one described in \cite{dybjer-setzer}. In any
+case, we would like to avoid extending the metatheory with any forms of
+induction-recursion, given our goal to keep the metatheory as small and
+essential as possible.
+
+In the next section we transform our current inductive-recursive definition to
+one that doesn't use induction-recursion.
 %
@@
 novel way to account for the presence of $\mProp$-valued types. To our
-knowledge, this is the first time this reduction trick is applied to an
+knowledge, this is the first time this reduction method is applied to an
 inductive-recursive type of this kind.
 
-\subsection{Inductive-inductive universe}\label{ii-universe}
-
-We will follows the same method outlined in Section~\ref{ir-universes}. In
+\subsection{Inductive-inductive setoid universe}\label{ii-universe}
+
+We will follow the same method outlined in Section~\ref{ir-universes}. In
 addition to $\ad{in-U}$ for defining $\U$, we also introduce a family
@@
 \begin{align*}
-  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
-  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1
+  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
+  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \mType\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \mProp) \to \mType_1
 \end{align*}
@@
 that allows the definition of a type mutually with other types that contain the
-first one in their signature \cite{ii}. The main example of induction-induction
-can be found in intrinsic definitions of Type Theory in Type Theory
-\cite{tt-in-tt}.
+first one in their signature \cite{ii}.\footnote{The main example of
+  induction-induction is the intrinsic definition of a dependent type theory in
+  type theory \cite{tt-in-tt}.}
 
@@
 
-\subsection{Inductive universe}
+\subsection{Inductive setoid universe}
+
+This section outlines a way to encode the inductive-inductive universe of
+setoids from the previous section without assuming arbitrary inductive-inductive
+definitions in the metatheory.
+
+Before turning our attention to the setoid universe, we recall the known,
+systematic method to reduce finitary inductive-inductive types to inductive
+families.
 
@@
 reduce finitary inductive-inductive definitions to inductive families. However,
-it is not clear whether this method extends to infinitary induction-induction,
-of which the setoid universe defined in Section~\ref{ii-universe} is an
-instance.
+it is not clear whether this method extends to \emph{infinitary}
+induction-induction, of which the setoid universe defined in
+Section~\ref{ii-universe} is an instance.
 %
@@
 \textsf{Ty}\ \GG$. Already with this simple example, we run into problems as
-soon as we try to define the general eliminator, since it requires function
-extensionality. \todo{Is this true?}
-
-Luckily, in our case we can take advantage of our specific circumstances. First
-of all, we can define the well-typing predicates as $\mProp$-valued families,
-rather than in $\mType$:
+soon as we try to define the eliminator. One issue is that the definition of the
+eliminator relies of a proof that the well-typing predicates $\ad{in-U}_1,
+\ad{in-U∼}_1$ are propositional, that is, any two of their elements are equal.
+%
+Without further assumptions this proof can only be done by induction, and
+requires function extensionality since these predicates include higher-order
+constructors.
+
+One way to get around this is to define the well-typing predicates as
+$\mProp$-valued families, rather than in $\mType$:
 
@@
 \end{align*}
-\todo{check that these signatures are correct}
-
-Defining the well-typing predicates in $\mProp$ does help, since otherwise we
-would need function extensionality to prove that these types are propositions
-(with respect to metatheoretic propositional equality).
+
+Using $\mProp$ avoid the issue of proving propositionality altogether, since the
+predicates are now propositional by definition.
 %
@@
 
-\section{Further work}\label{further-work}
-
-Previous work \cite{abel2019failure} seem to suggest that the strong transport
+Once the inductive encoding of the inductive-inductive universe is done, the
+setoid universe can be defined just as in Section~\ref{ii-universe}.
+
+\section{Conclusions and further work}\label{further-work}
+
+We have described the construction of a universe of setoids in the setoid model
+of type theory; this is given in several steps, first as an inductive-recursive
+definition, then as an inductive-inductive definition, and finally as an
+inductive type. Every encoding is obtained from the previous by adapting known
+data type transformation methods in a novel way that accounts for the
+peculiarities of our construction.
+
+Previous work \cite{abel2019failure} seems to suggest that the strong transport
 rule for proof-irrelevant identity types is a non-trivial addition to the
@@
 
+The groupoid model of type theory \cite{groupoid} can be seen as a
+generalization of the setoid model, since groupoids are essentially
+proof-relevant setoids.
+%
+One question is therefore whether it is possible to design a type theory
+internalizing the groupoid model, in the same way that SeTT represents a syntax
+for the setoid model. A further question is whether such ``groupoid type
+theory'' can be justified, similarly to SeTT, via a syntactic translation,
+perhaps with SeTT itself as the target theory.
+
 \bibliographystyle{alpha}

### 2020-06-30 paper.tex
More paper writing

@@
 
-Previous work \cite{abel2019failure} seems to suggest that the strong transport
-rule for proof-irrelevant identity types is a non-trivial addition to the
-metatheory.
-%
-We would like to investigate a construction of the setoid universe that does not
-require such rule, not only to avoid unnecessary extensions to the metatheory,
-but also because we don't yet know whether $MLTT^{Prop}$ extended with this rule
-is normalizing.
-
-We want to figure out whether our inductive encoding of the setoid universe can
-support universe induction. A related question is whether the alternative
-definitions using induction-recursion and induction-induction can support it.
-
-SeTT \cite{mpc19} and XTT \cite{xtt} are both syntactic representations of the
-setoid model with similar design choices, like definitional
-proof-irrelevance. We would like to know whether their respective notions of
-models are equivalent, that is, whether we can obtain an XTT model from a SeTT
-model, and vice versa. Since XTT universes support universe induction, it is
-possible that we would need to extend our own universe with the same principle
-(see discussion in Section~\ref{setoid-model} and the previous paragraph). A
-further question is whether this mapping of models is functorial.
+As further work, we would like to investigate a construction of the setoid
+universe that does not require a strong transport rule for the proof-irrelevant
+identity type. Not only to avoid unnecessary extensions to the metatheory, but
+also because we don't yet know whether $\mlttp$ extended with this rule is
+normalizing.
+%
+Indeed, previous work in this respect \cite{abel2019failure} seems to suggest
+that this rule is a non-trivial addition to the metatheory.
+
+Additionally, we want to figure out whether our inductive encoding of the setoid
+universe can support universe induction. A related question is whether the
+alternative definitions using induction-recursion and induction-induction can
+support it.
+
+Another question regards the relationship between SeTT \cite{mpc19} and XTT
+\cite{xtt}. Both systems are syntactic representations of the setoid model with
+similar design choices, like definitional proof-irrelevance. We would like to
+know whether their respective notions of models are equivalent, that is, whether
+we can obtain an XTT model from a SeTT model, and vice versa. Since XTT
+universes support universe induction, it is possible that we would need to
+extend our own universe with the same principle (see discussion in
+Section~\ref{setoid-model} and the previous paragraph). A further question is
+whether this mapping of models is functorial.
 

### 2020-06-30 paper.tex
Add reference to agda formalization
Reduce space

@@
 
+Finally, the mathematical contents of this paper have been formalized in the
+proof-assistant Agda (see \cite{agda-code}).
+
 \paragraph{Structure of the paper}
@@
 
-We also require indexed W-types, both in $\mType$ and $\mProp$:
-
-\begin{align*}
-  W & : (S : \mType) \to (S \to \mType) \to \mType \\
-  W_p & : (S : \mType) \to (S \to \mType) \to \mProp
-\end{align*}
-
-From $W_p$ we can define propositional truncation:
-
-\begin{align*}
-  & \|\_\| : \mType \to \mProp \\
-  & |\_| : \{A : \mType \} \to A \to \| A \| \\
-  & \textsf{elim}_{\|\_\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P
-\end{align*}
+We also require indexed W-types, both in $\mType$ and $\mProp$; hence we have $W
+: (S : \mType) \to (S \to \mType) \to \mType$ and $W_p : (S : \mType) \to (S \to
+\mType) \to \mProp$.
+%
+From $W_p$ we can define propositional truncation $\|\_\| : \mType \to \mProp$, with constructor
+  $|\_| : \{A : \mType \} \to A \to \| A \|$ and eliminator
+  $\textsf{elim}_{\|\_\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P$.
 
@@
 \mProp$:
-
+%
 \begin{align*}
@@
 \textsf{transp}.
-
+%
 \begin{align*}
@@
 with the proof-irrelevant identity type and the strong transport rule added via
-postulates and rewriting.
+postulates and rewriting. The formalization can be found in \cite{agda-code}.
 
@@
 follow from proof-irrelevance in the metatheory, and thus hold definitionally.
-
+\\
+~
+\\
+\begin{minipage}{0.5\textwidth}
 \[
@@
     & | \sigma | : | \GG | \to | \GD | \\
-    & \sigma^\sim : \forall\{\rho_0\, \rho_1\} \to \GG^\sim\, \rho_0\, \rho_1 \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
+    & \sigma^\sim : \forall\{\rho_0\, \rho_1\} \to \GG^\sim\, \rho_0\, \rho_1 \\
+    & \quad \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
   \end{aligned}
@@
 \]
-
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
 \[
@@
     & t^\sim : \forall\{\gamma_0\, \gamma_1\}
-    (p : \GG^\sim\ \gamma_0\ \gamma_1) \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
+    (p : \GG^\sim\ \gamma_0\ \gamma_1) \\
+    & \quad \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
   \end{aligned}
@@
 \]
-
+\end{minipage}
+\\
+~
+\\
 We can show that the setoid model validates the usual basic type formers ($\Pi,
@@
 
+\begin{minipage}{0.45\textwidth}
 \begin{align*}
@@
   & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
-  & \El\ \textsf{bool} = \mathbbm{2} \\
-  & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
-\end{align*}
+\end{align*}
+\end{minipage}
+\begin{minipage}{0.45\textwidth}
+\begin{align*}
+  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
+  & \El\ (\textsf{pi}\ A\ B) :\equiv (a : \El\ A) \to \El\ (B\ a)
+\end{align*}
+\end{minipage}
 
@@
 $\U$ and $\El$ can be given as follows:
-\begin{align*}
-  \U & = \Sigma (A : \mType)\ (\textsf{in-U}\ A) \\
-  \El & = \pi_1
-\end{align*}
+\[
+\U = \Sigma (A : \mType)\ (\textsf{in-U}\ A) \qquad \El = \pi_1
+\]
 
@@
 setoid; in particular, we need a type of codes together with an equivalence
-relation reflecting equality of codes:
-%
+relation reflecting equality of codes, in addition to proofs that these are
+indeed equivalence relations:
+%
+
+\begin{minipage}{0.5\textwidth}
 \begin{align*}
@@
 \end{align*}
-%
-in addition to proofs that these are indeed equivalence relations
-%
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
 \begin{align*}
@@
 \end{align*}
+\end{minipage}
+\\
+~
+\\
 
@@
 definitions of $\setoidU$ and $\El$:
-%
+
+\begin{minipage}{0.5\textwidth}
 \begin{align*}
@@
   :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
-  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x' \to B\ x \sim_\setoidU B\ x')
+  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
+  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
   \to \setoidU
 \end{align*}
-%
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
 \begin{align*}
@@
   & \qquad \quad
-  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \to \eqel{h\ p}{f\ x}{f\ x'})
-\end{align*}
+  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
+  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
+\end{align*}
+\end{minipage}
 
@@
 
-We are now ready to give a full definition of the setoid universe:
-%
+We can now give a full definition of the setoid universe, and of $\EL\ A$ for
+any $A : \Tm\ \GG\ \Uty$:
+%
+
+\begin{minipage}{0.5\textwidth}
 \begin{align*}
@@
 \end{align*}
-
-Given $A : \Tm\ \GG\ \Uty$, we define $\EL\ A$ as follows:
-%
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
 \begin{align*}
@@
 
+\begin{minipage}{0.4\textwidth}
 \begin{align*}
   & \setoidU : \mType_1 \\
-  & \setoidU = \Sigma\ (X : \mType)\ (\ad{in-U}\ X) \\
+  & \setoidU = \Sigma\ (X : \mType)\ (\ad{in-U}\ X)
+\end{align*}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}
+\begin{align*}
   & \El : \setoidU \to \mType \\
-  & \El = \textsf{proj}_1
-\end{align*}
+  & \El = \pi_1
+\end{align*}
+\end{minipage}
+\\
+~
+\\
 
@@
 expect $A \sim_{\setoidU} B$ to be defined as follows:
-
-\[
-A \sim_{\setoidU} B :\equiv \Sigma (R: \El\,A \to \El\,B \to \mProp)(\ad{in-U∼}\ (\pi_2\,A)\ (\pi_2\,B)\ R)
-\]
-
-or, using pattern matching:
 
@@
 
+\begin{minipage}{0.4\textwidth}
 \begin{align*}
@@
   & \Pi_0 : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \\
+\end{align*}
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
+\begin{align*}
   & \bullet_1 : \textsf{Con}_1\ \bullet_0 \\
-  & \_,_1\_ : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
+  & \_,_1\_ : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
+  & \qquad \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
   & \iota_1 : \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
-  & \Pi_1 : \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
-\end{align*}
+  & \Pi_1 : \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \\
+  & \qquad \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \\
+  & \qquad \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
+\end{align*}
+\end{minipage}
+\\
+~
+\\
 
@@
 data type transformation methods in a novel way that accounts for the
-peculiarities of our construction.
+peculiarities of our construction. The mathematical contents of this paper have
+been formalized in the proof-assistant Agda (see \cite{agda-code}).
 

### 2020-07-04 paper.tex
Adapt section Indcutve-recursive

@@
 generalization of the simple inductive-recursive definition just shown. We only
-consider a very small universe with unit type $\mathds{1}$ and $\Pi$ for
+consider a very small universe with bool type $\mathbbm{2}$ and $\Pi$ for
 simplicity; a more realistic universe can be found in the Agda formalization,
@@
 
-We give an inductive definition of $\setoidU$, mutually with a recursive
-definition of everything else. Below we only show the more interesting
+We give an inductive definition of $\setoidU$, mutually with a recursive definition
+of the 4 functions
+$\_\sim_\setoidU\_$, $\textsf{refl}_\setoidU$, $\El$ and $\_\vdash\_\sim_\El\_$.
+The other functions are then recursively defined:
+\begin{itemize}
+\item $\textsf{refl}_\El$ alone,
+\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
+\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
+\end{itemize}
+The whole construction is quite long, below we only show the more interesting
 definitions of $\setoidU$ and $\El$:
@@
   & \textsf{data}\ \setoidU : \Set_1 \\
-  & \quad \textsf{one} : \setoidU \\
+  & \quad \textsf{bool} : \setoidU \\
   & \quad \textsf{pi}
@@
 \begin{align*}
-  & \El\ \textsf{one} :\equiv \textbf{1} \\
+  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
   & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
@@
 
-We can show that $\Uty$ is closed under $\Pi$ types and unit types, including
+We can show that $\Uty$ is closed under $\Pi$ types and bool type, including
 $\beta$ and $\eta$ laws. Moreover, all the necessary equations (including
 substitution laws) hold definitionally in the metatheory.
+The universe can be closed under more constructions if more codes are added to
+$\setoidU$.
 

### 2020-07-04 paper.tex
Small corrections

@@
 unfortunately equality reflection makes the problem of type-checking ETT terms
-computationally unfeasible, if not impossible.
+computationally unfeasible: it is undecidable.
 
@@
 properties like decidable type checking that can make it more suitable for
-computer implementation, but as normally defined (for example, in \cite{mltt75})
+computer implementation, but as usually defined (for example, in \cite{mltt75})
 it severely lacks extensionality.
@@
 
-A first fundamental consideration has to with the very definition of the setoid
+A first fundamental consideration has to do with the very definition of the setoid
 universe: as any type in the setoid model, this universe must be a setoid and
@@
 
-In the definition above, $p$ has type $\|| \Sigma (R: \El\,A \to \El\,B \to
-\mProp)(...) \||$. If the type wasn't propositionally truncated, we could define
+In the definition above, $p$ has type $\parallel \Sigma\ (R: \El\,A \to \El\,B \to
+\mProp)\ (...) \parallel$. If the type wasn't propositionally truncated, we could define
 $p \vdash a \sim_{\El} b$ by extracting the relation out of the first component

### 2020-07-04 paper.tex
small stuff

@@
 
-\newcommand{\metaprop}{\mathbf{Prop}}
-
 \newcommand{\List}{\mathrm{List}}
@@
   \Pi_p\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
-  \Sigma_P\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
+  \Sigma_p\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
   \mathbf{1}_p & :\equiv \| \mathbf{1} \|
@@
   \Ty &: (\GG : \Con) \to \mType \\
-  \Sub &: (\GG \GD : \Con) \to \mType \\
+  \Sub &: (\GG\ \GD : \Con) \to \mType \\
   \Tm &: (\GG : \Con) \to \Ty\ \GG \to \mType
@@
 In our presentation of the setoid model, contexts are given by setoids, that is
-types together with an equivalence relation.
+types together with an equivalence relation. A key point of the model is that the
+equivalence relation is valued in $\mProp$ and is thus definitionally proof
+irrelevant.
 
@@
 interpreted as target contexts together with a target type representing the
-equivalence relation, and so on.
-\footnote{%
+equivalence relation, and so on.\footnote{%
 Semantically, this translation corresponds to a model construction, in
@@
 that every context and type comes equipped with a setoid equivalence
-relation. As in the model, equality for dependent types is indexed over context
+relation. Note that $\mProp$ is the universe of the metatheory while $\Prop$ is
+the internal one. As in the model, equality for dependent types is indexed over context
 equality.
@@
 \infer{
-  \GG^\sim\ \rho_0\ \rho_1 : \metaprop
+  \GG^\sim\ \rho_0\ \rho_1 : \mProp
 }{
@@
 We can also derive Martin-L\"of's J eliminator for this homogeneous identity
-type.
+type. The only caveat is that $\textsf{transp}$ and the J eliminator do not
+compute definitionally on reflexivity.
 
@@
 \begin{align*}
+  & \El\ : \U \to \mType \\
   & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
-  & \El\ (\textsf{pi}\ A\ B) :\equiv (a : \El\ A) \to \El\ (B\ a)
+  & \El\ (\textsf{pi}\ A\ B) :\equiv (a : \El\ A) \to \El\ (B\ a) \\
 \end{align*}
@@
 \begin{align*}
-  & \El : \setoidU \to \Set \\
-  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
+  & \El : \setoidU \to \mType \\
+  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
 \end{align*}
@@
 We also require data and proofs that make sure we get setoids out of $\El$:
-
 \begin{align*}
@@
 \end{align*}
-
+%
 We give an inductive definition of $\setoidU$, mutually with a recursive definition
@@
 Two codes $A, B$ in the universe $\setoidU$ are equal when there exists a setoid
-equivalence relation on their respective sets $\El\ A$ and $\El B$. Intuitively,
+equivalence relation on their respective sets $\El\ A$ and $\El\ B$. Intuitively,
 since elements of a setoid are only ever compared to elements of the same
-setoid, this should only be possible if $A, B$ are codes for the same setoid,
+setoid, this should only be possible if $A$ and $B$ are codes for the same setoid,
 that is, if $A \sim_{\setoidU} B$. Existence and well-formedness of such
@@
 Unfortunately this definition only manages to capture the idea, but does not
-actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be $\Prop_1$-valued
+actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be $\mProp_1$-valued
 relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
@@
 One possible solution is actually quite simple, and it just involves truncating
-the $\Sigma$ type above to force it to be in $\Prop_1$.
-%
-\begin{align*}
-  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \Prop_1 \\
+the $\Sigma$ type above to force it to be in $\mProp_1$.
+%
+\begin{align*}
+  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \mProp_1 \\
   & (X , p) \sim_{\setoidU} (X' , p') =
-   \ \parallel \Sigma\ (R : X \to X' \to \Prop)\ (\ad{in-U∼}\ p\ p'\ R) \parallel
+   \ \parallel \Sigma\ (R : X \to X' \to \mProp)\ (\ad{in-U∼}\ p\ p'\ R) \parallel
 \end{align*}

### 2020-07-06 paper.tex
typos

@@
 The first step towards a solution is to give up the idea of propositional
-equality as a single inductive definition given homogeneously for an arbitrary
-type. Instead, equality should be \emph{specific} to each type former in the
+equality as a single inductive definition given homogeneously for arbitrary
+types. Instead, equality should be \emph{specific} to each type former in the
 type theory, or in other words, every type former should be introduced alongside
@@
 and funext hold in the setoid model. Moreover, it was shown by Altenkirch
-\cite{setoid99} that is the model construction is carried out in a type
+\cite{setoid99} that if the model construction is carried out in a type
 theoretic metatheory with a universe of strict (definitionally proof-irrelevant)

### 2020-07-06 paper.tex
typos

@@
 first examples of two kinds of data type reductions in an intensional
-metatheory: the first involving an inductive-recursive type which including
+metatheory: the first involving an inductive-recursive type which includes
 strict propositions, and the second involving an infinitary inductive-inductive
@@
 
-\section{$\mlttp$}\label{metatheory}
+\section{$\mlttp$}\label{sec:metatheory}
 
@@
 
-\section{Setoid model}\label{setoid-model}
+\section{Setoid model}\label{sec:setoid-model}
 

### 2020-07-06 paper.tex
small fixes

@@
 We then provide, in Section~\ref{ir-setoid-universe}, a first complete
-definition of the setoid universe using a form of induction-recursion. This is
-followed by an alternative definition of the universe in
+definition of the setoid universe using a special form of induction-recursion.
+This form of induction-recursion is not known to be reducible to plain inductive types.
+Then we describe an alternative definition of the universe in
 Section~\ref{ii-universe}, that does not rely on induction-recursion but instead
@@
 %
-In \cite{abel:nbe09}, Abel given a proof of normalization by evaluation for the
+In \cite{abel:nbe09}, Abel gives a proof of normalization for the
 Logical Framework extended with a similar proof-irrelevant equality type.

### 2020-07-06 paper.tex
small fixes

@@
 : (S : \mType) \to (S \to \mType) \to \mType$ and $W_p : (S : \mType) \to (S \to
-\mType) \to \mProp$.
+\mType) \to \mProp$. The elimination principle of $W_p$ only allows defining functions into $\mProp$s.
 %
@@
 
-By \emph{setoid model} we intend a class of models of type theory where
+By \emph{setoid model} we mean a class of models of type theory where
 contexts/closed types are interpreted as setoids, i.e. sets with an equivalence
@@
 %
-A setoid model was first given for intensional type theory by M. Hofmann in
-\cite{hofmann}, in order to give a semantics to extensionality principles such
+A setoid model was first given for intensional type theory by M.\ Hofmann
+\cite{hofmann}, in order to provide a semantics for extensionality principles such
 as function and propositional extensionality.
@@
 construction effectively gives an implementation of various extensionality
-principles it in terms of a small, completely intensional theory.
+principles in terms of a small, completely intensional theory.
 
@@
 
-In our presentation of the setoid model, contexts are given by setoids, that is
+In our presentation of the setoid model, contexts are given by setoids, that is,
 types together with an equivalence relation. A key point of the model is that the
@@
 This definition of types in the setoid model is different from the one in
-\cite{setoid99}, but it is equivalent to it \todo{add citation}. The main
+\cite{setoid99}, but it is equivalent to it \cite[Section 1.6.1]{simonthesis}. The main
 difference here is in the use of a heterogeneous equivalence relation $A^\sim$
@@
 propositions satisfying propositional extensionality \cite{setoid99}. Note that
-we don't need neither identity types nor inductive types (W-types) to do this
-much.
+we don't need identity types or inductive types (W-types) for this.
 

### 2020-07-06 paper.tex
small fixes

@@
 }
+\qquad
+\infer{
+  u \equiv v
+}{
+  u : \Tm\ \GG\ \underline{P} && v : \Tm\ \GG\ \underline{P}
+}
 \]
@@
 the construction of the universe. In fact, one of the main aspects of the setoid
-model construction recalled in Section~\ref{setoid-model} and shown originally
+model construction recalled in Section~\ref{sec:setoid-model} and shown originally
 in \cite{setoid99} is that it can be carried out in a very small type theoretic
@@
 semantics, that is, if our encoding of the setoid universe in the model allows
-to define such universe eliminator.
+to define such a universe eliminator.
 %
@@
 Another design choice has to do with how the setoid universe relates to the
-other universes. One could have $\Prop$ be a code of the setoid universe, rather
-than a type. Moreover, the setoid universes could form a hierarchy, possibly
+other universes. One could provide a code for $\Prop$ in the setoid universe.
+Moreover, the setoid universes could form a hierarchy, possibly
 cumulative.
@@
 using the equivalence of $I$-indexed families of types and types over $I$
-(that is, $\mType_i \to I$).
+(that is, $\mType_i \to I$) \cite{malatasta13smallir}.
 
@@
 words, $\ad{in-U}$ is a predicate that holds for any type that would have been
-obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed on
+obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed by
 the type of codes, the definition of $\ad{in-U}$ quite expectedly reflects the

### 2020-07-06 paper.tex
small fixes

@@
 on infinitary induction-induction. This inductive-inductive encoding of the
-universe is obtained from the inductive-recursive one, following the method of
+universe is obtained from the inductive-recursive one, inspired by the method of
 Section~\ref{ir-universes}.
@@
 consider a very small universe with bool type $\mathbbm{2}$ and $\Pi$ for
-simplicity; a more realistic universe can be found in the Agda formalization,
-and includes more type formers.
+simplicity; a more realistic universe that includes more type formers
+can be found in the Agda formalization.
 
@@
 \Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
-Section~\ref{setoid-model}, these are essentially record types made of several
+Section~\ref{sec:setoid-model}, these are essentially record types made of several
 components. Since $\Uty$ is a closed type, it requires the same data of a
@@
 
-We will follow the same method outlined in Section~\ref{ir-universes}. In
+We will follow the method outlined in Section~\ref{ir-universes}. In
 addition to $\ad{in-U}$ for defining $\U$, we also introduce a family
@@
 %
-To illustrate the idea, let us consider a well-known example of finitary
+To illustrate the idea, let us consider a well-known example of a finitary
 inductive-inductive type, the intrinsic encoding of type theory in type theory
-itself. Actually, we only need to consider the type of contexts $\textsf{Con} :
+itself. Actually, we only consider the type of contexts $\textsf{Con} :
 \mType$ and the type of types $\textsf{Ty} : \textsf{Con} \to \mType$; since the
@@
 The general method to eliminate induction-induction is to split the original
-inductive-inductive types into a type of codes and an associated well-formedness
-predicate. In our \textsf{Con/Ty} example, these would be respectively given by
+inductive-inductive types into a type of codes and associated well-formedness
+predicates. In our \textsf{Con/Ty} example, these would be respectively given by
 codes $\textsf{Con}_0, \textsf{Ty}_0 : \mType$ and predicates $\textsf{Con}_1 :
@@
 The next step is to prove that these relations are functional, by induction on
-the untyped codes $\textsf{Con}_0$ and $\textsf{Ty}_0$. From this result,
+the untyped codes $\textsf{Con}_0$ and $\textsf{Ty}_0$ \cite{induction-is-enough}. From this result,
 defining the eliminators is immediate.
@@
 soon as we try to define the eliminator. One issue is that the definition of the
-eliminator relies of a proof that the well-typing predicates $\ad{in-U}_1,
+eliminator relies on a proof that the well-typing predicates $\ad{in-U}_1,
 \ad{in-U∼}_1$ are propositional, that is, any two of their elements are equal.
@@
 
-Using $\mProp$ avoid the issue of proving propositionality altogether, since the
+Using $\mProp$ avoids the issue of proving propositionality altogether, since the
 predicates are now propositional by definition.
@@
 extend our own universe with the same principle (see discussion in
-Section~\ref{setoid-model} and the previous paragraph). A further question is
+Section~\ref{sec:setoid-model} and the previous paragraph). A further question is
 whether this mapping of models is functorial.

### 2020-07-06 paper.tex
Phrasing.

@@
 similar design choices, like definitional proof-irrelevance. We would like to
-know whether their respective notions of models are equivalent, that is, whether
-we can obtain an XTT model from a SeTT model, and vice versa. Since XTT
-universes support universe induction, it is possible that we would need to
+know whether their respective notions of models are equivalent.
+In particular, we would like to obtain an XTT model from a SeTT model, and vice versa. Since XTT
+universes support universe induction, for one direction we would need to
 extend our own universe with the same principle (see discussion in

### 2020-07-06 paper.tex
changes discussed today

@@
 %
-Unfortunately, one can show that canonicity and function extensionality cannot
-coexist if propositional equality is given by Martin-L\"of's inductively-defined
-identity types.
+Unfortunately, canonicity for Martin-L{\"o}f's inductively defined
+identity type says that if two terms are propositionally equal in the
+empty context, then they are also definitionally equal. This rules out
+function extensionality.
 %
 The first step towards a solution is to give up the idea of propositional
-equality as a single inductive definition given homogeneously for arbitrary
+equality as a single inductive definition given generically for arbitrary
 types. Instead, equality should be \emph{specific} to each type former in the
@@
 natural and useful semantic domain in which to interpret type theory. The first
-setoid model was constructed to show consistency of funext for ITT
-\cite{hofmann}. This consistency result is implied by the observation that UIP
-and funext hold in the setoid model. Moreover, it was shown by Altenkirch
+setoid model was constructed to justify function extensionality without relying
+on funext in the metatheory \cite{hofmann}. Moreover, it was shown by Altenkirch
 \cite{setoid99} that if the model construction is carried out in a type
@@
 In addition to type formers in $\mType$, we will need the propositional versions
-of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can be defined from their
+of $\mathbf{0}$, $\mathbf{1}$, $\Pi$, and $\Sigma$. The latter three can be defined from their
 $\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
@@
 \begin{align*}
+  \mathbf{1}_p & :\equiv \| \mathbf{1} \| \\
   \Pi_p\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
-  \Sigma_p\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
-  \mathbf{1}_p & :\equiv \| \mathbf{1} \|
-\end{align*}
+  \Sigma_p\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \|
+\end{align*}
+We assume that we have
+$\mathbf{0}_p:\mProp$ together with $\mathsf{exfalso}_p :\{A:\mType \} \to \mathbf{0}_p \to A$.
 
@@
 
-Types in a context $\GG$ are given by setoid morphisms from $\GG$ to the
-(strict) category of setoids. The definition can be unfolded as follows.
+Types in a context $\GG$ are given by displayed setoids over $\GG$
+with a fibration condition given by $\textsf{coe}$, $\textsf{coh}$.
 
@@
 SeTT's internal universe of propositions, here we state the context equalities
-as elements of the external, metatheoretic universe $\mProp$. Equality on
-dependent types is defined as before in \cite{mpc19}.
+as elements of the external, metatheoretic universe $\mProp$. This makes it easier
+to define models of SeTT as contexts do not need to be partially fibrant. Equality on
+types is defined as before in \cite{mpc19}.
 
@@
 Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
-$\El$ is a setoid morphism, hence in particular it must respect the setoid
+$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
 equality of $\setoidU$.
@@
 
-We can show that $\Uty$ is closed under $\Pi$ types and bool type, including
-$\beta$ and $\eta$ laws. Moreover, all the necessary equations (including
-substitution laws) hold definitionally in the metatheory.
-The universe can be closed under more constructions if more codes are added to
-$\setoidU$.
+We can show that $\Uty$ is closed under $\Pi$ types and booleans such
+that $\EL\,(\mathsf{pi}\,A\,B) \equiv \Pi\,(\EL\,A)\,(\EL\,B)$ and
+$\EL\,\mathsf{bool} = \mathsf{Bool}$. The universe can be closed under
+more constructions if more codes are added to $\setoidU$.
 
@@
   & \textsf{data}\ \ad{in-U}_1 : (A : \mType) \to \ad{in-U}_0\ A \to \mProp_1 \\
-  & \textsf{data}\ \ad{in-U∼}_1 : \{A\ A' : \mType\} \to (R : A \to A' \to \mProp) \to \ad{in-U∼}_0\ R \to \mType_1
+  & \textsf{data}\ \ad{in-U∼}_1 : \{A\ A' : \mType\} \to (R : A \to A' \to \mProp) \to \ad{in-U∼}_0\ R \to \mProp_1
 \end{align*}
@@
 to derive the general recursive-recursive eliminators we need to prove that the
-correponding graph relations are functional, which can't be done without funext.
+corresponding graph relations are functional, which can't be done without funext.
 
 Luckily, it turns out that the \emph{simple} elimination principle is sufficient
-for our purposes. This eliminator can be defined by simple pattern matching on
+for our purposes: all functions of Section \ref{ii-universe} are defined using simple mutual pattern matching without recursion-recursion.
+The simple eliminator can be defined by pattern matching on
 the untyped codes, and does not require extensionality or any extra principles

### 2020-07-06 paper.tex
affiliations

@@
 \and
-  Simon Boulier
+  Simon Boulier \inst{2}
 \and
-  Ambrus Kaposi \inst{2}
+  Ambrus Kaposi \inst{3}\thanks{Supported by the National Research, Development and Innovation Fund of Hungary, financed under the Thematic Excellence Programme funding scheme, Project no.
+ED\_18-1-2019-0030 and the Bolyai Fellowship of the Hungarian Academy of Sciences, Project no.\ BO/00659/19/3.}
 \and
-  Christian Sattler\inst{3}\thanks{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.}
+  Christian Sattler\inst{4}\thanks{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.}
 \and
@@
   \email{\{psztxa,pszcs1,psxfs5\}@nottingham.ac.uk}
+  \and
+  Inria, Nantes, France \\
+  \email{simon.boulier@inria.fr}
   \and

### 2020-07-06 paper.tex
apply lipics style

--- 2020-07-06 paper.tex
@@
-\documentclass{easychair}
+\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2019}
 
@@
 \usepackage{proof}
-\usepackage{tikz}
 \usepackage{bbm}
-\usepackage[bw]{agda}
-\usetikzlibrary{cd}
+%\usepackage[bw]{agda}
+
+\bibliographystyle{plainurl}% the mandatory bibstyle
 
@@
 \newcommand{\setoidU}{\mathcal{U}}
-\newcommand{\ad}[1]{\AgdaFunction{#1}}
+% \newcommand{\ad}[1]{\AgdaFunction{#1}}
+\newcommand{\inU}{\mathsf{inU}}
+\newcommand{\inUU}{\mathsf{inU{\sim}}}
 \newcommand{\Set}{\textsf{Set}}
@@
 
-% \usepackage{amsmath}
-% \usepackage{amssymb}
-% %\usepackage{mathpartir}
-
-%\usepackage{natbib}
-%\setcitestyle{round}
-% Note that the \doi command from the doi package doesn't enable the
-% same kinds of line breaks as the command below.
-\newcommand{\doi}[1]{doi:\href{http://doi.org/#1}{%
-    \urlstyle{same}\nolinkurl{#1}}}
-
 % TODO notes.
-\usepackage[textsize=small]{todonotes}
-\setlength{\marginparwidth}{2cm}
-\newcommand{\txatodo}[1]{\todo[fancyline,color=green!20]{#1}{}}
+%\usepackage[textsize=small]{todonotes}
+%\setlength{\marginparwidth}{2cm}
+%\newcommand{\txatodo}[1]{\todo[fancyline,color=green!20]{#1}{}}
 
 \title{Constructing a universe for the setoid model}
-\titlerunning{Constructing a universe for the setoid model}
-\authorrunning{Altenkirch, Boulier, Kaposi, Sattler, Sestini}
-
-\author{
-  Thorsten Altenkirch \inst{1}\thanks{Supported by USAF grant FA9550-16-1-0029.}
-\and
-  Simon Boulier
-\and
-  Ambrus Kaposi \inst{2}
-\and
-  Christian Sattler\inst{3}\thanks{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.}
-\and
-  Filippo Sestini\inst{1}
-}
-
-\institute{
-  School of Computer Science, University of Nottingham, UK\\
-  \email{\{psztxa,pszcs1,psxfs5\}@nottingham.ac.uk}
-  \and
-  E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary\\
-  \email{akaposi@inf.elte.hu}
-  \and
-  Chalmers University of Technology, Gothenburg, Sweden
-  \email{sattler@chalmers.se}
-}
+%\titlerunning{Constructing a universe for the setoid model}
+
+\author{Thorsten Altenkirch}{School of Computer Science, University of Nottingham, UK}{psztxa@nottingham.ac.uk}{orcid}{Supported by USAF grant FA9550-16-1-0029.}
+\author{Simon Boulier}{Inria, Nantes, France}{simon.boulier@inria.fr}{}{}
+\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{orcid}{Supported by the National Research, Development and Innovation Fund of Hungary, financed under the Thematic Excellence Programme funding scheme, Project no.
+ED\_18-1-2019-0030 and the Bolyai Fellowship of the Hungarian Academy of Sciences, Project no.\ BO/00659/19/3.}
+\author{Christian Sattler}{Chalmers University of Technology, Gothenburg, Sweden}{sattler@chalmers.se}{}{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.}
+\author{Filippo Sestini}{School of Computer Science, University of Nottingham, UK}{Filippo.Sestini@nottingham.ac.uk}{}{}
+
+\authorrunning{T. Altenkirch and S. Boulier and A. Kaposi and C. Sattler and F. Sestini}
+
+\Copyright{Thorsten Altenkirch and Simon Boulier and Ambrus Kaposi and Christian Sattler and Filippo Sestini} %TODO mandatory, please use full first names.
+LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+
+\ccsdesc[500]{Theory of computation~Logic~Type theory} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+
+\keywords{type theory, function extensionality, univalence, setoid model, induction-recursion, induction-induction} %TODO mandatory; please add comma-separated list of keywords
+
+%\category{} %optional, e.g. invited paper
+
+%\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
+%\relatedversion{A full version of the paper is available at \url{...}.}
+
+%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
+
+%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
+Please enter author specific funding statements as fifth argument of the \author macro.
+
+%\acknowledgements{I want to thank \dots}%optional
+
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\EventEditors{John Q. Open and Joan R. Access}
+\EventNoEds{2}
+\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
+\EventShortTitle{CVIT 2016}
+\EventAcronym{CVIT}
+\EventYear{2016}
+\EventDate{December 24--27, 2016}
+\EventLocation{Little Whinging, United Kingdom}
+\EventLogo{}
+\SeriesVolume{42}
+\ArticleNo{23}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \maketitle
+
+\begin{abstract}
+The setoid model of type theory can be used to show the consistency of
+intensional type theory with certain extensionality principles, like function
+extensionality. In addition, Altenkirch's formulation of such model justifies a
+universe of strict (definitionally-proof irrelevant) propositions supporting
+propositional extensionality, a limited form of univalence according to which
+logically equivalent propositions are equal. The appeal of this model
+construction is that it takes place in a small, intensional, type theoretic
+metatheory, therefore giving a method to boostrap extensionality.
+
+Recently, Altenkirch's setoid model has been turned into a formal system, under
+the name of Setoid Type Theory (SeTT). SeTT can be presented as an extension of
+intensional Martin-Loef type theory with syntactic constructs that internalize
+setoid reasoning. These give full access to the extensionality principles that
+hold in the setoid model, while retaining the convenience of an intensional
+theory.
+
+As currently defined, SeTT is already a rich theory. However, its introspection
+capabilities are currently lacking, since it only includes (univalent) universes
+for propositions. We would like to extend SeTT with a universe of setoids
+internalizing the notion of type beyond propositions. This amounts to extending
+the setoid model with a semantics for such a universe.
+
+In this paper, we present the construction of a (non-univalent) universe of
+setoids within the setoid model. The universe is given in steps, first as an
+inductive-recursive definition, which is then translated to an
+inductive-inductive definition, and finally to an inductive family. Every step
+is obtained from the previous by applying known reduction methods, although in a
+novel way. Additionally, our construction shows the setoid model with the
+additional encoding of the universe can still be defined in a relatively small
+metatheory, that however must include a universe of strict propositions, indexed
+W types, and a proof-irrelevant identity type with a strong transport rule. All
+the mathematical developments described in the paper have been formalized in the
+Agda proof-assistant.
+\end{abstract}
 
@@
 
-\paragraph{Contributions}
+\paragraph*{Contributions}
 
@@
 
-\paragraph{Structure of the paper}
+\paragraph*{Structure of the paper}
 
@@
 formers. The core structure of a CwF can be given as the following signature:
-
 \begin{align*}
@@
 irrelevant.
-
 \[
@@
 Types in a context $\GG$ are given by displayed setoids over $\GG$
-with a fibration condition given by $\textsf{coe}$, $\textsf{coh}$.
-
+with a fibration condition given by $\textsf{coe}$, $\textsf{coh}$. In
+the following, we somtimes omit implicit quantifications such as the $\forall\{\gamma_0\ \gamma_1\}$ in the type of $\textsf{sym}\,\GG$.
 \[
@@
     \to A^\sim\ p\ a_0\ a_1 \to A^\sim\ (\textsf{sym}\ \GG\ p)\ a_1\ a_0 \\
-    & \textsf{trans*} : \forall\{\gamma_0\,\gamma_1\,\gamma_2\,a_0\,a_1\,a_2\}\{p_0 : \GG^\sim\ \gamma_0\ \gamma_1\}\{p_1 : \GG^\sim\ \gamma_1\ \gamma_2\}\\
-    & \ \qquad \to A^\sim\ p_0\ a_0\ a_1\to A^\sim\ p_1\ a_1\ a_2\to A^\sim\ (\textsf{trans}\,\GG\,p_0\,p_1)\ a_0\ a_2\\
-    & \textsf{coe} : \forall\{\gamma_0\ \gamma_1 : | \GG |\} \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
-    & \textsf{coh} : \forall\{\gamma_0\ \gamma_1 : | \GG |\} \ (p : \GG^\sim\ \gamma_0\ \gamma_1) (a : | A | \gamma_0)
+    & \textsf{trans*} : A^\sim\ p_0\ a_0\ a_1\to A^\sim\ p_1\ a_1\ a_2\to A^\sim\ (\textsf{trans}\,\GG\,p_0\,p_1)\ a_0\ a_2\\
+    & \textsf{coe} : \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
+    & \textsf{coh} : (p : \GG^\sim\ \gamma_0\ \gamma_1) (a : | A | \gamma_0)
     \to A^\sim\,p\,a\,(\textsf{coe}\,A\,p\,a) \\
@@
 \\
-~
-\\
 \begin{minipage}{0.5\textwidth}
@@
     & | \sigma | : | \GG | \to | \GD | \\
-    & \sigma^\sim : \forall\{\rho_0\, \rho_1\} \to \GG^\sim\, \rho_0\, \rho_1 \\
-    & \quad \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
+    & \sigma^\sim : \GG^\sim\, \rho_0\, \rho_1 \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
   \end{aligned}
@@
     & | t | : (\gamma : | \GG |) \to | A | \, \gamma \\
-    & t^\sim : \forall\{\gamma_0\, \gamma_1\}
-    (p : \GG^\sim\ \gamma_0\ \gamma_1) \\
-    & \quad \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
+    & t^\sim : (p : \GG^\sim\ \gamma_0\ \gamma_1) \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
   \end{aligned}
@@
 \\
-~
-\\
+
 We can show that the setoid model validates the usual basic type formers ($\Pi,
-\Sigma$, etc.), including function extensionality and a universe of strict
-propositions satisfying propositional extensionality \cite{setoid99}. Note that
+\Sigma$, etc.), function extensionality and a universe of strict
+propositions ith propositional extensionality \cite{setoid99}. Note that
 we don't need identity types or inductive types (W-types) for this.
@@
 We have a universe of propositions $\Prop$ defined as follows:
-
 \[
@@
 \infer{
-  u \equiv v
+  u \equiv v\vphantom{\Prop}
 }{
@@
 equality.
-
 \[
@@
 recall reflexivity:
-
 \[
@@
 SeTT respects setoid equality, so that we can transport along any such equality:
-
     \[
@@
 with transport implemented via coercion.
-
 \[
@@
 As an example, consider the following definition of a small universe with
-booleans and $\Pi$ types.
+bool and $\Pi$.
 
@@
 
-In our case, we can encode $\U$ as an inductive type $\ad{in-U}$ that
+In our case, we can encode $\U$ as an inductive type $\inU$ that
 \emph{carves out} all types in $\mType$ that are in the image of $\El$. In other
-words, $\ad{in-U}$ is a predicate that holds for any type that would have been
+words, $\inU$ is a predicate that holds for any type that would have been
 obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed by
-the type of codes, the definition of $\ad{in-U}$ quite expectedly reflects the
+the type of codes, the definition of $\inU$ quite expectedly reflects the
 inductive structure of codes.
 %
-\begin{align*}
-  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
-  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
-  & \quad \textsf{inPi}
-  :  \{A : \mType\} \{B : A \to \mType\}\\
-  & \ \qquad \to \ad{in-U}\ A
-  \to ((a : A) \to \ad{in-U}\ (B\ a))
-  \to \ad{in-U}\ ((a : A) \to (B\ a))
-\end{align*}
-
-$\U$ and $\El$ can be given as follows:
-\[
-\U = \Sigma (A : \mType)\ (\textsf{in-U}\ A) \qquad \El = \pi_1
-\]
+\begin{alignat*}{10}
+  & \textsf{data}\ \inU && : \mType \to \mType_1 \\
+  & \quad \textsf{inBool} && : \textsf{in-U}\ \mathbbm{2} \\
+  & \quad \textsf{inPi} && :  \inU\ A \to ((a : A) \to \inU\ (B\ a)) \to \inU\ ((a : A) \to (B\ a))
+\end{alignat*}
+
+$\U$ and $\El$ can be given by $\U :\equiv \Sigma (A : \mType)\ (\textsf{in-U}\ A)$ and $\El :\equiv \pi_1$.
 
@@
 relation reflecting equality of codes, in addition to proofs that these are
-indeed equivalence relations:
-%
-
+indeed equivalence relations: \\
+%
 \begin{minipage}{0.5\textwidth}
@@
 \end{minipage}
-\\
-~
 \\
@@
 The whole construction is quite long, below we only show the more interesting
-definitions of $\setoidU$ and $\El$:
-
+definitions of $\setoidU$ and $\El$: \\
 \begin{minipage}{0.5\textwidth}
@@
 \end{minipage}
+\\
 
@@
 We can now give a full definition of the setoid universe, and of $\EL\ A$ for
-any $A : \Tm\ \GG\ \Uty$:
-%
-
+any $A : \Tm\ \GG\ \Uty$: \\
+%
 \begin{minipage}{0.5\textwidth}
@@
 \\
-~
-\\
 
@@
 of induction-recursion like the one described in \cite{dybjer-setzer}. In any
-case, we would like to avoid extending the metatheory with any forms of
-induction-recursion, given our goal to keep the metatheory as small and
+case, we would like to avoid extending the metatheory with any form of
+induction-recursion in order to keep the metatheory as small and
 essential as possible.
@@
 We will follow the method outlined in Section~\ref{ir-universes}. In
-addition to $\ad{in-U}$ for defining $\U$, we also introduce a family
-$\ad{in-U∼}$ of binary relations between types in the universe, from which we
+addition to $\inU$ for defining $\U$, we also introduce a family
+$\inUU$ of binary relations between types in the universe, from which we
 then define $\_\sim_{\setoidU}\_$.
@@
 \begin{align*}
-  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
-  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \mType\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \mProp) \to \mType_1
-\end{align*}
-
-Note that $\ad{in-U}$ appears in the sort of $\ad{in-U∼}$. Since these types are
+  & \textsf{data}\ \inU : \mType \to \mType_1 \\
+  & \textsf{data}\ \inUU : \{A\ A' : \mType\} \to \inU\ A \to \inU\ A' \to (A \to A' \to \mProp) \to \mType_1
+\end{align*}
+
+Note that $\inU$ appears in the sort of $\inUU$. Since these types are
 mutually defined, they form an instance of \emph{induction-induction}, a schema
@@
 $\setoidU$ as a $\Sigma$ type, and $\El$ as the corresponding first projection.
-
-\begin{minipage}{0.4\textwidth}
-\begin{align*}
-  & \setoidU : \mType_1 \\
-  & \setoidU = \Sigma\ (X : \mType)\ (\ad{in-U}\ X)
-\end{align*}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}
-\begin{align*}
-  & \El : \setoidU \to \mType \\
-  & \El = \pi_1
-\end{align*}
-\end{minipage}
-\\
-~
-\\
+\begin{alignat*}{10}
+  & \setoidU && : \mType_1 && \El : \setoidU \to \mType \\
+  & \setoidU && :\equiv \Sigma\ (X : \mType)\ (\inU\ X) \hspace{5em} && \El :\equiv \pi_1
+\end{alignat*}
 
@@
 that is, if $A \sim_{\setoidU} B$. Existence and well-formedness of such
-relations is expressed via the type $\ad{in-U~}$ just defined, hence we would
+relations is expressed via the type $\inUU$ just defined, hence we would
 expect $A \sim_{\setoidU} B$ to be defined as follows:
-
 \[
-(A, p_A) \sim_{\setoidU} (B, p_B) :\equiv \Sigma (R: A \to B \to \mProp)(\ad{in-U∼}\ p_A\ p_B\ R)
+(A, p_A) \sim_{\setoidU} (B, p_B) :\equiv \Sigma (R: A \to B \to \mProp)(\inUU\ p_A\ p_B\ R)
 \]
@@
   & (X , p) \sim_{\setoidU} (X' , p') =
-   \ \parallel \Sigma\ (R : X \to X' \to \mProp)\ (\ad{in-U∼}\ p\ p'\ R) \parallel
+   \ \parallel \Sigma\ (R : X \to X' \to \mProp)\ (\inUU\ p\ p'\ R) \parallel
 \end{align*}
@@
 up being logically equivalent to the proposition we would have obtained by
-$\ad{proj}_1\ p\ x\ y$ if there were no truncation.
+$\mathsf{proj}_1\ p\ x\ y$ if there were no truncation.
 %
 More precisely, we need to construct proofs that for any concrete $R$ and
-$\ad{inR}$, the types $| (R, \ad{inR}) | \vdash x \sim_{\El} y$ and $R\ x\ y$
+$\mathsf{inR}$, the types $| (R, \mathsf{inR}) | \vdash x \sim_{\El} y$ and $R\ x\ y$
 are logically equivalent. These in turn need to be defined mutually with
@@
 $\Pi$ types.
-
-\begin{align*}
-  & \bullet : \textsf{Con} \\
-  & \_,\_ : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \\
-  & \iota : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
-  & \Pi : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
-\end{align*}
+\begin{alignat*}{10}
+  & \bullet && : \textsf{Con}                                                          &&  \iota && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
+  & \_,\_ && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{5em} &&  \Pi && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
+\end{alignat*}
 
@@
 $\textsf{Con}_0$ and $\textsf{Ty}_0$ are defined mutually, their sorts are not
-indexed.
-
+indexed. \\
 \begin{minipage}{0.4\textwidth}
-\begin{align*}
-  & \bullet_0 : \textsf{Con}_0 \\
-  & \_,_0\_ : (\GG : \textsf{Con}_0) \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
-  & \iota_0 : (\GG : \textsf{Con}_0) \to \textsf{Ty}_0 \\
-  & \Pi_0 : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \\
-\end{align*}
+\begin{alignat*}{10}
+  & \bullet_0 && : \textsf{Con}_0 \\
+  & \_,_0\_ && : (\GG : \textsf{Con}_0) \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
+  & \iota_0 && : (\GG : \textsf{Con}_0) \to \textsf{Ty}_0 \\
+  & \Pi_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \hspace{2.8em} \\
+\end{alignat*}
 \end{minipage}
 \begin{minipage}{0.5\textwidth}
-\begin{align*}
-  & \bullet_1 : \textsf{Con}_1\ \bullet_0 \\
-  & \_,_1\_ : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
-  & \qquad \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
-  & \iota_1 : \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
-  & \Pi_1 : \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \\
-  & \qquad \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \\
-  & \qquad \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
-\end{align*}
+\vspace{-1.3em}
+\begin{alignat*}{10}
+  & \bullet_1 && : \textsf{Con}_1\ \bullet_0 \\
+  & \_,_1\_ && : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
+  & && \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
+  & \iota_1 && : \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
+  & \Pi_1 && : \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \\
+  & && \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \\
+  & && \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
+\end{alignat*}
 \end{minipage}
-\\
-~
 \\
@@
 is straighforward:
-
-\begin{align*}
-  & \bullet :\equiv (\bullet_0 , \bullet_1) \\
-  & (\GG_0 , \GG_1) , (A_0 , A_1) :\equiv ((\GG_0 ,_0 A_0) , (\GG_1 ,_1 A_1)) \\
-  & \iota\ (\GG_0 , \GG_1) :\equiv (\iota_0 \GG_0 , \iota_1\ \GG_1) \\
-  & \Pi\ \{\GG_0 , \GG_1\}(A_0 , A_1) (B_0 , B_1) :\equiv
-    ((\Pi_0\ \GG_0\ A_0\ B_0) , (\Pi_1\ \GG_1\ A_1\ B_1))
-\end{align*}
+\begin{alignat*}{10}
+  & \bullet && :\equiv (\bullet_0 , \bullet_1) \\
+  & (\GG_0 , \GG_1) , (A_0 , A_1) && :\equiv ((\GG_0 \mathbin{,_0} A_0) , (\GG_1 ,_1 A_1)) \\
+  & \iota\ (\GG_0 , \GG_1) && :\equiv (\iota_0\ \GG_0 , \iota_1\ \GG_1) \\
+  & \Pi\ \{\GG_0 , \GG_1\}(A_0 , A_1) (B_0 , B_1) && :\equiv
+    (\Pi_0\ \GG_0\ A_0\ B_0 , \Pi_1\ \GG_1\ A_1\ B_1)
+\end{alignat*}
 
@@
 \mType$, the general eliminators have the following signatures:
-
-\begin{align*}
-  & \textsf{elim}_{\textsf{Con}} : (\GG : \textsf{Con}) \to C\ \GG \\
-  & \textsf{elim}_{\textsf{Ty}} : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG)
-  \to T\ \GG\ A\ (\textsf{elim}_{\textsf{Con}}\ \GG)
+\begin{align*}
+  & \textsf{elim}_{\textsf{Con}} : (\GG : \textsf{Con}) \to C\ \GG \hspace{5em} & \textsf{elim}_{\textsf{Ty}} : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to T\ \GG\ A\ (\textsf{elim}_{\textsf{Con}}\ \GG)
 \end{align*}
@@
 the graph of the eliminators in the form of inductively-generated relations:
-
 \begin{align*}
@@
 soon as we try to define the eliminator. One issue is that the definition of the
-eliminator relies on a proof that the well-typing predicates $\ad{in-U}_1,
-\ad{in-U∼}_1$ are propositional, that is, any two of their elements are equal.
+eliminator relies on a proof that the well-typing predicates $\inU_1,
+\inUU_1$ are propositional, that is, any two of their elements are equal.
 %
@@
 $\mProp$-valued families, rather than in $\mType$:
-
-\begin{align*}
-  & \textsf{data}\ \ad{in-U}_0 : \mType \to \mType_1 \\
-  & \textsf{data}\ \ad{in-U∼}_0 : \{A\ A' : \mType\} \to (A \to A' \to \mProp) \to \mType_1 \\
-  & \textsf{data}\ \ad{in-U}_1 : (A : \mType) \to \ad{in-U}_0\ A \to \mProp_1 \\
-  & \textsf{data}\ \ad{in-U∼}_1 : \{A\ A' : \mType\} \to (R : A \to A' \to \mProp) \to \ad{in-U∼}_0\ R \to \mProp_1
-\end{align*}
+\begin{alignat*}{10}
+  & \textsf{data}\ \inU_0 && : \mType \to \mType_1 \\
+  & \textsf{data}\ \inUU_0 && : \{A\ A' : \mType\} \to (A \to A' \to \mProp) \to \mType_1 \\
+  & \textsf{data}\ \inU_1 && : (A : \mType) \to \inU_0\ A \to \mProp_1 \\
+  & \textsf{data}\ \inUU_1 && : \{A\ A' : \mType\} \to (R : A \to A' \to \mProp) \to \inUU_0\ R \to \mProp_1
+\end{alignat*}
 
@@
 %
-However, it introduces a different issue: $\ad{in-U}_1$ and $\ad{in-U∼}_1$ give
+However, it introduces a different issue: $\inU_1$ and $\inUU_1$ give
 rise to equational constraints on their indices, in the form of proofs of the
-$\mProp$-valued identity type. The definition of the eliminators for $\ad{in-U}$
-and $\ad{in-U∼}$ relies on the ability to transport along these proofs, hence
+$\mProp$-valued identity type. The definition of the eliminators for $\inU$
+and $\inUU$ relies on the ability to transport along these proofs, hence
 the need to extend our metatheory with a primitive, strong form of transport for
@@
 However, we would still need extensionality to derive the general eliminators
-for $\ad{in-U}$ and $\ad{in-U∼}$. In fact, as explained in the previous section,
+for $\inU$ and $\inUU$. In fact, as explained in the previous section,
 to derive the general recursive-recursive eliminators we need to prove that the
@@
 
-\bibliographystyle{alpha}
 \bibliography{bibliography}

### 2020-07-07 paper.tex
orcid

@@
 
-\author{Thorsten Altenkirch}{School of Computer Science, University of Nottingham, UK}{psztxa@nottingham.ac.uk}{orcid}{Supported by USAF grant FA9550-16-1-0029.}
+\author{Thorsten Altenkirch}{School of Computer Science, University of Nottingham, UK}{psztxa@nottingham.ac.uk}{https://orcid.org/0000-0002-6582-5025}{Supported by USAF grant FA9550-16-1-0029.}
 \author{Simon Boulier}{Inria, Nantes, France}{simon.boulier@inria.fr}{}{}
-\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{orcid}{Supported by the National Research, Development and Innovation Fund of Hungary, financed under the Thematic Excellence Programme funding scheme, Project no.
+\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{Supported by the National Research, Development and Innovation Fund of Hungary, financed under the Thematic Excellence Programme funding scheme, Project no.
 ED\_18-1-2019-0030 and the Bolyai Fellowship of the Hungarian Academy of Sciences, Project no.\ BO/00659/19/3.}

### 2020-07-08 paper.tex
minor

@@
 
-Induction-recursion is arguably a nicer and more natural way to define internal
+Induction-recursion is arguably a nice and natural way to define internal
 universes in type theory, however it is not always strictly required.
@@
 
-$\U$ and $\El$ can be given by $\U :\equiv \Sigma (A : \mType)\ (\textsf{in-U}\ A)$ and $\El :\equiv \pi_1$.
+$\U$ and $\El$ can be given by $\U :\equiv \Sigma\ (A : \mType)\ (\textsf{in-U}\ A)$ and $\El :\equiv \pi_1$.
 
@@
 \begin{align*}
-  & \textsf{data}\ \setoidU : \Set_1 \\
+  & \textsf{data}\ \setoidU : \mType_1 \\
   & \quad \textsf{bool} : \setoidU \\
@@
 \[
-(A, p_A) \sim_{\setoidU} (B, p_B) :\equiv \Sigma (R: A \to B \to \mProp)(\inUU\ p_A\ p_B\ R)
+(A, p_A) \sim_{\setoidU} (B, p_B) :\equiv \Sigma\ (R: A \to B \to \mProp)\ (\inUU\ p_A\ p_B\ R)
 \]
@@
 like $\reflu, \reflel$, etc., which can be adapted from their IR counterparts
-more or less straighforwardly. The final result does not use
+more or less straightforwardly. The final result does not use
 induction-recursion, but it's nevertheless an instance of infinitary
@@
 We can recover the original inductive-inductive type as $\textsf{Con} :\equiv
-\Sigma (\GG_0 : \textsf{Con}_0)(\textsf{Con}_1\ \GG_0)$ and $\textsf{Ty}\ \GG
-:\equiv \Sigma(A_0 :
-\textsf{Ty}_0)(\textsf{Ty}_1\ (\pi_1\ \GG)\ A_0)$. Recovering the constructors
-is straighforward:
+\Sigma\ (\GG_0 : \textsf{Con}_0)\ (\textsf{Con}_1\ \GG_0)$ and $\textsf{Ty}\ \GG
+:\equiv \Sigma\ (A_0 :
+\textsf{Ty}_0)\ (\textsf{Ty}_1\ (\pi_1\ \GG)\ A_0)$. Recovering the constructors
+is straightforward:
 \begin{alignat*}{10}

### 2020-07-08 paper.tex
Minor change to the conclusions

@@
 
-As further work, we would like to investigate a construction of the setoid
-universe that does not require a strong transport rule for the proof-irrelevant
-identity type. Not only to avoid unnecessary extensions to the metatheory, but
-also because we don't yet know whether $\mlttp$ extended with this rule is
-normalizing.
-%
-Indeed, previous work in this respect \cite{abel2019failure} seems to suggest
-that this rule is a non-trivial addition to the metatheory.
+Our inductive definition of the setoid universe relies on a metatheory with a
+strong transport rule. As future work, we would like to prove normalization for
+this metatheory, since we don't yet know if this property holds in the presence
+of such rule, and previous work in this respect \cite{abel2019failure} seems to
+suggest that is represents a non-trivial addition.
+%
+An alternative path worth investigating is to find a construction of the
+universe that does not rely on strong transport. Note that this issue only
+involves the inductive definition, as the inductive-recursive and
+inductive-inductive versions do not need strong transport.
 

### 2020-07-08 paper.tex
Compress concls to stay within 15 pgs

@@
 
-Additionally, we want to figure out whether our inductive encoding of the setoid
-universe can support universe induction. A related question is whether the
-alternative definitions using induction-recursion and induction-induction can
-support it.
-
 Another question regards the relationship between SeTT \cite{mpc19} and XTT
@@
 similar design choices, like definitional proof-irrelevance. We would like to
-know whether their respective notions of models are equivalent.
-In particular, we would like to obtain an XTT model from a SeTT model, and vice versa. Since XTT
-universes support universe induction, for one direction we would need to
-extend our own universe with the same principle (see discussion in
-Section~\ref{sec:setoid-model} and the previous paragraph). A further question is
-whether this mapping of models is functorial.
-
-The groupoid model of type theory \cite{groupoid} can be seen as a
-generalization of the setoid model, since groupoids are essentially
-proof-relevant setoids.
-%
-One question is therefore whether it is possible to design a type theory
-internalizing the groupoid model, in the same way that SeTT represents a syntax
-for the setoid model. A further question is whether such ``groupoid type
-theory'' can be justified, similarly to SeTT, via a syntactic translation,
-perhaps with SeTT itself as the target theory.
+know whether their respective notions of models are equivalent, that is, if we
+can obtain an XTT model from a SeTT model, and vice versa. Since XTT universes
+support universe induction, for one direction we would need to extend our own
+universe with the same principle (see discussion in
+Section~\ref{sec:setoid-model} and the previous paragraph). Thus a related
+question is whether our encodings of the setoid universe can support universe
+induction. A further question is whether this mapping of models is functorial.
+
+Groupoids can be regarded as generalized setoids. In the future we would like to
+design a type theory internalizing the groupoid model of type theory
+\cite{groupoid}, in the same way that SeTT represents a syntax for the setoid
+model. A further question is whether such ``groupoid type theory'' can be
+justified, similarly to SeTT, via a syntactic translation, perhaps with SeTT
+itself as the target theory.
 

### 2020-07-08 paper.tex
Add paragraphs on missing formalization

@@
 
-Luckily, it turns out that the \emph{simple} elimination principle is sufficient
-for our purposes: all functions of Section \ref{ii-universe} are defined using simple mutual pattern matching without recursion-recursion.
-The simple eliminator can be defined by pattern matching on
-the untyped codes, and does not require extensionality or any extra principles
-beyond strong transport.
+Luckily, it would seem the \emph{simple} elimination principle is sufficient for
+our purposes: careful inspection of the code suggests that all functions
+described in Section \ref{ii-universe} can be defined using simple mutual
+pattern matching without recursion-recursion. The simple eliminator can be
+defined by pattern matching on the untyped codes, and does not require
+extensionality or any extra principles beyond strong transport. Note that
+producing a full formalization that uses exclusively simple eliminators appears
+to be extremely time-consuming; we therefore leave this task to future work.
 
@@
 data type transformation methods in a novel way that accounts for the
-peculiarities of our construction. The mathematical contents of this paper have
-been formalized in the proof-assistant Agda (see \cite{agda-code}).
-
-Our inductive definition of the setoid universe relies on a metatheory with a
-strong transport rule. As future work, we would like to prove normalization for
-this metatheory, since we don't yet know if this property holds in the presence
-of such rule, and previous work in this respect \cite{abel2019failure} seems to
+peculiarities of our construction.
+
+Most of the mathematical contents of this paper have been formalized in the
+proof-assistant Agda (see \cite{agda-code}). We aim to complete the
+formalization in the near future, in particular to obtain a machine-checked
+proof that our inductive construction of the setoid universe does not rely on
+recursion-recursion.
+
+The inductive definition of the universe relies on a metatheory with a strong
+transport rule. As future work, we would like to prove normalization for this
+metatheory, since we don't yet know if this property holds in the presence of
+such rule, and previous work in this respect \cite{abel2019failure} seems to
 suggest that is represents a non-trivial addition.

### 2020-07-08 paper.tex
small changes

@@
 
-We also require indexed W-types, both in $\mType$ and $\mProp$; hence we have $W
-: (S : \mType) \to (S \to \mType) \to \mType$ and $W_p : (S : \mType) \to (S \to
-\mType) \to \mProp$. The elimination principle of $W_p$ only allows defining functions into $\mProp$s.
-%
-From $W_p$ we can define propositional truncation $\|\_\| : \mType \to \mProp$, with constructor
+We also require indexed W-types, both in $\mType$ and $\mProp$; hence we have $W_\Box
+: (S : I\to\mType) \to ((i:I)\to S\ i\to I \to \mType) \to I\to \Box$ where $\Box \in \{\mType, \mProp\}$.
+The elimination principle of $W_\mProp$ only allows defining functions into $\mProp$s.
+%
+From $W_\mProp$ we can define propositional truncation $\|\_\| : \mType \to \mProp$, with constructor
   $|\_| : \{A : \mType \} \to A \to \| A \|$ and eliminator
@@
 \begin{align*}
-  \mathbf{1}_p & :\equiv \| \mathbf{1} \| \\
-  \Pi_p\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
-  \Sigma_p\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \|
+  \mathbf{1}_\mProp & :\equiv \| \mathbf{1} \| \\
+  \Pi_\mProp\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
+  \Sigma_\mProp\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \|
 \end{align*}
 We assume that we have
-$\mathbf{0}_p:\mProp$ together with $\mathsf{exfalso}_p :\{A:\mType \} \to \mathbf{0}_p \to A$.
+$\mathbf{0}_\mProp:\mProp$ together with $\mathsf{exfalso}_\mProp :\{A:\mType \} \to \mathbf{0}_\mProp \to A$.
 
@@
 %
-Note that this identity type must necessarily be assumed as primitive; in fact,
-even though we could have defined \textsf{Id} from $W_p$, \textsf{transp} is not
-definable from $W_p$ elimination.
+% Note that this identity type must necessarily be assumed as
+% primitive; in fact, even though we could have defined \textsf{Id}
+% from $W_\mProp$, \textsf{transp} is not definable from $W_\mProp$
+% elimination.
 
@@
 
-The inductive definition of the universe relies on a metatheory with a strong
-transport rule. As future work, we would like to prove normalization for this
-metatheory, since we don't yet know if this property holds in the presence of
-such rule, and previous work in this respect \cite{abel2019failure} seems to
-suggest that is represents a non-trivial addition.
-%
-An alternative path worth investigating is to find a construction of the
-universe that does not rely on strong transport. Note that this issue only
-involves the inductive definition, as the inductive-recursive and
-inductive-inductive versions do not need strong transport.
+In contrast to the inductive-recursive and inductive-inductive
+versions of the universe, the inductive definition relies on a
+metatheory with a strong transport rule. As future work, we would like
+to prove normalization for this metatheory, since we don't yet know if
+this property holds in the presence of such rule, and previous work in
+this respect \cite{abel2019failure} seems to suggest that is
+represents a non-trivial addition.
 

### 2020-07-08 paper.tex
last minute updates

@@
 inductive identity type. The main difference is that our identity type is a
-$\mProp$-valued relation. We also include a transport combinator
-\textsf{transp}.
-%
-\begin{align*}
-  & \textsf{Id} : \{A : \mType \} \to A \to A \to \mProp \\
-  & \textsf{refl} : \{A : \mType \} (a : A) \to \textsf{Id}\ a\ a \\
-  & \textsf{transp} : \{A : \mType \} (C : A \to \mType) \{a_0\ a_1 : A\} \to \textsf{Id}\ a_0\ a_1 \to C\ a_0 \to C\ a_1
-\end{align*}
+$\mProp$-valued relation. We have a transport combinator
+\textsf{transp} from which \textsf{J} is derivable.
+%
+\begin{alignat*}{10}
+  & \textsf{Id} && :\,\, && \{A : \mType \} \to A \to A \to \mProp \\
+  & \textsf{refl} && : && \{A : \mType \} (a : A) \to \textsf{Id}\ a\ a \\
+  & \textsf{transp} && : && \{A : \mType \} (C : A \to \mType) \{a_0\ a_1 : A\} \to \textsf{Id}\ a_0\ a_1 \to C\ a_0 \to C\ a_1 \\
+  & && && \textsf{transp}\ C\ \{x\}\ \{x\}\ e\ u \equiv u
+\end{alignat*}
 
@@
 %
-The reason to have an identity type in $\mProp$ with such a strong transport
-rule are made clear in Section~\ref{inductive-universe}.
+We only use this identity type with such a strong transport in Section~\ref{inductive-universe}.
 %

### 2020-07-08 paper.tex
ERC support + compression

@@
 \author{Thorsten Altenkirch}{School of Computer Science, University of Nottingham, UK}{psztxa@nottingham.ac.uk}{https://orcid.org/0000-0002-6582-5025}{Supported by USAF grant FA9550-16-1-0029.}
-\author{Simon Boulier}{Inria, Nantes, France}{simon.boulier@inria.fr}{}{}
+\author{Simon Boulier}{Inria, Nantes, France}{simon.boulier@inria.fr}{}{Supported by ERC Starting Grant CoqHoTT 637339.}
 \author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{Supported by the National Research, Development and Innovation Fund of Hungary, financed under the Thematic Excellence Programme funding scheme, Project no.
@@
 peculiarities of our construction.
-
+%
 Most of the mathematical contents of this paper have been formalized in the
@@
 metatheory with a strong transport rule. As future work, we would like
-to prove normalization for this metatheory, since we don't yet know if
-this property holds in the presence of such rule, and previous work in
+to prove normalization for this metatheory since previous work in
 this respect \cite{abel2019failure} seems to suggest that is

### 2020-10-12 paper.tex
add Palmgren
Adapt paper and abstract to LNCS format

@@
-\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2019}
-
+%% \documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{llncs}
+\documentclass[autoref]{llncs}
+
+\usepackage{hyperref}
 \usepackage{amsmath}
@@
 
+\begin{document}
+
 \title{Constructing a universe for the setoid model}
@@
 
-\author{Thorsten Altenkirch}{School of Computer Science, University of Nottingham, UK}{psztxa@nottingham.ac.uk}{https://orcid.org/0000-0002-6582-5025}{Supported by USAF grant FA9550-16-1-0029.}
-\author{Simon Boulier}{Inria, Nantes, France}{simon.boulier@inria.fr}{}{Supported by ERC Starting Grant CoqHoTT 637339.}
-\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{Supported by the National Research, Development and Innovation Fund of Hungary, financed under the Thematic Excellence Programme funding scheme, Project no.
+%% \author{Thorsten Altenkirch}{School of Computer Science, University of Nottingham, UK}{psztxa@nottingham.ac.uk}{https://orcid.org/0000-0002-6582-5025}{Supported by USAF grant FA9550-16-1-0029.}
+%% \author{Simon Boulier}{Inria, Nantes, France}{simon.boulier@inria.fr}{}{Supported by ERC Starting Grant CoqHoTT 637339.}
+%% \author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{Supported by the National Research, Development and Innovation Fund of Hungary, financed under the Thematic Excellence Programme funding scheme, Project no.
 ED\_18-1-2019-0030 and the Bolyai Fellowship of the Hungarian Academy of Sciences, Project no.\ BO/00659/19/3.}
-\author{Christian Sattler}{Chalmers University of Technology, Gothenburg, Sweden}{sattler@chalmers.se}{}{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.}
-\author{Filippo Sestini}{School of Computer Science, University of Nottingham, UK}{Filippo.Sestini@nottingham.ac.uk}{}{}
+%% \author{Christian Sattler}{Chalmers University of Technology, Gothenburg, Sweden}{sattler@chalmers.se}{}{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.}
+%% \author{Filippo Sestini}{School of Computer Science, University of Nottingham, UK}{Filippo.Sestini@nottingham.ac.uk}{}{}
+
+\author{Thorsten Altenkirch\inst{1}
+  \thanks{Supported by USAF grant FA9550-16-1-0029.}
+  \orcidID{0000-0002-6582-5025} \and
+  Simon Boulier\inst{2}\thanks{Supported by ERC Starting Grant CoqHoTT 637339.} \and
+  Ambrus Kaposi\inst{3}\orcidID{0000-0001-9897-8936}
+  \thanks{Supported by the National Research, Development and Innovation Fund of Hungary, financed under the Thematic Excellence Programme funding scheme, Project no.
+ED\_18-1-2019-0030 and the Bolyai Fellowship of the Hungarian Academy of Sciences, Project no.\ BO/00659/19/3.}\and
+  Christian Sattler\inst{4}\thanks{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.} \and
+  Filippo Sestini\inst{1}\orcidID{0000-0002-8701-5613}}
+
+\institute{School of Computer Science, University of Nottingham, UK
+  \email{\{psztxa,psxfs5\}@nottingham.ac.uk} \and
+  Inria, Nantes, France\email{simon.boulier@inria.fr} \and
+  E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary
+  \email{akaposi@inf.elte.hu} \and
+  Chalmers University of Technology, Gothenburg, Sweden
+  \email{sattler@chalmers.se}
+}
 
@@
 
-\Copyright{Thorsten Altenkirch and Simon Boulier and Ambrus Kaposi and Christian Sattler and Filippo Sestini} %TODO mandatory, please use full first names.
+\maketitle              % typeset the header of the contribution
+
+%% \Copyright{Thorsten Altenkirch and Simon Boulier and Ambrus Kaposi and Christian Sattler and Filippo Sestini} %TODO mandatory, please use full first names.
 LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
-\ccsdesc[500]{Theory of computation~Logic~Type theory} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
-
-\keywords{type theory, function extensionality, univalence, setoid model, induction-recursion, induction-induction} %TODO mandatory; please add comma-separated list of keywords
-
-%\category{} %optional, e.g. invited paper
-
-%\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
-%\relatedversion{A full version of the paper is available at \url{...}.}
-
-%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
-
-%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-%\acknowledgements{I want to thank \dots}%optional
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-\maketitle
+%% \ccsdesc[500]{Theory of computation~Logic~Type theory} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
 
 \begin{abstract}
-The setoid model of type theory can be used to show the consistency of
-intensional type theory with certain extensionality principles, like function
-extensionality. In addition, Altenkirch's formulation of such model justifies a
-universe of strict (definitionally-proof irrelevant) propositions supporting
-propositional extensionality, a limited form of univalence according to which
-logically equivalent propositions are equal. The appeal of this model
-construction is that it takes place in a small, intensional, type theoretic
-metatheory, therefore giving a method to boostrap extensionality.
-
-Recently, Altenkirch's setoid model has been turned into a formal system, under
-the name of Setoid Type Theory (SeTT). SeTT can be presented as an extension of
-intensional Martin-Loef type theory with syntactic constructs that internalize
-setoid reasoning. These give full access to the extensionality principles that
-hold in the setoid model, while retaining the convenience of an intensional
-theory.
-
-As currently defined, SeTT is already a rich theory. However, its introspection
-capabilities are currently lacking, since it only includes (univalent) universes
-for propositions. We would like to extend SeTT with a universe of setoids
-internalizing the notion of type beyond propositions. This amounts to extending
-the setoid model with a semantics for such a universe.
-
-In this paper, we present the construction of a (non-univalent) universe of
-setoids within the setoid model. The universe is given in steps, first as an
-inductive-recursive definition, which is then translated to an
-inductive-inductive definition, and finally to an inductive family. Every step
-is obtained from the previous by applying known reduction methods, although in a
-novel way. Additionally, our construction shows the setoid model with the
-additional encoding of the universe can still be defined in a relatively small
-metatheory, that however must include a universe of strict propositions, indexed
-W types, and a proof-irrelevant identity type with a strong transport rule. All
-the mathematical developments described in the paper have been formalized in the
-Agda proof-assistant.
+The setoid model is a model of intensional type theory that validates certain
+extensionality principles, like function extensionality and propositional
+extensionality, the latter being a limited form of univalence that equates
+logically equivalent propositions. The appeal of this model construction is that
+it can be constructed in a small, intensional, type theoretic metatheory,
+therefore giving a method to boostrap extensionality.
+%
+The setoid model has been recently adapted into a formal system, namely Setoid
+Type Theory (SeTT). SeTT is an extension of intensional Martin-L\"of type theory
+with constructs that give full access to the extensionality principles that hold
+in the setoid model.
+
+Although already a rich theory as currently defined, SeTT currently lacks a way
+to internalize the notion of type beyond propositions. This paper is a step
+towards extending SeTT with a universe of setoids.
+%
+To this aim, we present the construction of a (non-univalent) universe of
+setoids within the setoid model, first as an inductive-recursive definition,
+which is then translated to an inductive-inductive definition and finally to an
+inductive family. These translations from more powerful definition schemas to
+simpler ones ensure that our construction can still be defined in a relatively
+small metatheory, that however must include a proof-irrelevant identity type
+with a strong transport rule. \keywords{type theory \and
+  function extensionality \and univalence \and setoid model \and
+  induction-recursion \and induction-induction}
 \end{abstract}
@@
 it is left open whether the construction presented here supports this principle.
+%
+Palmgren and Wilander \cite{2014arXiv1408.1364P} construct a setoid
+universe using a translation into constructive set theory. In
+contrast, we target a basic intensional type theory and hence also
+give a computational semantics.
 
@@
 \mProp$:
-%
+\vspace{-0.2em}
 \begin{align*}
@@
 \textsf{transp} from which \textsf{J} is derivable.
-%
+\vspace{-0.2em}
 \begin{alignat*}{10}
@@
 formers. The core structure of a CwF can be given as the following signature:
+\vspace{-0.2em}
 \begin{align*}
@@
 bool and $\Pi$.
-
+\vspace{-0.2em}
 \begin{minipage}{0.45\textwidth}
@@
 inductive structure of codes.
-%
+\vspace{-0.2em}
 \begin{alignat*}{10}
@@
 indeed equivalence relations: \\
-%
 \begin{minipage}{0.5\textwidth}
+\vspace{-0.2em}
 \begin{align*}
@@
 \begin{minipage}{0.5\textwidth}
+\vspace{-0.2em}
 \begin{align*}
@@
 relation.
-%
+\vspace{-0.2em}
 \begin{align*}
@@
 definitions of $\setoidU$ and $\El$: \\
+\vspace{-0.2em}
 \begin{minipage}{0.5\textwidth}
@@
 any $A : \Tm\ \GG\ \Uty$: \\
-%
+\vspace{-0.2em}
 \begin{minipage}{0.5\textwidth}
@@
 then define $\_\sim_{\setoidU}\_$.
-%
+\vspace{-0.2em}
 \begin{align*}
@@
 $\setoidU$ as a $\Sigma$ type, and $\El$ as the corresponding first projection.
+\vspace{-0.2em}
 \begin{alignat*}{10}
@@
 the $\Sigma$ type above to force it to be in $\mProp_1$.
-%
+\vspace{-0.2em}
 \begin{align*}
@@
 We are now left to define the indexed equivalence relation on $\El$:
-%
+\vspace{-0.2em}
 \begin{align*}
@@
 $\Pi$ types.
+\vspace{-0.2em}
 \begin{alignat*}{10}
@@
 indexed. \\
+\vspace{-0.2em}
 \begin{minipage}{0.4\textwidth}
@@
 is straightforward:
+\vspace{-0.2em}
 \begin{alignat*}{10}
@@
 \begin{align*}
-  & \textsf{elim}_{\textsf{Con}} : (\GG : \textsf{Con}) \to C\ \GG \hspace{5em} & \textsf{elim}_{\textsf{Ty}} : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to T\ \GG\ A\ (\textsf{elim}_{\textsf{Con}}\ \GG)
+  & \textsf{elim}_{\textsf{Con}} : (\GG : \textsf{Con}) \to C\ \GG \hspace{5em} \\
+  & \textsf{elim}_{\textsf{Ty}} : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to T\ \GG\ A\ (\textsf{elim}_{\textsf{Con}}\ \GG)
 \end{align*}
@@
 $\mProp$-valued families, rather than in $\mType$:
+\vspace{-0.2em}
 \begin{alignat*}{10}

### 2020-10-12 paper.tex
Fix typos

@@
 satisfies more definitional equalities. For instance, whereas in cubical type
-theory equality of functions is isomorphic to pair-wise equality, in SeTT the
+theory equality of functions is isomorphic to pointwise equality, in SeTT the
 isomorphism is replaced by a definitional equality. SeTT is also a syntactically
 straightforward extension of Martin-L\"of Type Theory, that does not require
-exotic objects like the internal pretype. In turn, the obvious advantage of
+exotic objects like the interval pretype. In turn, the obvious advantage of
 cubical type theory is that it is not limited to setoids.
@@
 with a fibration condition given by $\textsf{coe}$, $\textsf{coh}$. In
-the following, we somtimes omit implicit quantifications such as the $\forall\{\gamma_0\ \gamma_1\}$ in the type of $\textsf{sym}\,\GG$.
+the following, we sometimes omit implicit quantifications such as the $\forall\{\gamma_0\ \gamma_1\}$ in the type of $\textsf{sym}\,\GG$.
 \[
@@
 \Sigma$, etc.), function extensionality and a universe of strict
-propositions ith propositional extensionality \cite{setoid99}. Note that
+propositions with propositional extensionality \cite{setoid99}. Note that
 we don't need identity types or inductive types (W-types) for this.
@@
 
-We can show that $\Uty$ is closed under $\Pi$ types and booleans such
-that $\EL\,(\mathsf{pi}\,A\,B) \equiv \Pi\,(\EL\,A)\,(\EL\,B)$ and
-$\EL\,\mathsf{bool} = \mathsf{Bool}$. The universe can be closed under
-more constructions if more codes are added to $\setoidU$.
+We can show that $\Uty$ is closed under $\Pi$ types and booleans, and satisfies
+$\EL\,(\mathsf{pi}\,A\,B) \equiv \Pi\,(\EL\,A)\,(\EL\,B)$ and
+$\EL\,\mathsf{bool} = \mathsf{Bool}$. The universe can be closed under more
+constructions if more codes are added to $\setoidU$.
 

### 2020-10-13 paper.tex
Add brief clarification on inU~
small fixes abstract

@@
 Although already a rich theory as currently defined, SeTT currently lacks a way
-to internalize the notion of type beyond propositions. This paper is a step
-towards extending SeTT with a universe of setoids.
+to internalize the notion of type beyond propositions, hence we want
+to extend SeTT with a universe of setoids.
 %
@@
 simpler ones ensure that our construction can still be defined in a relatively
-small metatheory, that however must include a proof-irrelevant identity type
-with a strong transport rule. \keywords{type theory \and
+small metatheory which includes a proof-irrelevant identity type
+with a strong transport rule.
+
+\keywords{type theory \and
   function extensionality \and univalence \and setoid model \and
@@
 
+Just as the role of $\inU$ is, as before, to classify all types that are image
+of $\El$, in the same way $\inUU\ p\ q$ classifies all relations of type $A \to
+A' \to \mProp$ that are image of $\_\vdash\_\sim_{\El}\_$,
+given proofs $p : \inU\ A, q : \inU\ A'$.
+%
 Note that $\inU$ appears in the sort of $\inUU$. Since these types are

### 2020-10-13 paper.tex
Fix citations

@@
 a full-fledged foundation in which to develop constructive mathematics
-\cite{mltt75, libretto}.
+\cite{mltt75,libretto}.
 %
@@
 syntactic translation. This is in contrast with other type theories based on the
-setoid model, like Observational Type Theory \cite{alti:ott-conf, alti:ott-conf}
+setoid model, like Observational Type Theory \cite{alti:ott-conf,alti:ott-conf}
 and XTT \cite{xtt}, which instead rely on ETT for their justification. A major
@@
 formal system derived from this model construction \cite{mpc19}. Observational
-Type Theory (OTT) \cite{alti:ott-conf, alti:ott-conf} is a syntax for the setoid
+Type Theory (OTT) \cite{alti:ott-conf,alti:ott-conf} is a syntax for the setoid
 model differing from SeTT for the use of a different notion of heterogeneous
@@
 induction-recursion (IR), a definition schema developed by Dybjer
-\cite{ir-dybjer, dybjer-setzer}. Inductive-recursive definitions can be found
+\cite{ir-dybjer,dybjer-setzer}. Inductive-recursive definitions can be found
 throughout the literature, from the already mentioned type theoretic universes,

### 2020-10-13 paper.tex
Add full definition of inU and inU~

@@
   & \textsf{data}\ \inU : \mType \to \mType_1 \\
-  & \textsf{data}\ \inUU : \{A\ A' : \mType\} \to \inU\ A \to \inU\ A' \to (A \to A' \to \mProp) \to \mType_1
+  & \qquad \textsf{bool} : \inU\ \mathbbm{2} \\
+  & \qquad \pi : \forall\{A\;A_\sim\;B\;B_\sim\}(a : \inU\;A)
+  (b : (x : A) \to \inU\;(B\;x))\\
+  & \qquad\ \to \inU\;a\;a\;A_\sim \to (\forall\{x_0\;x_1\}(x_{01} : A_\sim\;x_0\;x_1)
+  \to \inUU\;(b\;x_0)\;(b\;x_1)\;(B_\sim\;x_{01})) \\
+  & \qquad\ \to \inU\;
+  (\Sigma\;(f : (x : A) \to B\;x) \\
+  &\qquad\qquad\qquad\quad\, ((x_0\;x_1 : A)(x_{01} : A_\sim\;x_0\;x_1) \to B_\sim\;x_{01}\;(f\;x_0)\;(f\;x_1))) \\
+  %%%%%%
+  & \textsf{data}\ \inUU : \{A\ A' : \mType\} \to \inU\ A \to \inU\ A' \to (A \to A' \to \mProp) \to \mType_1 \\
+  & \qquad \mathsf{bool}_\sim : \inUU\;\textsf{bool}\;\textsf{bool}\;
+  (\lambda x_0\;x_1\; .\; x_0 \stackrel{?}{=}_{\mathbbm{2}} x_1) \\
+  & \qquad \pi_\sim :
+  \forall\{A_0\;A_1\;A_{0\sim}\;A_{01\sim}\;B_0\;B_1\;B_{0\sim}\;B_{1\sim}\;B_{01\sim}\}
+  \{ a_0 : \inU\;A_0 \}\{ a_1 : \inU\;A_1 \}\\
+  & \qquad\quad\quad\ \; \{b_0 : (x_0 : A_0) \to \inU\;(B_0\;x_0)\}
+  \{b_1 : (x_1 : A_1) \to \inU\;(B_1\;x_1)\}\\
+  & \qquad\quad\quad\ \; \{a_{0\sim} : \inUU\;a_0\;a_0\;A_{0\sim}\}
+  \{a_{1\sim} : \inUU\;a_1\;a_1\;A_{1\sim}\} \\
+  & \qquad\quad\quad\ \; \{ b_{0\sim} : \forall\{x_0\;x_1\}(x_{01} : A_{0\sim}\;x_0\;x_1)
+    \to \inUU\;(b_0\;x_0)\;(b_0\;x_1)\;(B_{0\sim}\;x_{01}) \}\\
+  & \qquad\quad\quad\ \; \{ b_{1\sim} : \forall\{x_0\;x_1\}(x_{01} : A_{1\sim}\;x_0\;x_1)
+    \to \inUU\;(b_1\;x_0)\;(b_1\;x_1)\;(B_{1\sim}\;x_{01}) \}\\
+  & \qquad\quad \to \inUU\;a_0\;a_1\;A_{01\sim} \\
+    & \qquad\quad \to (\forall\{x_0\;x_1\}(x_{01} : A_{01\sim}\;x_0\;x_1)
+      \to \inUU\;(b_0\;x_0)\;(b_1\;x_1)\;(B_{01\sim}\;x_{01})) \\
+  & \qquad\quad \to \inUU\;(\pi\;a_0\;a_{0\sim}\;b_0\;b_{0\sim})\;
+      (\pi\;a_1\;a_{1\sim}\;b_1\;b_{1\sim}) \\
+  & \qquad\qquad\qquad\quad\,(\lambda f_0\;f_1\;.\;\forall(x_0\;x_1)\to A_{01\sim}\;x_0\;x_1
+        \to B_{01\sim}\;x_{01}\;(\pi_1\;f_0\;x_0)\;(\pi_1\;f_1\;x_1))
 \end{align*}
@@
 of $\El$, in the same way $\inUU\ p\ q$ classifies all relations of type $A \to
-A' \to \mProp$ that are image of $\_\vdash\_\sim_{\El}\_$,
-given proofs $p : \inU\ A, q : \inU\ A'$.
+A' \to \mProp$ that are image of $\_\vdash\_\sim_{\El}\_$, given proofs $p :
+\inU\ A, q : \inU\ A'$.
+%
+In particular, this definition of $\inUU$ states that the appropriate
+equivalence for boolean elements is the obvious syntactic equality
+$\_\stackrel{?}{=}_{\mathbbm{2}}\_$, whereas functions are to be compared
+pointwise.
 %

### 2020-10-14 paper.tex
small fixes

@@
 As an example, consider the following definition of a small universe with
-bool and $\Pi$.
+bool and $\Pi$. \\
 \vspace{-0.2em}
@@
   (b : (x : A) \to \inU\;(B\;x))\\
-  & \qquad\ \to \inU\;a\;a\;A_\sim \to (\forall\{x_0\;x_1\}(x_{01} : A_\sim\;x_0\;x_1)
+  & \qquad\ \to \inUU\;a\;a\;A_\sim \to (\forall\{x_0\;x_1\}(x_{01} : A_\sim\;x_0\;x_1)
   \to \inUU\;(b\;x_0)\;(b\;x_1)\;(B_\sim\;x_{01})) \\
@@
   & \qquad \pi_\sim :
-  \forall\{A_0\;A_1\;A_{0\sim}\;A_{01\sim}\;B_0\;B_1\;B_{0\sim}\;B_{1\sim}\;B_{01\sim}\}
+  \forall\{A_0\;A_1\;A_{0\sim}\;A_{1\sim}\;A_{01\sim}\;B_0\;B_1\;B_{0\sim}\;B_{1\sim}\;B_{01\sim}\}
   \{ a_0 : \inU\;A_0 \}\{ a_1 : \inU\;A_1 \}\\
@@
 Just as the role of $\inU$ is, as before, to classify all types that are image
-of $\El$, in the same way $\inUU\ p\ q$ classifies all relations of type $A \to
-A' \to \mProp$ that are image of $\_\vdash\_\sim_{\El}\_$, given proofs $p :
-\inU\ A, q : \inU\ A'$.
+of $\El$, in the same way $\inUU\ a\ a'$ classifies all relations of type $A \to
+A' \to \mProp$ that are image of $\_\vdash\_\sim_{\El}\_$, given proofs $a :
+\inU\ A, a' : \inU\ A'$.
 %
@@
 \[
-(A, p_A) \sim_{\setoidU} (B, p_B) :\equiv \Sigma\ (R: A \to B \to \mProp)\ (\inUU\ p_A\ p_B\ R)
+(A, a) \sim_{\setoidU} (B, b) :\equiv \Sigma\ (R: A \to B \to \mProp)\ (\inUU\ a\ b\ R)
 \]
@@
   & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \mProp_1 \\
-  & (X , p) \sim_{\setoidU} (X' , p') =
-   \ \parallel \Sigma\ (R : X \to X' \to \mProp)\ (\inUU\ p\ p'\ R) \parallel
+  & (A , a) \sim_{\setoidU} (B , b) =
+   \ \parallel \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \parallel
 \end{align*}
@@
   & \_\vdash\_\sim_{\El}\_ : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
-  & p \vdash a \sim_{\El} b :\equiv \ ?
+  & p \vdash x \sim_{\El} y :\equiv \ ?
 \end{align*}
@@
 \mProp)\ (...) \parallel$. If the type wasn't propositionally truncated, we could define
-$p \vdash a \sim_{\El} b$ by extracting the relation out of the first component
-of $p$, and apply it to $a, b$. That is, $p \vdash a \sim_{\El} b :\equiv
-\pi_1\,p\,a\,b$. This would make the definition of $\_\sim_\setoidU\_$ and
+$p \vdash x \sim_{\El} y$ by extracting the relation out of the first component
+of $p$, and apply it to $x, y$. That is, $p \vdash x \sim_{\El} y :\equiv
+\pi_1\,p\,x\,y$. This would make the definition of $\_\sim_\setoidU\_$ and
 $\_\vdash\_\sim_{\El}\_$ in line with how we defined $\setoidU$ and $\El$.
@@
 up being logically equivalent to the proposition we would have obtained by
-$\mathsf{proj}_1\ p\ x\ y$ if there were no truncation.
+$\pi_1\ p\ x\ y$ if there were no truncation.
 %
@@
   & \bullet && : \textsf{Con}                                                          &&  \iota && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
-  & \_,\_ && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{5em} &&  \Pi && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
+  & \_,\_ && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{3em} &&  \Pi && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
 \end{alignat*}
@@
   & \bullet_0 && : \textsf{Con}_0 \\
-  & \_,_0\_ && : (\GG : \textsf{Con}_0) \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
-  & \iota_0 && : (\GG : \textsf{Con}_0) \to \textsf{Ty}_0 \\
+  & \_,_0\_ && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
+  & \iota_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \\
   & \Pi_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \hspace{2.8em} \\
@@
 \mType$, the general eliminators have the following signatures:
-\begin{align*}
-  & \textsf{elim}_{\textsf{Con}} : (\GG : \textsf{Con}) \to C\ \GG \hspace{5em} \\
-  & \textsf{elim}_{\textsf{Ty}} : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to T\ \GG\ A\ (\textsf{elim}_{\textsf{Con}}\ \GG)
-\end{align*}
+\begin{alignat*}{10}
+  & \textsf{elim}_{\textsf{Con}} && : (\GG : \textsf{Con}) \to C\ \GG \hspace{5em} \\
+  & \textsf{elim}_{\textsf{Ty}} && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to T\ \GG\ A\ (\textsf{elim}_{\textsf{Con}}\ \GG)
+\end{alignat*}
 
@@
 the graph of the eliminators in the form of inductively-generated relations:
-\begin{align*}
-  & \textsf{data}\ \textsf{R-Con} :
+\begin{alignat*}{10}
+  & \textsf{data}\ \textsf{R-Con} && :
     (\GG : \textsf{Con}) \to C\ \GG \to \mType \\
-  & \textsf{data}\ \textsf{R-Ty} : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG)
+  & \textsf{data}\ \textsf{R-Ty} && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG)
   (\gamma : C\ \GG)
@@
   \to \mType
-\end{align*}
+\end{alignat*}
 

### 2021-01-14 paper.tex
added todo for final version

@@
 syntactic translation. This is in contrast with other type theories based on the
-setoid model, like Observational Type Theory \cite{alti:ott-conf,alti:ott-conf}
+setoid model, like Observational Type Theory \cite{alti:ott-conf}
 and XTT \cite{xtt}, which instead rely on ETT for their justification. A major
@@
 universe should support universe induction, thus exposing the inductive
-structure of the codes. Such elimination principle is known to be inconsistent
+structure of the codes. Such an elimination principle is known to be inconsistent
 with univalence, although this is not an issue in our case; nevertheless it's

### 2021-01-16 paper.tex
added Palmgren and small fixes

@@
   Ambrus Kaposi\inst{3}\orcidID{0000-0001-9897-8936}
-  \thanks{Supported by the National Research, Development and Innovation Fund of Hungary, financed under the Thematic Excellence Programme funding scheme, Project no.
-ED\_18-1-2019-0030 and the Bolyai Fellowship of the Hungarian Academy of Sciences, Project no.\ BO/00659/19/3.}\and
+  \thanks{Supported by the Bolyai Fellowship of the Hungarian Academy
+    of Sciences (BO/00659/19/3) and by the ``Application Domain
+    Specific Highly Reliable IT Solutions'' project that has been
+    implemented with support from the National Research, Development
+    and Innovation Fund of Hungary, financed under the Thematic
+    Excellence Programme TKP2020-NKA-06 funding scheme.}\and
   Christian Sattler\inst{4}\thanks{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.} \and
@@
   \email{\{psztxa,psxfs5\}@nottingham.ac.uk} \and
-  Inria, Nantes, France\email{simon.boulier@inria.fr} \and
+  Inria, Nantes, France \email{simon.boulier@inria.fr} \and
   E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary
@@
 formal system derived from this model construction \cite{mpc19}. Observational
-Type Theory (OTT) \cite{alti:ott-conf,alti:ott-conf} is a syntax for the setoid
-model differing from SeTT for the use of a different notion of heterogeneous
+Type Theory (OTT) \cite{alti:ott-conf} is a syntax for the setoid
+model differing from SeTT in the use of a different notion of heterogeneous
 equality. Moreover, the consistency proof for OTT relies on Extensional Type
@@
 Palmgren and Wilander \cite{2014arXiv1408.1364P} construct a setoid
-universe using a translation into constructive set theory. In
-contrast, we target a basic intensional type theory and hence also
-give a computational semantics.
+universe using a translation into constructive set theory. Palmgren
+\cite{2019arXiv190901414P} constructs an encoding of ETT in ITT
+through Aczel's encoding of set theory in type theory
+\cite{ACZEL197855}. He uses type theory as a language for his
+formalisation but his construction is set-theoretic in nature. Setoids
+are utilized to encode sets as arbitrarily branching well-founded
+trees quotiented by bisimulation. His notion of family of setoids does
+not use strict propositions and it has a weaker form of proof
+irrelevance which seems to be not enough to obtain a model of SeTT.
 

### 2021-01-16 paper.tex
further work

@@
 
+It is known that finitary IITs can be reduced to inductive types in an
+extensional setting \cite{induction-is-enough}. In our paper we reduce
+an infinitary IIT to inductive types in an intensional setting. In the
+future, we would like to investigate whether this reduction can be
+generalised to arbitrary infinitary IITs.
+
 In contrast to the inductive-recursive and inductive-inductive

### 2021-01-19 paper.tex
changes from todo-new

@@
 satisfying propositional extensionality. The setoid model thus satisfies all the
-extensionality principles that we would like to have in type theory. The
+extensionality principles that we would like to have in a set-level
+type theory
+\footnote{In the sense of HoTT we mean a type theory limited to h-sets.}
+. The
 question is whether there exists a version of intensional type theory that
@@
 XTT \cite{xtt} is a cubical variant of OTT where the equality type is defined
-using an interval pretype. XTT's universes support universe induction, whereas
+using an interval pretype
+\\footnote{To quote one of the referees: \emph{the fact that the
+    interval is a pretype is but the easiest part of the story.}}
+. XTT's universes support universe induction, whereas
 it is left open whether the construction presented here supports this principle.
@@
 data type transformation methods in a novel way that accounts for the
-peculiarities of our construction.
+peculiarities of our construction. In \cite{mpc19} we present rules
+for SetTT, clearly these rules need to be extended by the rules for a
+universe reflecting the semantics presented here.
 %

### 2021-01-21 paper.tex
Typo.

@@
 
-We conclude the paper with a discussion of further work in
-Section~\ref{further-work}.
+% We conclude the paper with a discussion of further work in
+% Section~\ref{further-work}.
 
@@
 using an interval pretype
-\\footnote{To quote one of the referees: \emph{the fact that the
+\footnote{To quote one of the referees: \emph{the fact that the
     interval is a pretype is but the easiest part of the story.}}
@@
 impredicativity with a strong transport rule results in terms that fail to
-normalize.
+normalize but this is irrelevant in our setting.
 
@@
 then define $\_\sim_{\setoidU}\_$.
-\vspace{-0.2em}
+{\small\vspace{-0.2em}
 \begin{align*}
@@
         \to B_{01\sim}\;x_{01}\;(\pi_1\;f_0\;x_0)\;(\pi_1\;f_1\;x_1))
-\end{align*}
+\end{align*}}
 
@@
 Unfortunately this definition only manages to capture the idea, but does not
-actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be $\mProp_1$-valued
+actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be a $\mProp_1$-valued
 relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the

### 2021-01-21 paper.tex
small changes
after meeting
Clarified sort hierarchy. @Ambrus: Check for consistency with formalization.

@@
 go beyond inductive families.
-
-$\mlttp$ is thus an intensional type theory in the style of Martin-L\"of type
-theory with $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$, and a universe
-of strict propositions $\mProp$. $\Sigma$ types are defined negatively by
-pairing $\_,\_$ and projections $\pi_1, \pi_2$. We also have definitional $\eta$
-rules for $\Pi, \Sigma$, and $\mathbf{1}$ types. Propositions can be lifted to
-types via a lifting $\textsf{Lift} : \mProp \to \mType$, with constructor
+$\mlttp$ is thus an intensional type theory in the style of Martin-L\"of type theory.
+
+We have sorts $\Type_i$ of types and $\mProp_i$ of strict propositions for $i \in \{0, 1\}$.
+Here, $i = 0$ means ``small'' (and we will omit the subscript) and $i = 1$ means ``large''.
+We have implicit lifting from $i = 0$ to $i = 1$, but do not assume type formers are preserved.
+$\Type_1$ has universes for $\Type$ and $\mProp$.
+We do not distinguish notationally between universes and sorts.
+We continue to describe only the case $i = 0$; everything introduced has an analogue at level $i = 1$.
+Propositions lift to types via $\textsf{Lift} : \mProp \to \mType$, with constructor
 $\textsf{lift} : \{P : \mProp \} \to P \to \textsf{Lift}\ P$ and destructor
@@
 
+We have standard type formers $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$ in $\Type$.
+$\Sigma$-types are defined negatively by pairing $\_,\_$ and projections $\pi_1, \pi_2$.
+We also have definitional $\eta$-rules for $\Pi, \Sigma$, and $\mathbf{1}$ types.
 We also require indexed W-types, both in $\mType$ and $\mProp$; hence we have $W_\Box
 : (S : I\to\mType) \to ((i:I)\to S\ i\to I \to \mType) \to I\to \Box$ where $\Box \in \{\mType, \mProp\}$.
-The elimination principle of $W_\mProp$ only allows defining functions into $\mProp$s.
-%
-From $W_\mProp$ we can define propositional truncation $\|\_\| : \mType \to \mProp$, with constructor
+The elimination principle of $W_{\mProp}$ only allows defining functions into elements of $\mProp$.
+%
+From $W_{\mProp}$ we can define propositional truncation $\|\_\| : \mType \to \mProp$, with constructor
   $|\_| : \{A : \mType \} \to A \to \| A \|$ and eliminator
@@
 using the equivalence of $I$-indexed families of types and types over $I$
-(that is, $\mType_i \to I$) \cite{malatasta13smallir}.
+(that is, $\mType_i \to I$) \cite{malatasta13smallir}. % TODO: fix wrong formula
 

### 2021-01-21 paper.tex
Fix incorrect description of types over I.

@@
 using the equivalence of $I$-indexed families of types and types over $I$
-(that is, $\mType_i \to I$) \cite{malatasta13smallir}. % TODO: fix wrong formula
+(that is, $A : \mType$ with $A \to I$) \cite{malatasta13smallir}.
 

### 2021-01-21 paper.tex
Expand apostrophic abbreviations: * it's -> it is * can't -> cannot * wasn't -> was not * ...

@@
 types instead of deriving them from truncation. We operate under the assumption
-that everything can be equivalently carried out in $\mlttp$, although we haven't
+that everything can be equivalently carried out in $\mlttp$, although we have not
 fully checked all the necessary details.
@@
 propositions with propositional extensionality \cite{setoid99}. Note that
-we don't need identity types or inductive types (W-types) for this.
+we do not need identity types or inductive types (W-types) for this.
 
@@
 structure of the codes. Such an elimination principle is known to be inconsistent
-with univalence, although this is not an issue in our case; nevertheless it's
+with univalence, although this is not an issue in our case; nevertheless it is
 not immediately clear if the elimination principle can be justified by the
@@
 In the next section we transform our current inductive-recursive definition to
-one that doesn't use induction-recursion.
+one that does not use induction-recursion.
 %
@@
 relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
-$\Sigma$ type shown above clearly isn't, since it quantifies over a type of
+$\Sigma$ type shown above clearly is not, since it quantifies over a type of
 relations, which is not a proposition.
@@
 In the definition above, $p$ has type $\parallel \Sigma\ (R: \El\,A \to \El\,B \to
-\mProp)\ (...) \parallel$. If the type wasn't propositionally truncated, we could define
+\mProp)\ (...) \parallel$. If the type was not propositionally truncated, we could define
 $p \vdash x \sim_{\El} y$ by extracting the relation out of the first component
@@
 
-However, this doesn't work in our case, since the type of $p$ \emph{is}
+However, this does not work in our case, since the type of $p$ \emph{is}
 propositionally truncated, hence it cannot be eliminated to construct a
@@
 more or less straightforwardly. The final result does not use
-induction-recursion, but it's nevertheless an instance of infinitary
+induction-recursion, but it is nevertheless an instance of infinitary
 induction-induction. The ability to define arbitrary, infinitary
@@
 %
-Of course, the absence of a general reduction method does not mean that we can't
+Of course, the absence of a general reduction method does not mean that we cannot
 reduce particular concrete instances of infinitary induction-induction, which is
@@
 to derive the general recursive-recursive eliminators we need to prove that the
-corresponding graph relations are functional, which can't be done without funext.
+corresponding graph relations are functional, which cannot be done without funext.
 

### 2021-01-21 paper.tex
Use \blank for placeholders.

@@
 
+% for placeholders
+\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
+
 % TODO notes.
@@
 We have standard type formers $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$ in $\Type$.
-$\Sigma$-types are defined negatively by pairing $\_,\_$ and projections $\pi_1, \pi_2$.
-We also have definitional $\eta$-rules for $\Pi, \Sigma$, and $\mathbf{1}$ types.
-We also require indexed W-types, both in $\mType$ and $\mProp$; hence we have $W_\Box
-: (S : I\to\mType) \to ((i:I)\to S\ i\to I \to \mType) \to I\to \Box$ where $\Box \in \{\mType, \mProp\}$.
+$\Sigma$-types are defined negatively by pairing $\blank,\blank$ and projections $\pi_1, \pi_2$.
+We have definitional $\eta$-rules for $\Pi$-, $\Sigma$-, $\mathbf{1}$-types.
+We also require indexed W-types, both in $\mType$ and $\mProp$:
+$W_\Box : (S : I\to\mType) \to ((i:I)\to S\ i\to I \to \mType) \to I\to \Box$ where $\Box \in \{\mType, \mProp\}$.
 The elimination principle of $W_{\mProp}$ only allows defining functions into elements of $\mProp$.
 %
-From $W_{\mProp}$ we can define propositional truncation $\|\_\| : \mType \to \mProp$, with constructor
-  $|\_| : \{A : \mType \} \to A \to \| A \|$ and eliminator
-  $\textsf{elim}_{\|\_\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P$.
+From $W_{\mProp}$ we can define propositional truncation $\|\blank\| : \mType \to \mProp$, with constructor
+  $|\blank| : \{A : \mType \} \to A \to \| A \|$ and eliminator
+  $\textsf{elim}_{\|\blank\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P$.
 
@@
   & \textsf{data} \ \setoidU : \mType_1 \\
-  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
+  & \blank\sim_\setoidU\blank : \setoidU \to \setoidU \to \mProp_1
 \end{align*}
@@
   & \El : \setoidU \to \mType \\
-  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
+  & \blank\vdash\blank\sim_\El\blank : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
 \end{align*}
 
-Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
+Note that $\eqel{\blank}{\blank}{\blank}$ is indexed over equality on the universe, because
 $\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
@@
 of the 4 functions
-$\_\sim_\setoidU\_$, $\textsf{refl}_\setoidU$, $\El$ and $\_\vdash\_\sim_\El\_$.
+$\blank\sim_\setoidU\blank$, $\textsf{refl}_\setoidU$, $\El$ and $\blank\vdash\blank\sim_\El\blank$.
 The other functions are then recursively defined:
@@
 $\inUU$ of binary relations between types in the universe, from which we
-then define $\_\sim_{\setoidU}\_$.
+then define $\blank\sim_{\setoidU}\blank$.
 {\small\vspace{-0.2em}
@@
 of $\El$, in the same way $\inUU\ a\ a'$ classifies all relations of type $A \to
-A' \to \mProp$ that are image of $\_\vdash\_\sim_{\El}\_$, given proofs $a :
+A' \to \mProp$ that are image of $\blank\vdash\blank\sim_{\El}\blank$, given proofs $a :
 \inU\ A, a' : \inU\ A'$.
@@
 equivalence for boolean elements is the obvious syntactic equality
-$\_\stackrel{?}{=}_{\mathbbm{2}}\_$, whereas functions are to be compared
+$\blank\stackrel{?}{=}_{\mathbbm{2}}\blank$, whereas functions are to be compared
 pointwise.
@@
 Unfortunately this definition only manages to capture the idea, but does not
-actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be a $\mProp_1$-valued
+actually typecheck. In fact, $\blank\sim_{\setoidU}\blank$ should be a $\mProp_1$-valued
 relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
@@
 \begin{align*}
-  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \mProp_1 \\
+  & \blank\sim_{\setoidU}\blank : \setoidU \to \setoidU \to \mProp_1 \\
   & (A , a) \sim_{\setoidU} (B , b) =
@@
 \begin{align*}
-  & \_\vdash\_\sim_{\El}\_ : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
+  & \blank\vdash\blank\sim_{\El}\blank : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
   & p \vdash x \sim_{\El} y :\equiv \ ?
@@
 of $p$, and apply it to $x, y$. That is, $p \vdash x \sim_{\El} y :\equiv
-\pi_1\,p\,x\,y$. This would make the definition of $\_\sim_\setoidU\_$ and
-$\_\vdash\_\sim_{\El}\_$ in line with how we defined $\setoidU$ and $\El$.
+\pi_1\,p\,x\,y$. This would make the definition of $\blank\sim_\setoidU\blank$ and
+$\blank\vdash\blank\sim_{\El}\blank$ in line with how we defined $\setoidU$ and $\El$.
 
@@
 are logically equivalent. These in turn need to be defined mutually with
-$\_\vdash\_\sim_{\El}\_$. We direct the interested reader to the Agda
+$\blank\vdash\blank\sim_{\El}\blank$. We direct the interested reader to the Agda
 formalization for the full details of these definitions, as they are quite
@@
 Contexts in \textsf{Con} are formed out of empty contexts $\bullet$ and context
-extension $\_,\_$. Types in \textsf{Ty} are either the base type $\iota$ or
+extension $\blank,\blank$. Types in \textsf{Ty} are either the base type $\iota$ or
 $\Pi$ types.
@@
   & \bullet && : \textsf{Con}                                                          &&  \iota && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
-  & \_,\_ && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{3em} &&  \Pi && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
+  & \blank,\blank && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{3em} &&  \Pi && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
 \end{alignat*}
@@
   & \bullet_0 && : \textsf{Con}_0 \\
-  & \_,_0\_ && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
+  & \blank,_0\blank && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
   & \iota_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \\
@@
   & \bullet_1 && : \textsf{Con}_1\ \bullet_0 \\
-  & \_,_1\_ && : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
+  & \blank,_1\blank && : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
   & && \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\

### 2021-01-21 paper.tex
Fix gross LaTeX misuse: * \parallel is a relation symbol, not to be used for the truncation operator * replaced '...' in formulas by \ldots
Use macro \def with vertically centered colon for defining equalities.

@@
 \usepackage{bbm}
+\usepackage{mathtools}
 %\usepackage[bw]{agda}
@@
 
+% for definitions
+\newcommand{\defeq}{\mathrel{\vcentcolon\equiv}}
+
 % TODO notes.
@@
 \begin{align*}
-  \mathbf{1}_\mProp & :\equiv \| \mathbf{1} \| \\
-  \Pi_\mProp\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
-  \Sigma_\mProp\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \|
+  \mathbf{1}_\mProp & \defeq \| \mathbf{1} \| \\
+  \Pi_\mProp\ P\ Q & \defeq \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
+  \Sigma_\mProp\ P\ Q & \defeq \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \|
 \end{align*}
@@
 %% \begin{align*}
-%%   & (\Pi(a : A)\ B)^\sim\ p\ f_0\ f_1 = \Pi(a_0\,a_1 : A)... \\
-%%   & (\Sigma\ A\ B)^\sim\ p\ (a_0, b_0)\ (a_1, b_1) = ... \\
-%%   & \Prop^\sim\ p\ P\ Q = ...
+%%   & (\Pi(a : A)\ B)^\sim\ p\ f_0\ f_1 = \Pi(a_0\,a_1 : A)\ldots \\
+%%   & (\Sigma\ A\ B)^\sim\ p\ (a_0, b_0)\ (a_1, b_1) = \ldots \\
+%%   & \Prop^\sim\ p\ P\ Q = \ldots
 %% \end{align*}
@@
 \infer{
-  \textsf{Id}_A\ a_0\ a_1 :\equiv A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
+  \textsf{Id}_A\ a_0\ a_1 \defeq A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
 }{
@@
 \infer{
-  \textsf{transp}\ P\ p\ t :\equiv
+  \textsf{transp}\ P\ p\ t \defeq
     \coe\ P\ (\textsf{R}\ \textsf{id} , p)\ t : \Tm\ \GG\ P[a_1]
@@
   & \El\ : \U \to \mType \\
-  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
-  & \El\ (\textsf{pi}\ A\ B) :\equiv (a : \El\ A) \to \El\ (B\ a) \\
+  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
+  & \El\ (\textsf{pi}\ A\ B) \defeq (a : \El\ A) \to \El\ (B\ a) \\
 \end{align*}
@@
 
-$\U$ and $\El$ can be given by $\U :\equiv \Sigma\ (A : \mType)\ (\textsf{in-U}\ A)$ and $\El :\equiv \pi_1$.
+$\U$ and $\El$ can be given by $\U \defeq \Sigma\ (A : \mType)\ (\textsf{in-U}\ A)$ and $\El \defeq \pi_1$.
 
@@
 \begin{align*}
-  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
-  & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
+  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
+  & \El\ (\textsf{pi}\ A\ B\ h) \defeq \\
   & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
@@
 \begin{align*}
-  & | \Uty | :\equiv \lambda\, \gamma . \, \setoidU \\
-  & \Uty^\sim :\equiv \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
-  & \textsf{refl}\ \Uty :\equiv \reflu \\
-  & ... \\
-  & \textsf{coe}\ \Uty :\equiv \lambda\,p\,a . \, a \\
-  & \textsf{coh}\ \Uty :\equiv \lambda\,p . \, \reflu
+  & | \Uty | \defeq \lambda\, \gamma . \, \setoidU \\
+  & \Uty^\sim \defeq \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
+  & \textsf{refl}\ \Uty \defeq \reflu \\
+  & \ldots \\
+  & \textsf{coe}\ \Uty \defeq \lambda\,p\,a . \, a \\
+  & \textsf{coh}\ \Uty \defeq \lambda\,p . \, \reflu
 \end{align*}
@@
 \begin{align*}
-  & | \EL\ A | :\equiv \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
-  & (\EL\ A)^\sim :\equiv \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
-  & \textsf{refl}\ (\EL\ A) :\equiv \reflel \\
-  & ... \\
-  & \textsf{coe}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
-  & \textsf{coh}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
+  & | \EL\ A | \defeq \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
+  & (\EL\ A)^\sim \defeq \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
+  & \textsf{refl}\ (\EL\ A) \defeq \reflel \\
+  & \ldots \\
+  & \textsf{coe}\ (\EL\ A) \defeq \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
+  & \textsf{coh}\ (\EL\ A) \defeq \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
 \end{align*}
@@
   & \setoidU && : \mType_1 && \El : \setoidU \to \mType \\
-  & \setoidU && :\equiv \Sigma\ (X : \mType)\ (\inU\ X) \hspace{5em} && \El :\equiv \pi_1
+  & \setoidU && \defeq \Sigma\ (X : \mType)\ (\inU\ X) \hspace{5em} && \El \defeq \pi_1
 \end{alignat*}
@@
 \[
-(A, a) \sim_{\setoidU} (B, b) :\equiv \Sigma\ (R: A \to B \to \mProp)\ (\inUU\ a\ b\ R)
+(A, a) \sim_{\setoidU} (B, b) \defeq \Sigma\ (R: A \to B \to \mProp)\ (\inUU\ a\ b\ R)
 \]
@@
   & (A , a) \sim_{\setoidU} (B , b) =
-   \ \parallel \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \parallel
+   \| \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \|
 \end{align*}
@@
   & \blank\vdash\blank\sim_{\El}\blank : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
-  & p \vdash x \sim_{\El} y :\equiv \ ?
+  & p \vdash x \sim_{\El} y \defeq \ ?
 \end{align*}
 
-In the definition above, $p$ has type $\parallel \Sigma\ (R: \El\,A \to \El\,B \to
-\mProp)\ (...) \parallel$. If the type was not propositionally truncated, we could define
+In the definition above, $p$ has type $\| \Sigma\ (R: \El\,A \to \El\,B \to
+\mProp)\ (\ldots) \|$. If the type was not propositionally truncated, we could define
 $p \vdash x \sim_{\El} y$ by extracting the relation out of the first component
-of $p$, and apply it to $x, y$. That is, $p \vdash x \sim_{\El} y :\equiv
+of $p$, and apply it to $x, y$. That is, $p \vdash x \sim_{\El} y \defeq
 \pi_1\,p\,x\,y$. This would make the definition of $\blank\sim_\setoidU\blank$ and
@@
 
-We can recover the original inductive-inductive type as $\textsf{Con} :\equiv
+We can recover the original inductive-inductive type as $\textsf{Con} \defeq
 \Sigma\ (\GG_0 : \textsf{Con}_0)\ (\textsf{Con}_1\ \GG_0)$ and $\textsf{Ty}\ \GG
-:\equiv \Sigma\ (A_0 :
+\defeq \Sigma\ (A_0 :
 \textsf{Ty}_0)\ (\textsf{Ty}_1\ (\pi_1\ \GG)\ A_0)$. Recovering the constructors
@@
 \begin{alignat*}{10}
-  & \bullet && :\equiv (\bullet_0 , \bullet_1) \\
-  & (\GG_0 , \GG_1) , (A_0 , A_1) && :\equiv ((\GG_0 \mathbin{,_0} A_0) , (\GG_1 ,_1 A_1)) \\
-  & \iota\ (\GG_0 , \GG_1) && :\equiv (\iota_0\ \GG_0 , \iota_1\ \GG_1) \\
-  & \Pi\ \{\GG_0 , \GG_1\}(A_0 , A_1) (B_0 , B_1) && :\equiv
+  & \bullet && \defeq (\bullet_0 , \bullet_1) \\
+  & (\GG_0 , \GG_1) , (A_0 , A_1) && \defeq ((\GG_0 \mathbin{,_0} A_0) , (\GG_1 ,_1 A_1)) \\
+  & \iota\ (\GG_0 , \GG_1) && \defeq (\iota_0\ \GG_0 , \iota_1\ \GG_1) \\
+  & \Pi\ \{\GG_0 , \GG_1\}(A_0 , A_1) (B_0 , B_1) && \defeq
     (\Pi_0\ \GG_0\ A_0\ B_0 , \Pi_1\ \GG_1\ A_1\ B_1)

### 2021-01-21 paper.tex
Reduced some weird align hackery. Changed aligning to be with respect to the typing colon in several places. Undo commit if not desired.

@@
 \vspace{-0.2em}
-\begin{alignat*}{10}
-  & \textsf{Id} && :\,\, && \{A : \mType \} \to A \to A \to \mProp \\
-  & \textsf{refl} && : && \{A : \mType \} (a : A) \to \textsf{Id}\ a\ a \\
-  & \textsf{transp} && : && \{A : \mType \} (C : A \to \mType) \{a_0\ a_1 : A\} \to \textsf{Id}\ a_0\ a_1 \to C\ a_0 \to C\ a_1 \\
-  & && && \textsf{transp}\ C\ \{x\}\ \{x\}\ e\ u \equiv u
-\end{alignat*}
-
+\begin{align*}
+  \textsf{Id} &: \{A : \mType \} \to A \to A \to \mProp \\
+  \textsf{refl} &: \{A : \mType \} (a : A) \to \textsf{Id}\ a\ a \\
+  \textsf{transp} &: \{A : \mType \} (C : A \to \mType) \{a_0\ a_1 : A\} \to \textsf{Id}\ a_0\ a_1 \to C\ a_0 \to C\ a_1
+\end{align*}
+with $\textsf{transp}\ C\ \{x\}\ \{x\}\ e\ u \equiv u$.
 The \textsf{transp} combinator provides a strong elimination principle allowing
@@
   \begin{aligned}
-    & | \GG | : \mType \\
-    & \GG^\sim : | \GG | \to | \GG | \to \mProp \\
-    & \textsf{refl}\ \GG : (\gamma : | \GG |) \to \GG^\sim\,\gamma\,\gamma \\
-    & \textsf{sym} \ \GG : \forall\{\gamma_0\ \gamma_1\} \to \GG^\sim\,\gamma_0\,\gamma_1
+    | \GG | &: \mType \\
+    \GG^\sim &: | \GG | \to | \GG | \to \mProp \\
+    \textsf{refl}\ \GG &: (\gamma : | \GG |) \to \GG^\sim\,\gamma\,\gamma \\
+    \textsf{sym} \ \GG &: \forall\{\gamma_0\ \gamma_1\} \to \GG^\sim\,\gamma_0\,\gamma_1
     \to \GG^\sim\,\gamma_1\,\gamma_0 \\
-    & \textsf{trans} \ \GG : \forall\{\gamma_0\ \gamma_1\ \gamma_2\}
+    \textsf{trans} \ \GG &: \forall\{\gamma_0\ \gamma_1\ \gamma_2\}
     \to \GG^\sim\,\gamma_0\,\gamma_1
@@
   \begin{aligned}
-    & | A | : | \GG | \to \mType \\
-    & A^\sim : \{\gamma_0\ \gamma_1 : | \GG |\}
+    | A | &: | \GG | \to \mType \\
+    A^\sim &: \{\gamma_0\ \gamma_1 : | \GG |\}
     \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
-    & \textsf{refl*} : \{\gamma : | \GG | \} (a : | A | \gamma) \to A^\sim\ (\textsf{refl}\ \GG\ \gamma)\ a\ a \\
-    & \textsf{sym*} : \forall\{\gamma_0\,\gamma_1\,a_0\,a_1\}\{p : \GG^\sim\ \gamma_0\ \gamma_1\}
+    \textsf{refl*} &: \{\gamma : | \GG | \} (a : | A | \gamma) \to A^\sim\ (\textsf{refl}\ \GG\ \gamma)\ a\ a \\
+    \textsf{sym*} &: \forall\{\gamma_0\,\gamma_1\,a_0\,a_1\}\{p : \GG^\sim\ \gamma_0\ \gamma_1\}
     \to A^\sim\ p\ a_0\ a_1 \to A^\sim\ (\textsf{sym}\ \GG\ p)\ a_1\ a_0 \\
-    & \textsf{trans*} : A^\sim\ p_0\ a_0\ a_1\to A^\sim\ p_1\ a_1\ a_2\to A^\sim\ (\textsf{trans}\,\GG\,p_0\,p_1)\ a_0\ a_2\\
-    & \textsf{coe} : \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
-    & \textsf{coh} : (p : \GG^\sim\ \gamma_0\ \gamma_1) (a : | A | \gamma_0)
+    \textsf{trans*} &: A^\sim\ p_0\ a_0\ a_1\to A^\sim\ p_1\ a_1\ a_2\to A^\sim\ (\textsf{trans}\,\GG\,p_0\,p_1)\ a_0\ a_2\\
+    \textsf{coe} &: \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
+    \textsf{coh} &: (p : \GG^\sim\ \gamma_0\ \gamma_1) (a : | A | \gamma_0)
     \to A^\sim\,p\,a\,(\textsf{coe}\,A\,p\,a) \\
@@
   \begin{aligned}
-    & | \sigma | : | \GG | \to | \GD | \\
-    & \sigma^\sim : \GG^\sim\, \rho_0\, \rho_1 \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
+    | \sigma | &: | \GG | \to | \GD | \\
+    \sigma^\sim &: \GG^\sim\, \rho_0\, \rho_1 \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
   \end{aligned}
@@
   \begin{aligned}
-    & | t | : (\gamma : | \GG |) \to | A | \, \gamma \\
-    & t^\sim : (p : \GG^\sim\ \gamma_0\ \gamma_1) \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
+    | t | &: (\gamma : | \GG |) \to | A | \, \gamma \\
+    t^\sim &: (p : \GG^\sim\ \gamma_0\ \gamma_1) \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
   \end{aligned}
@@
 \begin{align*}
-  & \textsf{data} \ \setoidU : \mType_1 \\
-  & \blank\sim_\setoidU\blank : \setoidU \to \setoidU \to \mProp_1
+  \textsf{data} \ \setoidU &: \mType_1 \\
+  \blank\sim_\setoidU\blank &: \setoidU \to \setoidU \to \mProp_1
 \end{align*}
@@
 \begin{align*}
-  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
-  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
-  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
+  \textsf{refl}_\setoidU &: (A : \setoidU) \to \equ{A}{A} \\
+  \textsf{sym}_\setoidU &: \equ{A}{B} \to \equ{B}{A} \\
+  \textsf{trans}_\setoidU &: \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
 \end{align*}
@@
 \begin{align*}
-  & \El : \setoidU \to \mType \\
-  & \blank\vdash\blank\sim_\El\blank : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
+  \El &: \setoidU \to \mType \\
+  \blank\vdash\blank\sim_\El\blank &: \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
 \end{align*}
@@
 \begin{align*}
-  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
-  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
-  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
-  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
-  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
+  \textsf{refl}_\El &: (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
+  \textsf{sym}_\El &: \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
+  \textsf{trans}_\El &: \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
+  \coeel &: \equ{A}{B} \to \El\ A \to \El\ B \\
+  \cohel &: (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
 \end{align*}
@@
 \begin{align*}
-  & | \Uty | \defeq \lambda\, \gamma . \, \setoidU \\
-  & \Uty^\sim \defeq \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
-  & \textsf{refl}\ \Uty \defeq \reflu \\
-  & \ldots \\
-  & \textsf{coe}\ \Uty \defeq \lambda\,p\,a . \, a \\
-  & \textsf{coh}\ \Uty \defeq \lambda\,p . \, \reflu
+  | \Uty | &\defeq \lambda\, \gamma . \, \setoidU \\
+  \Uty^\sim &\defeq \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
+  \textsf{refl}\ \Uty &\defeq \reflu \\
+  &\ldots \\
+  \textsf{coe}\ \Uty &\defeq \lambda\,p\,a . \, a \\
+  \textsf{coh}\ \Uty &\defeq \lambda\,p . \, \reflu
 \end{align*}
@@
 \begin{align*}
-  & | \EL\ A | \defeq \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
-  & (\EL\ A)^\sim \defeq \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
-  & \textsf{refl}\ (\EL\ A) \defeq \reflel \\
-  & \ldots \\
-  & \textsf{coe}\ (\EL\ A) \defeq \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
-  & \textsf{coh}\ (\EL\ A) \defeq \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
+  | \EL\ A | &\defeq \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
+  (\EL\ A)^\sim &\defeq \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
+  \textsf{refl}\ (\EL\ A) &\defeq \reflel \\
+  &\ldots \\
+  \textsf{coe}\ (\EL\ A) &\defeq \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
+  \textsf{coh}\ (\EL\ A) &\defeq \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
 \end{align*}
@@
 \vspace{-0.2em}
-\begin{alignat*}{10}
-  & \setoidU && : \mType_1 && \El : \setoidU \to \mType \\
-  & \setoidU && \defeq \Sigma\ (X : \mType)\ (\inU\ X) \hspace{5em} && \El \defeq \pi_1
-\end{alignat*}
+\begin{align*}
+  \setoidU & : \mType_1 & \El &: \setoidU \to \mType \\
+  \setoidU & \defeq \Sigma\ (X : \mType)\ (\inU\ X) & \El &\defeq \pi_1
+\end{align*}
 
@@
 \begin{align*}
-  & \blank\sim_{\setoidU}\blank : \setoidU \to \setoidU \to \mProp_1 \\
-  & (A , a) \sim_{\setoidU} (B , b) =
+  \blank\sim_{\setoidU}\blank &: \setoidU \to \setoidU \to \mProp_1 \\
+  (A , a) \sim_{\setoidU} (B , b) &\defeq
    \| \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \|
@@
 \begin{align*}
-  & \blank\vdash\blank\sim_{\El}\blank : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
-  & p \vdash x \sim_{\El} y \defeq \ ?
+  \blank\vdash\blank\sim_{\El}\blank & : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
+  p \vdash x \sim_{\El} y & \defeq \ ?
 \end{align*}
@@
 \vspace{-0.2em}
-\begin{alignat*}{10}
-  & \bullet && : \textsf{Con}                                                          &&  \iota && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
-  & \blank,\blank && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{3em} &&  \Pi && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
-\end{alignat*}
+\begin{align*}
+  \bullet &: \textsf{Con}                                                                  &  \iota &: (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
+  \blank,\blank &: (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{3em} &  \Pi &: \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
+\end{align*}
 
@@
 \begin{minipage}{0.4\textwidth}
-\begin{alignat*}{10}
-  & \bullet_0 && : \textsf{Con}_0 \\
-  & \blank,_0\blank && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
-  & \iota_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \\
-  & \Pi_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \hspace{2.8em} \\
-\end{alignat*}
+\begin{align*}
+  \bullet_0 &: \textsf{Con}_0 \\
+  \blank,_0\blank &: \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
+  \iota_0 &: \textsf{Con}_0 \to \textsf{Ty}_0 \\
+  \Pi_0 &: \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \hspace{2.8em} \\
+\end{align*}
 \end{minipage}
@@
 \begin{alignat*}{10}
-  & \bullet_1 && : \textsf{Con}_1\ \bullet_0 \\
-  & \blank,_1\blank && : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
-  & && \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
-  & \iota_1 && : \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
-  & \Pi_1 && : \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \\
-  & && \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \\
-  & && \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
+  \bullet_1 &: \textsf{Con}_1\ \bullet_0 \\
+  \blank,_1\blank &: \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
+  & \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
+  \iota_1 &: \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
+  \Pi_1 &: \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \\
+  & \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \\
+  & \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
 \end{alignat*}

### 2021-01-21 paper.tex
Clarification on use of weird identity type.
fixes suggested by reviewer 2, saving some space when rewriting the intro of simple eliminators

@@
 %
-We only use this identity type with such a strong transport in Section~\ref{inductive-universe}.
+We only use this identity type in Section~\ref{inductive-universe}.
+For the rest of our constructions, the traditional Martin-L\"of's identity type suffices.
+
 %
@@
 SeTT's internal universe of propositions, here we state the context equalities
-as elements of the external, metatheoretic universe $\mProp$. This makes it easier
-to define models of SeTT as contexts do not need to be partially fibrant. Equality on
+as elements of the external, metatheoretic universe $\mProp$. This generalises the
+notion of model of SeTT thus making it easier to construct models. Equality on
 types is defined as before in \cite{mpc19}.
@@
 and one for sets (with propositions convertible to sets) or a single sort of
-types (sets), with propositions given by elements of a universe of proposition,
+types (sets), with propositions given by elements of a universe of propositions,
 which is a (large) type. We have chosen to present the second option to fit with
@@
 
-For any inductive-inductive type $A : \mType, B: A \to \mType$, the most general
-form of (dependent) eliminator is given by motives $F : A \to \mType$ and $G :
-(a : A) \to (b : B a) \to F\ a \to \mType$, and the functions $\textsf{elim}_A :
-(a : A) \to F\ a, \textsf{elim}_B : (a : A) \to (b : B\ a) \to
-G\ a\ b\ (\textsf{elim}_A\ a)$. A simpler notion of eliminator does not require
-the second component of the motive to mention the first. In particular, we have
-$F : A \to \mType$, $G : (a : A) \to B\ a \to \mType$, and eliminators
-$\textsf{elim}_A : (a : A) \to F\ a, \textsf{elim}_B : (a : A) (b : B\ a) \to
-G\ a\ b$. Note that this elimination principle---that we call \emph{simple
-  elimination} after \cite{ii}---is no longer recursive-recursive, since
-$\textsf{elim}_A$ is not mentioned in the signature of $\textsf{elim}_B$.
-
-Let us go back to our \textsf{Con}/\textsf{Ty} example. Given any dependent
-algebra of $\textsf{Con}$ and $\textsf{Ty}$ with motives $C : \textsf{Con} \to
-\mType$ and $T : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to C\ \GG \to
-\mType$, the general eliminators have the following signatures:
+Following \cite{ii}, we distinguish two versions of the eliminator:
+the simple and the general one. Note that this is orthogonal to the
+distinction between nondependent and dependent eliminators, from which
+we only consider the latter. The motives for the simple eliminator are
+$C':\textsf{Con}\to\mType$,
+$T':(\GG:\textsf{Con})(A:\textsf{Ty}\,\GG)\to \mType$ and the
+eliminators themselves have the following signatures:
 \begin{alignat*}{10}
-  & \textsf{elim}_{\textsf{Con}} && : (\GG : \textsf{Con}) \to C\ \GG \hspace{5em} \\
-  & \textsf{elim}_{\textsf{Ty}} && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to T\ \GG\ A\ (\textsf{elim}_{\textsf{Con}}\ \GG)
+  & \textsf{elim}_{\textsf{Con}}' && : (\GG : \textsf{Con}) \to C'\ \GG \hspace{3.3em} \textsf{elim}_{\textsf{Ty}}' && : \forall\{\GG\} (A : \textsf{Ty}\ \GG) \to T'\ \GG\ A \hphantom{(\textsf{elim}_{\textsf{Con}}\ \GG)}
 \end{alignat*}
-
-These can be derived from our encoding of $\textsf{Con}$ and $\textsf{Ty}$ via
+In the case of the general eliminator, the motive for $\textsf{Ty}$
+depends on the motive for $\textsf{Con}$, making the two eliminators
+\emph{recursive-recursive} functions. For motives $C :
+\textsf{Con} \to \mType$ and $T : (\GG : \textsf{Con}) \to
+\textsf{Ty}\ \GG \to C\ \GG \to \mType$ the signatures are:
+\begin{alignat*}{10}
+  & \textsf{elim}_{\textsf{Con}} && : (\GG : \textsf{Con}) \to C\ \GG \hphantom{'}\hspace{3.3em} \textsf{elim}_{\textsf{Ty}} && : \forall\{\GG\} (A : \textsf{Ty}\ \GG) \to T\ \GG\ A\ (\textsf{elim}_{\textsf{Con}}\ \GG)
+\end{alignat*}
+
+The general eliminators can be derived from our encoding of $\textsf{Con}$ and $\textsf{Ty}$ via
 untyped codes and well-typing predicates. The way to do it is to first define

### 2021-01-22 paper.tex
??

--- 2021-01-21 paper.tex
@@
 \usepackage{bbm}
-\usepackage{mathtools}
+\usepackage{marvosym}
 %\usepackage[bw]{agda}
@@
 
-% for placeholders
-\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
-
-% for definitions
-\newcommand{\defeq}{\mathrel{\vcentcolon\equiv}}
-
 % TODO notes.
@@
   \thanks{Supported by USAF grant FA9550-16-1-0029.}
-  \orcidID{0000-0002-6582-5025} \and
+  \orcidID{0000-0002-6582-5025}\Letter \and
   Simon Boulier\inst{2}\thanks{Supported by ERC Starting Grant CoqHoTT 637339.} \and
@@
 go beyond inductive families.
-$\mlttp$ is thus an intensional type theory in the style of Martin-L\"of type theory.
-
-We have sorts $\Type_i$ of types and $\mProp_i$ of strict propositions for $i \in \{0, 1\}$.
-Here, $i = 0$ means ``small'' (and we will omit the subscript) and $i = 1$ means ``large''.
-We have implicit lifting from $i = 0$ to $i = 1$, but do not assume type formers are preserved.
-$\Type_1$ has universes for $\Type$ and $\mProp$.
-We do not distinguish notationally between universes and sorts.
-We continue to describe only the case $i = 0$; everything introduced has an analogue at level $i = 1$.
-Propositions lift to types via $\textsf{Lift} : \mProp \to \mType$, with constructor
+
+$\mlttp$ is thus an intensional type theory in the style of Martin-L\"of type
+theory with $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$, and a universe
+of strict propositions $\mProp$. $\Sigma$ types are defined negatively by
+pairing $\_,\_$ and projections $\pi_1, \pi_2$. We also have definitional $\eta$
+rules for $\Pi, \Sigma$, and $\mathbf{1}$ types. Propositions can be lifted to
+types via a lifting $\textsf{Lift} : \mProp \to \mType$, with constructor
 $\textsf{lift} : \{P : \mProp \} \to P \to \textsf{Lift}\ P$ and destructor
@@
 
-We have standard type formers $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$ in $\Type$.
-$\Sigma$-types are defined negatively by pairing $\blank,\blank$ and projections $\pi_1, \pi_2$.
-We have definitional $\eta$-rules for $\Pi$-, $\Sigma$-, $\mathbf{1}$-types.
-We also require indexed W-types, both in $\mType$ and $\mProp$:
-$W_\Box : (S : I\to\mType) \to ((i:I)\to S\ i\to I \to \mType) \to I\to \Box$ where $\Box \in \{\mType, \mProp\}$.
-The elimination principle of $W_{\mProp}$ only allows defining functions into elements of $\mProp$.
-%
-From $W_{\mProp}$ we can define propositional truncation $\|\blank\| : \mType \to \mProp$, with constructor
-  $|\blank| : \{A : \mType \} \to A \to \| A \|$ and eliminator
-  $\textsf{elim}_{\|\blank\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P$.
+We also require indexed W-types, both in $\mType$ and $\mProp$; hence we have $W_\Box
+: (S : I\to\mType) \to ((i:I)\to S\ i\to I \to \mType) \to I\to \Box$ where $\Box \in \{\mType, \mProp\}$.
+The elimination principle of $W_\mProp$ only allows defining functions into $\mProp$s.
+%
+From $W_\mProp$ we can define propositional truncation $\|\_\| : \mType \to \mProp$, with constructor
+  $|\_| : \{A : \mType \} \to A \to \| A \|$ and eliminator
+  $\textsf{elim}_{\|\_\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P$.
 
@@
 \begin{align*}
-  \mathbf{1}_\mProp & \defeq \| \mathbf{1} \| \\
-  \Pi_\mProp\ P\ Q & \defeq \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
-  \Sigma_\mProp\ P\ Q & \defeq \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \|
+  \mathbf{1}_\mProp & :\equiv \| \mathbf{1} \| \\
+  \Pi_\mProp\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
+  \Sigma_\mProp\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \|
 \end{align*}
@@
 types instead of deriving them from truncation. We operate under the assumption
-that everything can be equivalently carried out in $\mlttp$, although we have not
+that everything can be equivalently carried out in $\mlttp$, although we haven't
 fully checked all the necessary details.
@@
 propositions with propositional extensionality \cite{setoid99}. Note that
-we do not need identity types or inductive types (W-types) for this.
+we don't need identity types or inductive types (W-types) for this.
 
@@
 %% \begin{align*}
-%%   & (\Pi(a : A)\ B)^\sim\ p\ f_0\ f_1 = \Pi(a_0\,a_1 : A)\ldots \\
-%%   & (\Sigma\ A\ B)^\sim\ p\ (a_0, b_0)\ (a_1, b_1) = \ldots \\
-%%   & \Prop^\sim\ p\ P\ Q = \ldots
+%%   & (\Pi(a : A)\ B)^\sim\ p\ f_0\ f_1 = \Pi(a_0\,a_1 : A)... \\
+%%   & (\Sigma\ A\ B)^\sim\ p\ (a_0, b_0)\ (a_1, b_1) = ... \\
+%%   & \Prop^\sim\ p\ P\ Q = ...
 %% \end{align*}
@@
 \infer{
-  \textsf{Id}_A\ a_0\ a_1 \defeq A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
+  \textsf{Id}_A\ a_0\ a_1 :\equiv A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
 }{
@@
 \infer{
-  \textsf{transp}\ P\ p\ t \defeq
+  \textsf{transp}\ P\ p\ t :\equiv
     \coe\ P\ (\textsf{R}\ \textsf{id} , p)\ t : \Tm\ \GG\ P[a_1]
@@
 structure of the codes. Such an elimination principle is known to be inconsistent
-with univalence, although this is not an issue in our case; nevertheless it is
+with univalence, although this is not an issue in our case; nevertheless it's
 not immediately clear if the elimination principle can be justified by the
@@
   & \El\ : \U \to \mType \\
-  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
-  & \El\ (\textsf{pi}\ A\ B) \defeq (a : \El\ A) \to \El\ (B\ a) \\
+  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
+  & \El\ (\textsf{pi}\ A\ B) :\equiv (a : \El\ A) \to \El\ (B\ a) \\
 \end{align*}
@@
 using the equivalence of $I$-indexed families of types and types over $I$
-(that is, $A : \mType$ with $A \to I$) \cite{malatasta13smallir}.
+(that is, $\mType_i \to I$) \cite{malatasta13smallir}.
 
@@
 
-$\U$ and $\El$ can be given by $\U \defeq \Sigma\ (A : \mType)\ (\textsf{in-U}\ A)$ and $\El \defeq \pi_1$.
+$\U$ and $\El$ can be given by $\U :\equiv \Sigma\ (A : \mType)\ (\textsf{in-U}\ A)$ and $\El :\equiv \pi_1$.
 
@@
   & \textsf{data} \ \setoidU : \mType_1 \\
-  & \blank\sim_\setoidU\blank : \setoidU \to \setoidU \to \mProp_1
+  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
 \end{align*}
@@
   & \El : \setoidU \to \mType \\
-  & \blank\vdash\blank\sim_\El\blank : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
+  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
 \end{align*}
 
-Note that $\eqel{\blank}{\blank}{\blank}$ is indexed over equality on the universe, because
+Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
 $\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
@@
 of the 4 functions
-$\blank\sim_\setoidU\blank$, $\textsf{refl}_\setoidU$, $\El$ and $\blank\vdash\blank\sim_\El\blank$.
+$\_\sim_\setoidU\_$, $\textsf{refl}_\setoidU$, $\El$ and $\_\vdash\_\sim_\El\_$.
 The other functions are then recursively defined:
@@
 \begin{align*}
-  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
-  & \El\ (\textsf{pi}\ A\ B\ h) \defeq \\
+  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
+  & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
   & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
@@
 \begin{align*}
-  & | \Uty | \defeq \lambda\, \gamma . \, \setoidU \\
-  & \Uty^\sim \defeq \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
-  & \textsf{refl}\ \Uty \defeq \reflu \\
-  & \ldots \\
-  & \textsf{coe}\ \Uty \defeq \lambda\,p\,a . \, a \\
-  & \textsf{coh}\ \Uty \defeq \lambda\,p . \, \reflu
+  & | \Uty | :\equiv \lambda\, \gamma . \, \setoidU \\
+  & \Uty^\sim :\equiv \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
+  & \textsf{refl}\ \Uty :\equiv \reflu \\
+  & ... \\
+  & \textsf{coe}\ \Uty :\equiv \lambda\,p\,a . \, a \\
+  & \textsf{coh}\ \Uty :\equiv \lambda\,p . \, \reflu
 \end{align*}
@@
 \begin{align*}
-  & | \EL\ A | \defeq \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
-  & (\EL\ A)^\sim \defeq \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
-  & \textsf{refl}\ (\EL\ A) \defeq \reflel \\
-  & \ldots \\
-  & \textsf{coe}\ (\EL\ A) \defeq \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
-  & \textsf{coh}\ (\EL\ A) \defeq \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
+  & | \EL\ A | :\equiv \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
+  & (\EL\ A)^\sim :\equiv \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
+  & \textsf{refl}\ (\EL\ A) :\equiv \reflel \\
+  & ... \\
+  & \textsf{coe}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
+  & \textsf{coh}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
 \end{align*}
@@
 In the next section we transform our current inductive-recursive definition to
-one that does not use induction-recursion.
+one that doesn't use induction-recursion.
 %
@@
 $\inUU$ of binary relations between types in the universe, from which we
-then define $\blank\sim_{\setoidU}\blank$.
+then define $\_\sim_{\setoidU}\_$.
 {\small\vspace{-0.2em}
@@
 of $\El$, in the same way $\inUU\ a\ a'$ classifies all relations of type $A \to
-A' \to \mProp$ that are image of $\blank\vdash\blank\sim_{\El}\blank$, given proofs $a :
+A' \to \mProp$ that are image of $\_\vdash\_\sim_{\El}\_$, given proofs $a :
 \inU\ A, a' : \inU\ A'$.
@@
 equivalence for boolean elements is the obvious syntactic equality
-$\blank\stackrel{?}{=}_{\mathbbm{2}}\blank$, whereas functions are to be compared
+$\_\stackrel{?}{=}_{\mathbbm{2}}\_$, whereas functions are to be compared
 pointwise.
@@
   & \setoidU && : \mType_1 && \El : \setoidU \to \mType \\
-  & \setoidU && \defeq \Sigma\ (X : \mType)\ (\inU\ X) \hspace{5em} && \El \defeq \pi_1
+  & \setoidU && :\equiv \Sigma\ (X : \mType)\ (\inU\ X) \hspace{5em} && \El :\equiv \pi_1
 \end{alignat*}
@@
 \[
-(A, a) \sim_{\setoidU} (B, b) \defeq \Sigma\ (R: A \to B \to \mProp)\ (\inUU\ a\ b\ R)
+(A, a) \sim_{\setoidU} (B, b) :\equiv \Sigma\ (R: A \to B \to \mProp)\ (\inUU\ a\ b\ R)
 \]
@@
 Unfortunately this definition only manages to capture the idea, but does not
-actually typecheck. In fact, $\blank\sim_{\setoidU}\blank$ should be a $\mProp_1$-valued
+actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be $\mProp_1$-valued
 relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
-$\Sigma$ type shown above clearly is not, since it quantifies over a type of
+$\Sigma$ type shown above clearly isn't, since it quantifies over a type of
 relations, which is not a proposition.
@@
 \begin{align*}
-  & \blank\sim_{\setoidU}\blank : \setoidU \to \setoidU \to \mProp_1 \\
+  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \mProp_1 \\
   & (A , a) \sim_{\setoidU} (B , b) =
-   \| \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \|
+   \ \parallel \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \parallel
 \end{align*}
@@
 \begin{align*}
-  & \blank\vdash\blank\sim_{\El}\blank : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
-  & p \vdash x \sim_{\El} y \defeq \ ?
+  & \_\vdash\_\sim_{\El}\_ : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
+  & p \vdash x \sim_{\El} y :\equiv \ ?
 \end{align*}
 
-In the definition above, $p$ has type $\| \Sigma\ (R: \El\,A \to \El\,B \to
-\mProp)\ (\ldots) \|$. If the type was not propositionally truncated, we could define
+In the definition above, $p$ has type $\parallel \Sigma\ (R: \El\,A \to \El\,B \to
+\mProp)\ (...) \parallel$. If the type wasn't propositionally truncated, we could define
 $p \vdash x \sim_{\El} y$ by extracting the relation out of the first component
-of $p$, and apply it to $x, y$. That is, $p \vdash x \sim_{\El} y \defeq
-\pi_1\,p\,x\,y$. This would make the definition of $\blank\sim_\setoidU\blank$ and
-$\blank\vdash\blank\sim_{\El}\blank$ in line with how we defined $\setoidU$ and $\El$.
-
-However, this does not work in our case, since the type of $p$ \emph{is}
+of $p$, and apply it to $x, y$. That is, $p \vdash x \sim_{\El} y :\equiv
+\pi_1\,p\,x\,y$. This would make the definition of $\_\sim_\setoidU\_$ and
+$\_\vdash\_\sim_{\El}\_$ in line with how we defined $\setoidU$ and $\El$.
+
+However, this doesn't work in our case, since the type of $p$ \emph{is}
 propositionally truncated, hence it cannot be eliminated to construct a
@@
 are logically equivalent. These in turn need to be defined mutually with
-$\blank\vdash\blank\sim_{\El}\blank$. We direct the interested reader to the Agda
+$\_\vdash\_\sim_{\El}\_$. We direct the interested reader to the Agda
 formalization for the full details of these definitions, as they are quite
@@
 more or less straightforwardly. The final result does not use
-induction-recursion, but it is nevertheless an instance of infinitary
+induction-recursion, but it's nevertheless an instance of infinitary
 induction-induction. The ability to define arbitrary, infinitary
@@
 Contexts in \textsf{Con} are formed out of empty contexts $\bullet$ and context
-extension $\blank,\blank$. Types in \textsf{Ty} are either the base type $\iota$ or
+extension $\_,\_$. Types in \textsf{Ty} are either the base type $\iota$ or
 $\Pi$ types.
@@
   & \bullet && : \textsf{Con}                                                          &&  \iota && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
-  & \blank,\blank && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{3em} &&  \Pi && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
+  & \_,\_ && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{3em} &&  \Pi && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
 \end{alignat*}
@@
   & \bullet_0 && : \textsf{Con}_0 \\
-  & \blank,_0\blank && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
+  & \_,_0\_ && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
   & \iota_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \\
@@
   & \bullet_1 && : \textsf{Con}_1\ \bullet_0 \\
-  & \blank,_1\blank && : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
+  & \_,_1\_ && : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
   & && \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
@@
 
-We can recover the original inductive-inductive type as $\textsf{Con} \defeq
+We can recover the original inductive-inductive type as $\textsf{Con} :\equiv
 \Sigma\ (\GG_0 : \textsf{Con}_0)\ (\textsf{Con}_1\ \GG_0)$ and $\textsf{Ty}\ \GG
-\defeq \Sigma\ (A_0 :
+:\equiv \Sigma\ (A_0 :
 \textsf{Ty}_0)\ (\textsf{Ty}_1\ (\pi_1\ \GG)\ A_0)$. Recovering the constructors
@@
 \begin{alignat*}{10}
-  & \bullet && \defeq (\bullet_0 , \bullet_1) \\
-  & (\GG_0 , \GG_1) , (A_0 , A_1) && \defeq ((\GG_0 \mathbin{,_0} A_0) , (\GG_1 ,_1 A_1)) \\
-  & \iota\ (\GG_0 , \GG_1) && \defeq (\iota_0\ \GG_0 , \iota_1\ \GG_1) \\
-  & \Pi\ \{\GG_0 , \GG_1\}(A_0 , A_1) (B_0 , B_1) && \defeq
+  & \bullet && :\equiv (\bullet_0 , \bullet_1) \\
+  & (\GG_0 , \GG_1) , (A_0 , A_1) && :\equiv ((\GG_0 \mathbin{,_0} A_0) , (\GG_1 ,_1 A_1)) \\
+  & \iota\ (\GG_0 , \GG_1) && :\equiv (\iota_0\ \GG_0 , \iota_1\ \GG_1) \\
+  & \Pi\ \{\GG_0 , \GG_1\}(A_0 , A_1) (B_0 , B_1) && :\equiv
     (\Pi_0\ \GG_0\ A_0\ B_0 , \Pi_1\ \GG_1\ A_1\ B_1)
@@
 %
-Of course, the absence of a general reduction method does not mean that we cannot
+Of course, the absence of a general reduction method does not mean that we can't
 reduce particular concrete instances of infinitary induction-induction, which is
@@
 to derive the general recursive-recursive eliminators we need to prove that the
-corresponding graph relations are functional, which cannot be done without funext.
+corresponding graph relations are functional, which can't be done without funext.
 

### 2021-01-22 paper.tex

--- 2021-01-21 paper.tex
@@
 \usepackage{bbm}
+\usepackage{marvosym}
 \usepackage{mathtools}

### 2021-01-22 paper.tex
18p

@@
 
+\setlength{\abovedisplayskip}{2pt}
+\setlength{\belowdisplayskip}{1pt}
 % TODO notes.
@@
 formers. The core structure of a CwF can be given as the following signature:
-\vspace{-0.2em}
+{\small\vspace{-0.2em}
 \begin{align*}
@@
   \Tm &: (\GG : \Con) \to \Ty\ \GG \to \mType
-\end{align*}
+\end{align*}}\vspace{-1em}
 
@@
 irrelevant.
-\[
+{\small\[
 \infer{
@@
 }
-\]
+\]}\vspace{-1em}
 
@@
 the following, we sometimes omit implicit quantifications such as the $\forall\{\gamma_0\ \gamma_1\}$ in the type of $\textsf{sym}\,\GG$.
-\[
+{\small\[
 \infer{
@@
 \\
-\begin{minipage}{0.5\textwidth}
+{\small\begin{minipage}{0.5\textwidth}
 \[
@@
 \end{minipage}
-\\
+\\}\vspace{-1em}
 
@@
 We have a universe of propositions $\Prop$ defined as follows:
-\[
+{\small\[
 \infer{
@@
 equality.
-\[
+{\small\[
 \infer{
@@
 recall reflexivity:
-\[
+{\small\[
 \infer{
@@
 }
-\]
+\]}\vspace{-1em}
 
 In addition, we also have rules representing the fact that every construction in
-SeTT respects setoid equality, so that we can transport along any such equality:
+SeTT respects setoid equality, so that we can transport along any such
+equality:
+{\small
     \[
@@
     }
-    \]
+    \]}\vspace{-1em}
 
@@
 with transport implemented via coercion.
-\[
+{\small\[
 \infer{
@@
 As an example, consider the following definition of a small universe with
-bool and $\Pi$. \\
-\vspace{-0.2em}
+bool and $\Pi$.
+
+{\small\vspace{-0.4em}
 \begin{minipage}{0.45\textwidth}
@@
 \end{align*}
-\end{minipage}
+\end{minipage}}\vspace{-1em}
 
@@
 inductive structure of codes.
-\vspace{-0.2em}
+{\small\vspace{-0.4em}
 \begin{alignat*}{10}
@@
   & \quad \textsf{inPi} && :  \inU\ A \to ((a : A) \to \inU\ (B\ a)) \to \inU\ ((a : A) \to (B\ a))
-\end{alignat*}
+\end{alignat*}}\vspace{-2em}
 
@@
 indeed equivalence relations: \\
-\begin{minipage}{0.5\textwidth}
+
+{\small\begin{minipage}{0.5\textwidth}
 \vspace{-0.2em}
@@
 \end{align*}
-\end{minipage}
-\\
+\end{minipage}}\vspace{0.4em}
 
@@
 relation.
-\vspace{-0.2em}
+{\small\vspace{-0.4em}
 \begin{align*}
@@
   \blank\vdash\blank\sim_\El\blank &: \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
-\end{align*}
+\end{align*}}\vspace{-1.4em}
 
@@
 We also require data and proofs that make sure we get setoids out of $\El$:
-\begin{align*}
+{\vspace{-0.4em}\small\begin{align*}
   \textsf{refl}_\El &: (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
@@
   \cohel &: (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
-\end{align*}
-%
+\end{align*}}\vspace{-1.6em}
+
 We give an inductive definition of $\setoidU$, mutually with a recursive definition
@@
 $\blank\sim_\setoidU\blank$, $\textsf{refl}_\setoidU$, $\El$ and $\blank\vdash\blank\sim_\El\blank$.
-The other functions are then recursively defined:
-\begin{itemize}
-\item $\textsf{refl}_\El$ alone,
-\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
-\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
-\end{itemize}
+The other functions are then recursively defined: $\textsf{refl}_\El$ alone,
+$\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
+$\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
+% The other functions are then recursively defined:
+% \begin{itemize}
+% \item $\textsf{refl}_\El$ alone,
+% \item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
+% \item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
+% \end{itemize}
 The whole construction is quite long, below we only show the more interesting
-definitions of $\setoidU$ and $\El$: \\
-\vspace{-0.2em}
+definitions of $\setoidU$ and $\El$:
+
+{\small\vspace{-1em}
 \begin{minipage}{0.5\textwidth}
@@
 \end{align*}
-\end{minipage}
-\\
+\end{minipage}}\vspace{1em}
 
@@
 any $A : \Tm\ \GG\ \Uty$: \\
-\vspace{-0.2em}
+{\small\vspace{-0.6em}
 \begin{minipage}{0.5\textwidth}
@@
 \end{align*}
-\end{minipage}
+\end{minipage}}
 \\
@@
 constructions if more codes are added to $\setoidU$.
-
+%
 This gives a complete definition of a universe of setoids, which is, however,
@@
   & \qquad \textsf{bool} : \inU\ \mathbbm{2} \\
-  & \qquad \pi : \forall\{A\;A_\sim\;B\;B_\sim\}(a : \inU\;A)
-  (b : (x : A) \to \inU\;(B\;x))\\
-  & \qquad\ \to \inUU\;a\;a\;A_\sim \to (\forall\{x_0\;x_1\}(x_{01} : A_\sim\;x_0\;x_1)
+  & \qquad \pi : \inUU\;a\;a\;A_\sim \to (\forall\{x_0\;x_1\}(x_{01} : A_\sim\;x_0\;x_1)
   \to \inUU\;(b\;x_0)\;(b\;x_1)\;(B_\sim\;x_{01})) \\
@@
   (\Sigma\;(f : (x : A) \to B\;x) \\
-  &\qquad\qquad\qquad\quad\, ((x_0\;x_1 : A)(x_{01} : A_\sim\;x_0\;x_1) \to B_\sim\;x_{01}\;(f\;x_0)\;(f\;x_1))) \\
-  %%%%%%
+  &\qquad\qquad\qquad\quad\, ((x_0\;x_1 : A)(x_{01} :
+    A_\sim\;x_0\;x_1) \to B_\sim\;x_{01}\;(f\;x_0)\;(f\;x_1)))
+\end{align*}
+\begin{align*}
   & \textsf{data}\ \inUU : \{A\ A' : \mType\} \to \inU\ A \to \inU\ A' \to (A \to A' \to \mProp) \to \mType_1 \\
@@
   (\lambda x_0\;x_1\; .\; x_0 \stackrel{?}{=}_{\mathbbm{2}} x_1) \\
-  & \qquad \pi_\sim :
-  \forall\{A_0\;A_1\;A_{0\sim}\;A_{1\sim}\;A_{01\sim}\;B_0\;B_1\;B_{0\sim}\;B_{1\sim}\;B_{01\sim}\}
-  \{ a_0 : \inU\;A_0 \}\{ a_1 : \inU\;A_1 \}\\
-  & \qquad\quad\quad\ \; \{b_0 : (x_0 : A_0) \to \inU\;(B_0\;x_0)\}
+  & \qquad \pi_\sim : \ \; \{b_0 : (x_0 : A_0) \to \inU\;(B_0\;x_0)\}
   \{b_1 : (x_1 : A_1) \to \inU\;(B_1\;x_1)\}\\
@@
         \to B_{01\sim}\;x_{01}\;(\pi_1\;f_0\;x_0)\;(\pi_1\;f_1\;x_1))
-\end{align*}}
+\end{align*}}\vspace{-1em}
 
@@
 $\setoidU$ as a $\Sigma$ type, and $\El$ as the corresponding first projection.
-\vspace{-0.2em}
+{\small\vspace{-0.2em}
 \begin{align*}
@@
   \setoidU & \defeq \Sigma\ (X : \mType)\ (\inU\ X) & \El &\defeq \pi_1
-\end{align*}
+\end{align*}}\vspace{-1em}
 
@@
 the $\Sigma$ type above to force it to be in $\mProp_1$.
-\vspace{-0.2em}
+{\small\vspace{-0.2em}
 \begin{align*}
@@
    \| \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \|
-\end{align*}
+\end{align*}}\vspace{-1em}
 
 We are now left to define the indexed equivalence relation on $\El$:
-\vspace{-0.2em}
+{\small\vspace{-0.2em}
 \begin{align*}
@@
   p \vdash x \sim_{\El} y & \defeq \ ?
-\end{align*}
+\end{align*}}\vspace{-1em}
 
@@
 $\Pi$ types.
-\vspace{-0.2em}
+{\small\vspace{-0.2em}
 \begin{align*}
@@
   \blank,\blank &: (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{3em} &  \Pi &: \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
-\end{align*}
+\end{align*}}\vspace{-1em}
 
@@
 indexed. \\
-\vspace{-0.2em}
+{\small\vspace{-0.2em}
 \begin{minipage}{0.4\textwidth}
@@
 \end{alignat*}
-\end{minipage}
-\\
+\end{minipage}}\vspace{0.4em}
 
@@
 is straightforward:
-\vspace{-0.2em}
+{\small\vspace{-0.2em}
 \begin{alignat*}{10}
@@
     (\Pi_0\ \GG_0\ A_0\ B_0 , \Pi_1\ \GG_1\ A_1\ B_1)
-\end{alignat*}
+\end{alignat*}}\vspace{-1em}
 
@@
 the graph of the eliminators in the form of inductively-generated relations:
-\begin{alignat*}{10}
+{\small\begin{alignat*}{10}
   & \textsf{data}\ \textsf{R-Con} && :
@@
   \to \mType
-\end{alignat*}
+\end{alignat*}}\vspace{-1em}
 
@@
 $\mProp$-valued families, rather than in $\mType$:
-\vspace{-0.2em}
+{\small\vspace{-0.2em}
 \begin{alignat*}{10}
@@
   & \textsf{data}\ \inUU_1 && : \{A\ A' : \mType\} \to (R : A \to A' \to \mProp) \to \inUU_0\ R \to \mProp_1
-\end{alignat*}
+\end{alignat*}}\vspace{-1em}
 
@@
 
-Groupoids can be regarded as generalized setoids. In the future we would like to
-design a type theory internalizing the groupoid model of type theory
-\cite{groupoid}, in the same way that SeTT represents a syntax for the setoid
-model. A further question is whether such ``groupoid type theory'' can be
-justified, similarly to SeTT, via a syntactic translation, perhaps with SeTT
-itself as the target theory.
+% Groupoids can be regarded as generalized setoids. In the future we would like to
+% design a type theory internalizing the groupoid model of type theory
+% \cite{groupoid}, in the same way that SeTT represents a syntax for the setoid
+% model. A further question is whether such ``groupoid type theory'' can be
+% justified, similarly to SeTT, via a syntactic translation, perhaps with SeTT
+% itself as the target theory.
 

### 2021-01-22 paper.tex
remove mentions of formalisation being unfinished; add back the last paragraph

@@
 
-This section outlines a way to encode the inductive-inductive universe of
-setoids from the previous section without assuming arbitrary inductive-inductive
+This section encodes the inductive-inductive universe of setoids from
+the previous section without assuming arbitrary inductive-inductive
 definitions in the metatheory.
@@
 Following \cite{ii}, we distinguish two versions of the eliminator:
-the simple and the general one. Note that this is orthogonal to the
+the \emph{simple} and the \emph{general} one. Note that this is orthogonal to the
 distinction between nondependent and dependent eliminators, from which
@@
 
-Luckily, it would seem the \emph{simple} elimination principle is sufficient for
-our purposes: careful inspection of the code suggests that all functions
-described in Section \ref{ii-universe} can be defined using simple mutual
-pattern matching without recursion-recursion. The simple eliminator can be
-defined by pattern matching on the untyped codes, and does not require
-extensionality or any extra principles beyond strong transport. Note that
-producing a full formalization that uses exclusively simple eliminators appears
-to be extremely time-consuming; we therefore leave this task to future work.
+Luckily, the \emph{simple} elimination principle is sufficient for our
+purposes: all functions described in Section \ref{ii-universe} can be
+defined just using the simple eliminator without
+recursion-recursion. The simple eliminator itself can be defined by
+pattern matching on the untyped codes, and does not require
+extensionality or any extra principles beyond strong transport.
 
@@
 %
-Most of the mathematical contents of this paper have been formalized in the
-proof-assistant Agda (see \cite{agda-code}). We aim to complete the
-formalization in the near future, in particular to obtain a machine-checked
-proof that our inductive construction of the setoid universe does not rely on
-recursion-recursion.
+% Most of the mathematical contents of this paper have been formalized in the
+% proof-assistant Agda (see \cite{agda-code}). We aim to complete the
+% formalization in the near future, in particular to obtain a machine-checked
+% proof that our inductive construction of the setoid universe does not rely on
+% recursion-recursion.
 
@@
 
-% Groupoids can be regarded as generalized setoids. In the future we would like to
-% design a type theory internalizing the groupoid model of type theory
-% \cite{groupoid}, in the same way that SeTT represents a syntax for the setoid
-% model. A further question is whether such ``groupoid type theory'' can be
-% justified, similarly to SeTT, via a syntactic translation, perhaps with SeTT
-% itself as the target theory.
+Groupoids can be regarded as generalized setoids. In the future we would like to
+design a type theory internalizing the groupoid model of type theory
+\cite{groupoid}, in the same way that SeTT represents a syntax for the setoid
+model. A further question is whether such ``groupoid type theory'' can be
+justified, similarly to SeTT, via a syntactic translation, perhaps with SeTT
+itself as the target theory.
 

### 2021-01-22 paper.tex
spacing

@@
 \end{minipage}
-\\}\vspace{-1em}
+\\}
 
@@
 
-As an example, consider the following definition of a small universe with
+An example is the following definition of a small universe with
 bool and $\Pi$.
 
-{\small\vspace{-0.4em}
+{\small\vspace{-0.5em}
 \begin{minipage}{0.45\textwidth}
@@
 inductive structure of codes.
-{\small\vspace{-0.4em}
+{\small
 \begin{alignat*}{10}
@@
   & \quad \textsf{inPi} && :  \inU\ A \to ((a : A) \to \inU\ (B\ a)) \to \inU\ ((a : A) \to (B\ a))
-\end{alignat*}}\vspace{-2em}
+\end{alignat*}}\vspace{-1em}
 
@@
 one universe.
+\vspace{-0.3em}
 
@@
 relation reflecting equality of codes, in addition to proofs that these are
-indeed equivalence relations: \\
-
-{\small\begin{minipage}{0.5\textwidth}
+indeed equivalence relations:\vspace{-1em} \\
+{\small\begin{minipage}{0.4\textwidth}
 \vspace{-0.2em}
@@
 We also require data and proofs that make sure we get setoids out of $\El$:
-{\vspace{-0.4em}\small\begin{align*}
+{\small\begin{align*}
   \textsf{refl}_\El &: (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
@@
   \cohel &: (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
-\end{align*}}\vspace{-1.6em}
+\end{align*}}\vspace{-1em}
 
@@
 The whole construction is quite long, below we only show the more interesting
-definitions of $\setoidU$ and $\El$:
-
-{\small\vspace{-1em}
+definitions of $\setoidU$ and $\El$: \\
+{\small\vspace{0em}
 \begin{minipage}{0.5\textwidth}
@@
 any $A : \Tm\ \GG\ \Uty$: \\
-{\small\vspace{-0.6em}
+{\small
 \begin{minipage}{0.5\textwidth}
@@
 indexed. \\
-{\small\vspace{-0.2em}
+{\small
 \begin{minipage}{0.4\textwidth}

### 2021-01-22 paper.tex
small change
reference to formalisation

@@
 funext and propext hold by definition. SeTT is based on the \emph{strict} setoid
-model of Altenkirch\footnote{a \emph{strict} model is one where every equation
-  holds definitionally}, which makes it possible to show consistency via a
+model of Altenkirch\footnote{A \emph{strict} model is one where every equation
+  holds definitionally.}, which makes it possible to show consistency via a
 syntactic translation. This is in contrast with other type theories based on the
@@
 indexed. \\
+
 {\small

### 2021-02-11 paper.tex
ordcid etc
Uni Nottingham is in Nottingham
open access block and avoid overfill

@@
 %% \documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{llncs}
-\documentclass[autoref]{llncs}
+%\documentclass[autoref]{llncs}
+\documentclass[runningheads]{llncs}
+
+\usepackage{graphicx}
+%\renewcommand\UrlFont{\color{blue}\rmfamily}
+
+%%%%%% To display ORCID Logo with link, Please add below definition and copy the ORCID_Color.eps in the manuscript package %%%%%
+
+\makeatletter
+\RequirePackage[bookmarks,unicode,colorlinks=true]{hyperref}%
+   \def\@citecolor{blue}%
+   \def\@urlcolor{blue}%
+   \def\@linkcolor{blue}%
+\def\UrlFont{\rmfamily}
+\def\orcidID#1{\smash{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid_color.eps}}}}}
+\makeatother
+
+\makeatletter
+\renewcommand*{\@fnsymbol}[1]{\ensuremath{\ifcase#1\or *\or \dagger\or \ddagger\or
+    \mathsection\or \mathparagraph\or \|\or **\or \dagger\dagger
+    \or \ddagger\ddagger \else\@ctrerr\fi}}
+\makeatother
 
@@
 \newcommand{\cat}[1]{\underline{\mathbf{#1}}}
-\newcommand{\U}{\textsf{U}}
+\renewcommand{\U}{\textsf{U}}
 \newcommand{\El}{\textsf{El}}
@@
 
-\institute{School of Computer Science, University of Nottingham, UK
+\institute{School of Computer Science, University of Nottingham, Nottingham, UK
   \email{\{psztxa,psxfs5\}@nottingham.ac.uk} \and
@@
 
-\authorrunning{T. Altenkirch and S. Boulier and A. Kaposi and C. Sattler and F. Sestini}
+\authorrunning{T. Altenkirch, S. Boulier, A. Kaposi, C. Sattler and F. Sestini}
 
@@
 
+\vfill
+
+{\small\medskip\noindent{\bf Open Access} This chapter is licensed under the terms of the Creative Commons\break Attribution 4.0 International License (\url{http://creativecommons.org/licenses/by/4.0/}), which permits use, sharing, adaptation, distribution and reproduction in any medium or format, as long as you give appropriate credit to the original author(s) and the source, provide a link to the Creative Commons license and indicate if changes were made.}
+
+{\small \spaceskip .28em plus .1em minus .1em The images or other third party material in this chapter are included in the chapter's Creative Commons license, unless indicated otherwise in a credit line to the material.~If material is not included in the chapter's Creative Commons license and your intended\break use is not permitted by statutory regulation or exceeds the permitted use, you will need to obtain permission directly from the copyright holder.}
+
+\medskip\noindent\includegraphics{cc_by_4-0.eps}
+
 \end{document}
